---
title: "Call Density Tables and Plots by Season"
author: "Brian Miller"
date: "2022-11-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

# setwd("S:/manuscripts/2022-callDensityExperiments")
files <- dir(pattern = paste0('^density_',p$siteCode), full.names = TRUE)
d <- read.csv(files[1])

for (i in 2:length(files))   {
  d <- rbind(d,read.csv(files[i]))
}
d$Dc <- d$Dc*1e3
d$X <- factor(d$X, levels=c('summer','autumn','winter','spring','year'))
names(d)[1]<-'season'
d<- with(d, d[order(season),])

# Trying to get 95% confidence intervals here, but not very confident 
d$CI.low <- (d$Dc-(1.96*d$Dc*d$CV.Dc))
d$CI.high <- (d$Dc+(1.96*d$Dc*d$CV.Dc))
# d <- format(d,digits=3,nsmall=0, scientific=T)

```

#### Table of inputs into the seasonal density estimates

```{r}
library(kableExtra, quietly=T, warn.conflicts = F)

# expSup <- function(x, digits=2) {
#   sprintf(paste0("%5.", digits, "f x 10^%d"), 
#           x/10^floor(log10(abs(x))), floor(log10(abs(x))))
# }

separateCommon <- TRUE

if (separateCommon){
  commonTable <- subset(d[1,], select=c('k','w','SLmean','SLsd'))
  rownames(commonTable) <- 'All'
  common <- kbl(commonTable, row.names=TRUE) %>% 
    kable_classic_2(full_width=FALSE)
  common
  
  inputTable <- subset(d,select=c('season','Nc','c','T','pa',
                                  'NLmean','NLsd') )
  inputs <- kbl( inputTable, row.names = FALSE,
                 digits=c(0,0,3,1,4,1,1) ) %>% 
    kable_classic_2(full_width=FALSE)
} else {
  inputTable <- subset(d,select=c('season','Nc','c','k','T','w','pa',
                                  'NLmean','NLsd','SLmean','SLsd') )
  
  inputs <- kbl( inputTable, row.names = FALSE,
                 digits=c(0,0,3,0,1,0,4,1,1,1,1) ) %>% 
    kable_classic_2(full_width=FALSE)
}

inputs

```

#### Table of resulting densities and CVs

```{r}

resultsTable <- subset(d, select=c('season','Dc',
                                   'CV.Nc','CV.c','CV.pa','CV.Dc') )

result<- kbl( resultsTable, row.names = FALSE, digits=c(0,3,4,3,3,3) ) %>% 
  kable_classic_2(full_width = FALSE)
result

```

## Plot of results

Figure - (A) Mean seasonal detection rate (in calls h{\^{-1}}) of automated detections from the spectrogram correlation detector (including Bm-Ant-A, Bm-Ant-B, and Bm-Ant-Z) for the site-year S. Kerguelen Plateau 2014. (B) Seasonal call densities of Antarctic blue whale song calls for the same site and detector, with error-bars showing 95% confidence intervals.

```{r, include=FALSE, warning=FALSE}
library(ggplot2,quietly=T,warn.conflicts = F)
library(gridExtra,quietly = T,warn.conflicts = F)
library(latex2exp, quietly=T, warn.conflicts=F)
library(lubridate, quietly = T, warn.conflicts = F)
library(epitools,quietly = T,warn.conflicts = F)
library(dplyr,quietly = T,warn.conflicts = F)

det<- read.csv(p$fullYearDetectionCsv)
det$t <-mat2Rdate(det$t0)
det$season <- time2season(det$t)
det$group <- floor_date(det$t, "1 hour")

effYear <- read.csv(p$fullYearEffortFile)
effYear$t<- mat2Rdate(effYear$startDate)
effYear$group <-floor_date(effYear$t, "1 hour")
effYear$season <- time2season(effYear$t)

detHourly <- det %>% 
  group_by(group=floor_date(t,"1 hour")) %>%
  dplyr::summarize(n=n())

de <- left_join(effYear, detHourly, by = "group")
de$n = ifelse(is.na(de$n), 0, de$n);

de$rate <- de$n/(de$duration/3600) # Convert to calls/hour
  
seasonRate <- de %>% 
  group_by(season) %>%
  dplyr::summarize(count=sum(n),            
            duration_h = sum(duration/3600),
            seasonRate= count/duration_h,
            meanHourlyRate=mean(rate),
            sdHourlyRate = sd(rate),
            CI.low = meanHourlyRate-1.96*sdHourlyRate,
            CI.high = meanHourlyRate+1.96*sdHourlyRate)
seasonRate

pois.exact(seasonRate$count,pt=seasonRate$duration_h)

t0 <- min(floor_date(effYear$t, "hour"))
tEnd <- max(ceiling_date(effYear$t, "day"))
hours <- seq(t0,tEnd,by='hours')
# plot(de$t,de$rate)


gRate <- ggplot(data=seasonRate, aes(x=season, y=meanHourlyRate, fill=season) )+
  geom_bar(stat="identity",show.legend=F)+
  geom_errorbar(aes(ymin=CI.low,ymax=CI.high, width=0.5))
gRate

```

```{r fig.height=6, fig.width=3.0, warning=FALSE}

gRate <- ggplot(data=d, aes(x=season, y=Nc/T, fill=season))+
  geom_bar(stat="identity",show.legend=F)+
  # scale_y_continuous(limits=c(0,15))+
  # geom_errorbar(aes(ymin=CI.low, ymax=CI.high), width=0.5, position=position_dodge(0.9))+
  ylab( TeX("Mean detection rate ($ calls \\cdot h^{-1})" ) )+
  xlab('')+
  ggtitle(siteCode, )+
  labs(tag="(A)") + 
  theme_minimal()+
  theme(plot.tag = element_text(),
        plot.tag.position = c(0.2, 0.975),
        plot.title = element_text(hjust = 0.5))

gRateCorrected <- ggplot(data=d, aes(x=season, y=(Nc*c)/T, fill=season))+
  geom_bar(stat="identity",show.legend=F)+
  # scale_y_continuous(limits=c(0,15))+
  # geom_errorbar(aes(ymin=CI.low, ymax=CI.high), width=0.5, position=position_dodge(0.9))+
  ylab( TeX("Mean detection rate ($ calls \\cdot h^{-1})" ) )+
  xlab('')+
  ggtitle(siteCode, )+
  labs(tag="(A)") + 
  theme_minimal()+
  theme(plot.tag = element_text(),
        plot.tag.position = c(0.2, 0.975),
        plot.title = element_text(hjust = 0.5))

gDen <- ggplot(data=d, aes(x=season, y=Dc, fill=season))+
  geom_bar(stat="identity",show.legend=F)+
  geom_errorbar(aes(ymin=CI.low, ymax=CI.high), width=0.5,)+
  # scale_y_continuous(limits=c(0,0.25))+
  ylab( TeX("Call density ($ calls \\cdot h^{-1} \\cdot 1000 \\cdot km^{-2}$)" ) )+
  xlab('')+
  labs(tag="(B)") + 
  theme_minimal()+
  theme(plot.tag = element_text(), 
                plot.tag.position = c(0.225, 0.975))

grid.arrange(gRate,gDen,ncol=1)


```
