---
title: "Main high-level RMarkdown script for estimating call density"
author: "Brian Miller"
date: "2022-10-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Background and method

This is a high-level script to estimate call densities from automated detections of whales in a single site-year dataset from the AAD.

This script is essentially a wrapper around code originally written by Danielle Harris and Franciele Castro.

$D_c = \frac{N_c*(1-c)}{kTp_a{\pi}w^2}$

where:

$D_c$ is call density

$N_c$ is number of calls

$c$ is false discovery rate

$k$ is number of sensors (here always 1)

$T$ is the duration of data analysed

$p_a$ is the probability of detection in the study area

${\pi}w^2$ is the study area (in km\^2)

## Define data files, subsets, and constants

This script uses as it's inputs a lot of files that are generated from the Matlab script prepareForDensityEstimation.m

```{r}
rm(list=ls())

# Source support functions for converting data from AcTUP, CaptureHistoryTable, 
# Matlab, etc. Also includes source function probabilityOfDetectionInArea.R
source('callDensitySupport.R') 

### Hard-coded parameters for each site (TODO: standardize file naming)
siteCode <- 'Kerguelen2014'
p <- read.csv(paste0(siteCode, '_callDensityParams.csv'))
snrLim <- c(-10,20)
pDetLim <- c(0,1)
seasons <- factor(c('summer', 'autumn', 'winter', 'spring', 'year'))
seasons <- factor(c('01','02','03','04','05','06','07','08','09','10','11','12'))
seasons <- factor(c('01','02','03','04','05','06',
                    '07','08','09','10','11','12',
                    'summer', 'autumn', 'winter', 'spring', 'year'))
```

Run the density estimation script for each season and year

## Density of blue whale ABZ calls in long-term moored recording dataset from `r siteCode`

### Before calculating call-density, explore which detection function is sensible

#### First check whether a GLM is sensible

```{r echo=TRUE, fig.height=12, fig.width=6, message=FALSE, warning=FALSE}

plotList <- list()
plotCount = 1;
for (season in seasons ){
  p <- defaultOutputFileNames(p,season)
  p$useGLM <- TRUE
  snrInfo <- capHist2snrInfo(p$capHistFile, season=season)
  gDetSnr <- showSNRdetectionFunc(fitSNRdetectionFunc(snrInfo, useGLM=p$useGLM) ) +
    ggtitle(paste(p$siteCode, season,'n=',dim(snrInfo)[1]) ) +
    xlim(snrLim) +
    ylim(pDetLim) +
    theme(text=element_text(size=10), #change font size of all text
        plot.title=element_text(size=10))
  
  plotList[[plotCount]] <- gDetSnr;
  plotCount =plotCount +1; 
}

gridExtra::grid.arrange(grobs=plotList,ncol=2)

```

#### Check that GAMs are sensible

```{r echo=TRUE,fig.height=12, fig.width=6, message=FALSE, warning=FALSE}

plotList <- list()
plotCount = 1;
for (season in seasons ){
  p <- defaultOutputFileNames(p,season)
  p$useGLM <- FALSE
  p$useSCAM <-FALSE
  p$numKnots <- 5

  snrInfo <- capHist2snrInfo(p$capHistFile, season=season)
  gDetSnr <- showSNRdetectionFunc(
    fitSNRdetectionFunc(snrInfo, useGLM=p$useGLM, useSCAM=p$useSCAM,
                        numKnots = p$numKnots) 
  ) +
    ggtitle(paste(p$siteCode, season,'n=',dim(snrInfo)[1]) ) +
    xlim(snrLim) +
    ylim(pDetLim) +
    theme(text=element_text(size=10), #change font size of all text
          plot.title=element_text(size=10))
  
  plotList[[plotCount]] <- gDetSnr;
  plotCount =plotCount +1; 
}

gridExtra::grid.arrange(grobs=plotList,ncol=2)


```

#### Finally, see if a constrained GAM is sensible

```{r echo=TRUE,fig.height=12, fig.width=6, message=FALSE, warning=FALSE}

plotList <- list()
plotCount = 1;
for (season in seasons ){
  p <- defaultOutputFileNames(p,season)
  p$useGLM <- FALSE
  p$useSCAM <-TRUE
  p$numKnots <- 5

  snrInfo <- capHist2snrInfo(p$capHistFile, season=season)
  gDetSnr <- showSNRdetectionFunc(
    fitSNRdetectionFunc(snrInfo, useGLM=p$useGLM, useSCAM=p$useSCAM,
                        numKnots = p$numKnots)
  ) +
    ggtitle(paste(p$siteCode, season,'n=',dim(snrInfo)[1]) ) +
    xlim(snrLim) +
    ylim(pDetLim) +
    theme(text=element_text(size=10), #change font size of all text
          plot.title=element_text(size=10))

  plotList[[plotCount]] <- gDetSnr;
  plotCount =plotCount +1;
}

gridExtra::grid.arrange(grobs=plotList,ncol=2)


```

Regular GAMs are problematic for some spring & full year, so either GLM or constrained GAM are the way to go here.

## Estimate $p_a$, probability of detection in area separately for each season

```{r echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}

for (season in seasons ){
  print(season) 
  p <- defaultOutputFileNames(p,season)
  p$outerloop <- 1000
  p$useGLM <- FALSE
  p$useSCAM <-TRUE
  p$numKnots<-5
  p$useSeparateFPdata <-TRUE
  
  # Call density estimation for season and parameters, p:
  cde(p,season)
}


```

## Load and plot call density estimates from `r siteCode` of Bm-Ant-ABZ

```{r, fig.height=6, fig.width=6.0, warning=FALSE, dev='ragg_png'}

d<- collect.density.results('.',p$siteCode)
inputs <- show.density.input.table(d,T)
results <- density.results.table(d)


# Separate monthly vs seasonal results for plotting
dSeason <- subset(d,subset =  (d$season=='summer'| d$season=='autumn' | 
                  d$season=='winter' | d$season=='spring' | d$season=='year') )
dMonth <-   subset(d,subset =  !(d$season=='summer'| d$season=='autumn' | 
                  d$season=='winter' | d$season=='spring' | d$season=='year') )

gRateSeason<- detection.rate.plot(dSeason)
gDenSeason <- density.plot(dSeason)
gRateMonth<- detection.rate.plot(dMonth)
gDenMonth <- density.plot(dMonth)
grid.arrange(gRateSeason, gRateMonth, gDenSeason, gDenMonth,ncol=2)


```

