---
title: "Call Density"
author: "Brian Miller"
date: "2022-10-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Density of blue whale D-calls (BmD) in long-term moored recording dataset from Kerguelen2014 Autumn

Apply methods of Castro, Harris, et al (in prep) to obtain call density.

$D_c = \frac{N_c*(1-c)}{kTP_a{\pi}w^2}$

where:

$D_c$ is call density

$N_c$ is number of calls

$c$ is false discovery rate

$k$ is number of sensors (here always 1)

$T$ is the duration of data analysed

$P_a$ is the probability of detection in the study area

${\pi}w^2$ is the study area (in km\^2)


Define some helper functions
```{r}

########## Testing purposes only - Remove/comment when using in production######
# rm(list=ls())
# p <- read.csv('casey2019_callDensityParams.csv')
# season <-'year'
# source('callDensitySupport.R')
# p<- defaultOutputFileNames(p,season)
################################################################################

source('callDensitySupport.R')

# Require that parameters have already been defined and are in data.frame, p
if (!exists("p")){
  warning(c('Required parameter variable, p, not found.
  Please define all call density parameters in data.frame, p, before running 
  this script.\n'))
}

if (!exists("season")){
  warning(c(  'Required parameter variable, season, not found.
  Please define a variable called season, and assign it a value of 
  summer, autumn, winter, spring, or year.\n\n
  Now assuming season<-year'))
  season <-'year'
}

```
Check inputs and create outputs
Store all parameters for call-density estimation in data frame called 'p'

```{r}

# Require that parameters have already been defined and are in data.frame, p
if (!exists("p")){
  warning(c(
  'Required parameter variable, p, not found.
  Please define all call density parameters in data.frame, p, before running 
  this script.\n'))
}

if (!exists("season")){
  warning(c(
  'Required parameter variable, season, not found.
  Please define a variable called season, and assign it a value of 
  summer, autumn, winter, spring, or year.\n\n
  Now assuming season<-year'))
  season <-'year'
}


```

### Estimate $c$ (False discovery rate)

False discovery rate, c, is defined as: FP/(FP+TP), where FP is the number of false-positives and TP is the number of true positives (https://en.wikipedia.org/wiki/Precision_and_recall; TODO: find a primary literature citation to use instead of citing wikipedia).

Here we estimate false-positive rate of the automated detector from the manually annotated dataset. The data-file for this is the capture-history table that contains reconciled annotated, and automated detections.

First, we count the number of FP that the AI produced from the capture history table. These are the sum of the rows where detect_t2==T & detect_t1==F. Then we calculate the number of true positives, the sum of the rows where detect_t2==T & detect_t1==T.

Alternatively, we could use the double observer mark-recapture to estimate the total number of calls in the dataset (see Miller et al 2022 - AI bests human observer - RSEC). This would require incorporating the Huggins/RMark script that can be found in the supplement to the Miller et al 2022 paper.

```{r}

effort <- readxl::read_excel(p$effortFile);
ch <- read.csv(file = p$capHistFile)
ch <- capHistTimeSeason(ch,season=season) 

if (season != 'year'){
  effort$season <- time2season(effort$StartTime)
  ix <- effort$season==season
  effort = subset(effort, subset=ix)
  ch <- ch[ch$season==season,]
} else {
  ch %>% 
    group_by(season) %>%
    dplyr::summarize(false.pos.num = sum(detect_t2 & !detect_t1),
                     true.pos.num = sum(detect_t2 & detect_t1),
                     c = false.pos.num/(false.pos.num+true.pos.num) 
                    )
}


FP <- sum(ch$detect_t2 & !ch$detect_t1)
TP <- sum(ch$detect_t2 & ch$detect_t1)
c <- FP/(FP+TP)
c # Using all detections, so full year results

```
The above calculation uses the annotated library, and requires that you have defined effort periods for which you've measured the false discovery rate.

Below is an alternative measure of false-discovery-rate where the analyst has inspected every nth detection to determine if it is true or false positive.



```{r}

if (p$useSeparateFPdata){
  ffp <- readxl::read_xlsx(p$manualFPfileName,sheet = 'conference')
  
  ffp$dt0 <-lubridate::ymd_hms(ffp$UTC)
  ffp$season <- time2season(ffp$dt0)
  if (season!= 'year'){
    ffp<- ffp[ffp$season==season,]
    fp<-  ffp %>% 
      mutate(date_col = season ) %>%
      group_by(date_col) %>%
      dplyr::summarize(detection.num = n(),
                       true.pos.num = sum(`True positive rate`),
                       false.pos.num = detection.num-true.pos.num,
                       c = false.pos.num/(false.pos.num+true.pos.num) 
      )
    
    c <- fp$c
  } else {
    c = sum(!ffp$`True positive rate`)/dim(ffp)[1]
  }
}

```


### $CV_c$ (Cochran approx.);

First calculate using annotated library

```{r}

var.wtd.mean.cochran <- function(x,w){
library(Hmisc,quietly = T, warn.conflicts = F)
  ix <- !is.na(x)
  x <- x[ix]
  w <- w[ix]
# Computes the variance of a weighted mean following Cochran 1977 definition
  n = length(w)
  xWbar = wtd.mean(x,w, na.rm = TRUE)
  wbar = mean(w, na.rm=TRUE)
  out = n/( (n-1)*sum(w)^2)*
    (sum((w*x-wbar*xWbar)^2)-2*xWbar*
       sum((w-wbar)*(w*x-wbar*xWbar))+xWbar^2*sum((w-wbar)^2)
    )
  return(out)
 }

fp <- read.csv(p$hourlyFalsePosFile)
fp$season <- time2season(lubridate::dmy_hms(fp$dt0))
if (season != 'year'){
  fp<- fp[fp$season==season,]
}

fp$fdr <- fp$num_fp/fp$num_detections
x<-fp$fdr # false discovery rate
wgts<-as.numeric(fp$num_detections) # w = n. of detections
# str(x)
# str(wgts)
# wtd.mean(x,wgts,na.rm=T)
var.wtd.mean.cochran(x,wgts)


# D.Harris: "Then, from the variance, you can estimate the CV for each c estimate. 
#This would be the square root of the variance, divided by the square root of 
#the number of samples (so estimate the standard error, then divide by the
#mean false discovery rate, to get the CV."

SEc=sqrt(var.wtd.mean.cochran(x,wgts))/sqrt(sum(!is.na(x)))
SEc

CV.c=SEc/wtd.mean(x,wgts)
CV.c
```

Now calculate CV of c where the analyst has inspected every 50th detection.

Here we group the detections by day within each season in order to estimate CVs.

```{r}
if (p$useSeparateFPdata){
  ffp <- readxl::read_xlsx(p$manualFPfileName,sheet = 'conference')

ffp$dt0 <-lubridate::ymd_hms(ffp$UTC)
ffp$season <- time2season(ffp$dt0)
if (season!= 'year'){
  ffp<- ffp[ffp$season==season,]
}
fp<-  ffp %>% 
    mutate(date_col = lubridate::date(dt0) ) %>%
             group_by(date_col) %>%
    dplyr::summarize(detection.num = n(),
                     true.pos.num = sum(`True positive rate`),
                     false.pos.num = detection.num-true.pos.num,
                     c = false.pos.num/(false.pos.num+true.pos.num) 
                    )

x <- fp$c
wgts <- fp$detection.num

SEc=sqrt(var.wtd.mean.cochran(x,wgts))/sqrt(sum(!is.na(x)))
SEc

CV.c=SEc/wtd.mean(x,wgts)
CV.c
}
```

### Calculate $T$ (time of deployment)

Sum the duration of the audio that has been analysed

```{r}

wavInfo <- read.csv(p$fullYearEffortFile) # File generated from Matlab wavFolderInfo
wavInfo$season <- time2season( mat2Rdate(wavInfo$startDate) )

if (season != 'year') {
  ix <- wavInfo$season==season
  wavInfo <- subset(wavInfo, ix)
}


T <- sum(wavInfo$duration)/3600 # in hours
T

```


### $P_a$ (Overall probability of detection)

```{r, echo=TRUE, result='hide', message=F, warning=F, cache=TRUE}

SNRinfo <- capHist2snrInfo(p$capHistFile,season)
NL <- SNRinfo %>% summarise(mean=mean(NoiseRL), sd=sd(NoiseRL), sampleSize=n())
SL <- data.frame(mean=p$SLmean, sd=p$slStd, sampleSize=p$SLsamplesize); 
TL<-read.csv(p$tlFile)
pDetInArea(SNRinfo, SL, TL,  NL, # Sonar equation inputs
           p$transectFile, p$simResultsFile, p$paFile, # file output names
           useGLM=p$useGLM, useSCAM=p$useSCAM, numKnots=p$numKnots, # detector SNR curves
           output.resolution.m=p$output.resolution.m, outerloop=p$outerloop)

```

```{r}
pa.all.transects <- read.csv(p$paFile,header = T, sep = ' ');
no.transects <- dim(pa.all.transects)[1]-1;
pa <- pa.all.transects[no.transects+1,1]
pa

# CV.pa <- pa.all.transects[9,2]/pa.all.transects[9,1]

```

### $CV_{P_a}$ (see Harris, 2012 for details);

Approach is to calculate CV using mean and SD from each radial?

```{r}
# Coef. of variation - Pa

names(pa.all.transects)<-c('Mean','SD')


# Radials from 0-315 degrees are in rows 1:8
y=pa.all.transects[1:no.transects,]


## IMPORTANT: to check if the var, SE and CV calculation is correct ##
## Per year:

varp.pa<-((sd(y$Mean)/sqrt(no.transects))^2)+((sum(y$SD)/no.transects)^2) 
sep.pa<-sqrt(varp.pa)
CV.pa<-sep.pa/mean(y$Mean)
CV.pa

```
### Count $N_c$ (number of detections).

$T$ and $N_c$ were calculated based, respectively, in the raw data duration and number of automatic detections correspondent to the full dataset.

```{r}
library(ggplot2, quietly=T, warn.conflicts = F)

det <- read.csv(p$fullYearDetectionCsv,sep=',')
det$t <- as.POSIXct(det$UTC, tz='UTC')
det$season <- time2season(det$t)
if (season !='year'){
  ix <- det$season==season
  det<-subset(det,ix)
} else {
  ggplot(data=det, aes(x=season))+
  stat_count() 
}
det$month <- lubridate::month(det$t)
Nc <- dim(det)[1]
Nc

```

### $CV_{N_c}$ (see Harris, 2012 for details);

```{r, warning=FALSE}
varn=Nc*(pa)*(1-pa)
varn

# Confirm how to calculate SD and CV (?)

SDn=sqrt(varn)
SDn
CV.Nc=SDn/Nc
CV.Nc


```
### Finally, estimate Dc (Density of calls) [calls $h^{-1} km^{-2}$]

```{r}

Dc <- (Nc * (1-c) )/( p$k * pi*p$w^2 * pa * T )
Dc     # per h per km^2
Dc*1e3 # per h per 1000 km^2?


```

### CV_Total (Delta Method)

```{r}

CV.Dc<-sqrt((CV.Nc^2)+(CV.pa^2)+(CV.c^2))
CV.Dc

```
## Collate into data frame and write to csv file
```{r}


result <- data.frame(p$siteCode,Nc,c,p$k,T,p$w,pa,
                     p$SLmean,p$SLsd,NL$mean,NL$sd,p$useGLM,
                     CV.Nc,CV.c,CV.pa,Dc,CV.Dc)
names(result) <- c('siteCode','Nc','c','k','T','w','pa',
                     'SLmean','SLsd','NLmean','NLsd','useGLM',
                     'CV.Nc','CV.c','CV.pa','Dc','CV.Dc')

write.csv(result, file = p$densityResultsFile, row.names = season)
result


```
