---
title: "callDensity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{callDensity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.height = 3.75, 
  comment = "#>"
)
```

```{r setup}
library(callDensity)
library(ggplot2)
library(kableExtra)
```

#Simple test of callDensity R package

Create a synthetic dataset with known (specified) values for all parameters for testing the callDensity R package which implements the methods described by (@Castro2024). This package implements a solution for call density estimation from single sensors (hydrophones) via the density equation
    
\begin{equation}
D_c = \frac{N_c(1-c)}{kTp_aA}
\end{equation}

where: $D_c$ is call density, $N_c$ is number of calls, $c$ is false discovery rate, $k$ is number of sensors (here always 1), $T$ is the duration of data analysed (in hours), $p_a$ is the probability of detection in the study area, and $A$ is the study area (in km\^2).

##1) Generate known call distribution in space.

Start with _n_ calls within radius _R_ of the hydrophone. This should produce a known call density with a uniform random distribution in a circle around the hydrophone. 

```{r call.spatial.distribution, warning=FALSE, message=FALSE}
# Pseudo-code to check callDensity 

# 1) Generate a distribution of n calls within radius R of the hydrophone with a
# known call density Uniform random in circle

# For ABW ABZ calls, 1 whale call every 10 s would be a maximum of 3.1536M calls
# In a continuous year of recording:  365*24*60*60/10 = 3.1536e6 

n = 1e5; # 100k calls; 
R = 1e6; # radius of 1000 km 



# Call density, D_c is n/A 
A = pi*(R/1e3)^2 # Circular study area in km^2
TrueCallDensity = n/A # calls/km^2

k <- 1      # number of sensors
Time <- 1   # years

print(paste(TrueCallDensity,"calls per km^2 per unit time"))

r = R*sqrt(runif(n));
t = 2*pi*runif(n);
sim <- data.frame(x = r*cos(t), y = r*sin(t), groundTruth=1);

# Date and time of detections span 1 full year (calendar year of)

minDate <- as.POSIXct("2025-01-01")
maxDate <- as.POSIXct("2025-12-31")

epoch <- "1970-01-01"
sim$datetime <- as.POSIXct(pmax(runif(n, minDate, maxDate), runif(n, minDate, maxDate)), origin = epoch, tz="UTC")

ggplot(data=sim, aes(x=x/1e3,y=y/1e3))+
  geom_point(colour = "black",alpha = 0.1, size=0.1)+
  coord_equal()+
  xlab("X location (km)")+
  ylab("Y location (km)")



```

## 2) Check distances from that uniform distribution are sensible (triangle distribution)

```{r call.distance.distribution.check,  warning=FALSE, message=FALSE}

sim$d <- sqrt(sim$x^2+sim$y^2)
ggplot(data=sim, aes(x=d/1e3))+
  geom_histogram(binwidth = 10)+
  xlab("Distance (km)")

```

## 3) Assign SL,NL, and TL to each call

Use SL from studies of ABW SL_ABW: mean=190, sd=3 dB re 1 uPa

#### Assign NL to each call from plausible distribution of NL:

NL: mean= 90, sd=4 dB re 1 uPa

#### Assign TL to each call using spherical spreading propagation model:

TL = 20 log10 d

```{r sonar.eq.properties}

SL <- data.frame(mean=190, sd=3, sampleSize=350)

sim$SL <- rnorm(n, mean=SL$mean, sd=SL$sd)

NL = data.frame(mean=90, sd = 4, sampleSize = n)
sim$NL <- rnorm(n, mean=NL$mean, sd=NL$sd)

range_m = seq(from=5, to=1e6, by=100)

# Spherical spreading for a single transect
tlTransectSpherical <- 20*log10(range_m)


tlTransect <- tlTransectSpherical
sim$TL <- 20*log10(sim$d)

TL <- data.frame(
  range_m,
  tl000 = tlTransect,
  tl090 = tlTransect,
  tl180 = tlTransect,
  tl270 = tlTransect
)
```

```{r, include=FALSE, eval=FALSE}

# # NOT RUN & NOT INCLUDED IN RMARKDOWN 

ggplot(data=TL, aes(x=range_m/1e3, y=tl000))+
  geom_point(alpha=0.1,size=0.1)+
  scale_x_log10(limits=c(0.1,1000),
                breaks=10^(-1:7),
                minor_breaks = rep(1:9,6)*(10^rep(-1:7,each=9)))+
  xlab("Distance (km)")+
  ylab("TL (dB)")+
  annotation_logticks(sides='') +
  theme_bw()

```

```{r surface.duct.TL, include=FALSE, eval=FALSE}

# NOT RUN & NOT INCLUDED IN RMARKDOWN 

# Model transmission losses, TL as a simple surface duct: spherical spreading
# out to transition range r0, and then cylindrical spreading beyond r0.
tlType='surfaceDuct'

# Simple surface duct model for a single transect:
#  TL= 20*log10                  } d < r0;
#  TL= 10*log10(ro)+10*log10(r0) } d>= r0;
r0 <- 10000
tlSurfaceDuct <- function (r){
  tl<- ifelse(r<r0, 20*log10(r), 10*log10(r0) + 10*log10(r))
  return(tl)
}
tlTransectSurfaceDuct <- sapply(range_m, tlSurfaceDuct)
tlTransect <- tlTransectSurfaceDuct
sim$TL = sapply(sim$d,tlSurfaceDuct)

TL <- data.frame(
  range_m,
  tl000 = tlTransect,
  tl090 = tlTransect,
  tl180 = tlTransect,
  tl270 = tlTransect
)

ggplot(data=TL, aes(x=range_m, y=tl000))+
  geom_point(alpha=0.1)+
  scale_x_log10(limits=c(0.1,100),
                breaks=10^(-1:6),
                minor_breaks = rep(1:9,6)*(10^rep(-2:6,each=9)))+
  xlab("Distance (km)")+
  ylab("SNR (dB)")

ggplot(data=sim, aes(x=d/1e3, y=TL))+
  geom_point(alpha=0.1)+
  scale_x_log10(limits=c(0.1,100),
                breaks=10^(-1:6),
                minor_breaks = rep(1:9,6)*(10^rep(-2:6,each=9)))+
  xlab("Distance (km)")+
  ylab("SNR (dB)")



```


## 6) Calculate SNR for each call

Passive sonar equation (in dB): SNR = SL-TL-NL

```{r call.snr.check,  warning=FALSE, message=FALSE}

sim$snr <- with(sim,SL-TL-NL)

ggplot(data=sim, aes(x=snr))+
  geom_histogram()+
  xlab("SNR (dB)")

ggplot(data=sim, aes(x=d/1e3, y=snr))+
  geom_point(alpha=0.1,size=0.1)+
  geom_density_2d()+
  xlab("Distance (km)")+
  ylab("SNR (dB)")+
  scale_x_log10(limits=c(0.01,1000),
                breaks=10^(-1:7),
                minor_breaks = rep(1:9,6)*(10^rep(-2:7,each=9)))+
  annotation_logticks(sides='') +
  theme_bw()



```

## Simulate detection process

Here's where things start to get tricky. We need to make some assumptions about the behaviour of the detector. Generally, it is assumed the detector has a high probability of detection at high SNR, and a low(er) probability of detection at low SNR. In quantitative sonar performance modelling sometimes a step-function or logistic curve is used when an analytical expression is required. If ground truth data on the detectors performance are available, these models can be fit to the detection data.

In addition to the probability of detection, there is also a probability of false alarm (false positive predictions) that likely needs to be addressed. There can be multiple mechanisms that can produce false positives. For example, they could be triggered by intense high-SNR sound in the same band as the signal, but from a different source. For some signal types, false positives can also arise from small deviations in ambient noise (producing false positive detections that contain very low SNR). Again, if ground truth data are available then the relationship between false positive detections and false-positive power to noise ratio (SNR of false positives) can be modeled in a similar manner to that of probability of detection.

#### Detector 1

Assume probability of detection, p_det, for detector 1 follows a logistic curve in SNR with location (p_det=0.5) at an SNR of 3 dB.

Assume detector 1 does not produce false positives.

```{r simulate.detector, warning=FALSE, message=FALSE}

det1params = data.frame( #detector 1 parameters
  location=2,   # AKA intercept?
  scale=2,      # AKA slope?
  func='plogis', # logistic function
  c=0.1           # False discovery rate
)

# Probability of detection for each simulated call
sim$p_det = plogis(sim$snr,
                   location=det1params$location,
                   scale=det1params$scale)


# Spatial distribution
p0 <- ggplot(data=sim, aes(x=x/1e3, y=y/1e3, colour=detected_observer1))+
  geom_point(alpha=0.1,size=0.1)+
  coord_equal()+
  xlab("X location (km)")+
  ylab("Y location (km)")

# Apply Bernoulli trial to SNRs to see which were detected
sim$detected_observer1 <- as.logical(rbinom(n,size=1,prob=sim$p_det))

# View spatial, distance, and SNR distributions of detections vs non-detections

# SNR Distribution
ggplot(data=sim, aes(x=snr, group=detected_observer1, fill=detected_observer1) )+
  geom_histogram()

ggplot(data=sim, aes(x=d/1e3, y=snr, colour=detected_observer1))+
  geom_point(alpha=0.1, size=1)+
  xlab("Distance (km)")+
  ylab("SNR (dB)")+
  scale_x_log10(limits=c(0.1,1000),
                breaks=10^(-1:7),
                minor_breaks = rep(1:9,6)*(10^rep(-2:7,each=9)))+
  annotation_logticks(sides='') +
  theme_bw()


```

## Simulate false positives

```{r simulate.false.positives, warning=FALSE, message=FALSE}

set.seed(2)

# Calculate number of false positives given the number of true positive
# detections for the detector and the specified false discovery rate, c, for
# that detector.
n_tp_det1 <- sum(sim$detected_observer1) 
n_fp_det1 <- n_tp_det1/(1-det1params$c)-n_tp_det1

# Store false positives in data.frame with same format as sim
fp <- data.frame(matrix(ncol=length(sim),nrow=n_fp_det1))
colnames(fp) <- colnames(sim)

# groundTruth=0 means that these are false positives (handy to keep in mind for
# when we include these in sim)
fp$groundTruth <- 0
fp$detected_observer1 <- 1

# Generate the right number of false positives uniformly over same time period
# as true positives.
epoch <- "1970-01-01"
fp$datetime <- as.POSIXct(pmax(runif(n_fp_det1, minDate, maxDate), 
                               runif(n_fp_det1, minDate, maxDate)),
                          origin = epoch, tz="UTC")

# False positives don't have a location, SL, or TL, but do have NL and SNR
# Use same distribution of NL as true positives
fp$NL <- rnorm(n_fp_det1, mean=90, sd=4)

# SNR distribution of false positives broadly matches that of Casey2019 human
# analysts
fp$snr <- rnorm(n_fp_det1, mean=1, sd=2) 

simWithFP <- rbind(sim,fp)

# We've generated the right number of false positive detections now and merged
# these into our simulation. 
# But the false positives are missing locations, distances, SL, and p_dets. Not
# sure that this actually matters though.

simWithFP$group <- factor(ifelse(simWithFP$groundTruth,
                              ifelse(simWithFP$detected_observer1, 
                                     "TruePositive","FalseNegative"),
                              "FalsePositive"),
                          levels=c("FalsePositive","TruePositive","FalseNegative")
                          )

# SNR Distribution
ggplot(data=simWithFP, aes(x=snr, group=group, fill=group) )+
  geom_histogram(position = position_stack())

ggplot(data=simWithFP, aes(x=snr, group=group, fill=group) )+
  geom_histogram(position = position_identity())+
  facet_wrap(group~., ncol=1,scales="free_y")

simWithFP <- simWithFP[order(simWithFP$datetime),]

```

## 7) Subsample data for detector characterisation

Format results so that they can be used by the callDensity package for testing.

```{r subsample.for.detector.characterisation}

n_subsample <- 30e3

subsample <- simWithFP[sample(nrow(simWithFP), n_subsample),]

n_fp_subsample <- sum(subsample$groundTruth==FALSE)

# False discovery rate is number.false.positives/number.predicted.positive
c_subsample <- n_fp_subsample/sum(subsample$detected_observer1)

```

```{r format.output.for.callDensity.package}


SNRinfo <- with(subsample,
                data.frame(
                  Detected = detected_observer1,
                  CallRL   = SL-TL,
                  NoiseRL  = NL,
                  SNR      = snr,
                  t        = datetime,
                  season   = callDensity::time2season(datetime)
                )
)

NLsamp <- callDensity::nlFromSnrInfo(SNRinfo)
truncationDistances <- R 

```

## 8) Write results 
[not actually required]
## 9) Calculate call densities using callDensity package

```{r probability.of.detection.in.area, warning=FALSE, message=FALSE}


snrDetFun.glm <- callDensity::fitSNRdetectionFunc(SNRinfo,
                                              modelType = "glm")
snrDetFun.gam <- callDensity::fitSNRdetectionFunc(SNRinfo,
                                              modelType = "gam",
                                              numKnots = 3)
snrDetFun.scam <- callDensity::fitSNRdetectionFunc(SNRinfo,
                                              modelType = "scam",
                                              numKnots = 5)


# View logistic function
p0<- ggplot(data=sim, aes(x=snr, y=p_det))+
  geom_point(alpha=0.1,color='red',size=0.1)

p1<-showSNRdetectionFunc(snrDetFun.glm)

gridExtra::grid.arrange(grobs=list(p0,p1))

Pa.glm <- callDensity::pDetInArea(snrDetFun.glm, SL=SL, TLlookup = TL, NL=NLsamp,
           transectFile = 'pDet_test_glm_year_transects.csv',
           simResultsFile = 'simResult_test_glm_year.txt',
           paFile= 'Pa_test_glm_year.txt',
           output.resolution.m=100,
           outerloop = 100,
           truncationDistance=1e6, # in m
           snrTruncationThreshold =-Inf)
pa.glm <- Pa.glm$overall


Pa.gam <- callDensity::pDetInArea(snrDetFun.gam, SL=SL, TLlookup = TL, NL=NLsamp,
           transectFile = 'pDet_test_gam_year_transects.csv',
           simResultsFile = 'simResult_test_gam_year.txt',
           paFile= 'Pa_test_gam_year.txt',
           output.resolution.m=100,
           outerloop = 100,
           truncationDistance=1e6, # in m
           snrTruncationThreshold =-Inf)
pa.gam <- Pa.gam$overall

Pa.scam <- callDensity::pDetInArea(snrDetFun.scam, SL=SL, TLlookup = TL, NL=NLsamp,
           transectFile = 'pDet_test_scam_year_transects.csv',
           simResultsFile = 'simResult_test_scam_year.txt',
           paFile= 'Pa_test_scam_year.txt',
           output.resolution.m=100,
           outerloop = 100,
           truncationDistance=1e6, # in m
           snrTruncationThreshold =-Inf)
pa.scam <- Pa.scam$overall



ggplot(data=Pa.glm$meanOfAllTransects, aes(x=range_m/1e3,y=pDet))+
  geom_point(alpha=0.1, size=0.1)+
  xlab("Range (km)")+
  ylab("Probability of detection\n(Mean across all transects)")+
  scale_x_log10(limits=c(0.1,1000),
                breaks=10^(-1:7),
                minor_breaks = rep(1:9,6)*(10^rep(-2:7,each=9)))+
  annotation_logticks(sides='') +
  theme_bw()

ggplot(data=Pa.gam$meanOfAllTransects, aes(x=range_m/1e3,y=pDet))+
  geom_point(alpha=0.1, size=0.1)+
  xlab("Range (km)")+
  ylab("Probability of detection\n(Mean across all transects)")+
  scale_x_log10(limits=c(0.1,1000),
                breaks=10^(-1:7),
                minor_breaks = rep(1:9,6)*(10^rep(-2:7,each=9)))+
  annotation_logticks(sides='') +
  theme_bw()


ggplot(data=Pa.scam$meanOfAllTransects, aes(x=range_m/1e3,y=pDet))+
  geom_point(alpha=0.1, size=0.1)+
  xlab("Range (km)")+
  ylab("Probability of detection\n(Mean across all transects)")+
  scale_x_log10(limits=c(0.1,1000),
                breaks=10^(-1:7),
                minor_breaks = rep(1:9,6)*(10^rep(-2:7,each=9)))+
  annotation_logticks(sides='') +
  theme_bw()

```

```{r}

Nc <- sum(sim$detected_observer1)

# Density estimation equation
Dc.glm <- (Nc * (1 - c_subsample) )/( k * A * pa.glm * Time )
Dc.gam <- (Nc * (1 - c_subsample) )/( k * A * pa.gam * Time )
Dc.scam <- (Nc * (1 - c_subsample) )/( k * A * pa.scam * Time )


```

## Results

Check the effects

```{r Call.Density.Results, echo=TRUE}

results <- data.frame(Dc = TrueCallDensity, SampleSize = n, A = A,   # Environment 
                      NLmean = NL$mean, NLsd = NL$sd,       # Noise
                      # Detector params
                        model=det1params$func, intercept=det1params$location,  
                        slope = det1params$scale, c = det1params$c, 
                        Pa = mean(sim$p_det), 
                      row.names='Truth')

snrDetFun.glm$r.sq <- with(summary(snrDetFun.glm), 1 - deviance/null.deviance)
results<-rbind(
           results, 
           data.frame(Dc=Dc.glm, SampleSize=n_subsample, A = A,           # Environment
                      NLmean = NLsamp$mean, NLsd = NLsamp$sd,# Noise
                      # Detector params
                        model='glm', intercept=snrDetFun.glm$r.sq,
                        slope =AIC(snrDetFun.glm), c=c_subsample, Pa=pa.glm,  
                      row.names = 'Estimate1'))

results<-rbind(
           results, 
           data.frame(Dc=Dc.gam, SampleSize=n_subsample, A = A,           # Environment
                      NLmean = NLsamp$mean, NLsd = NLsamp$sd,# Noise
                      # Detector params
                        model='gam', intercept=summary(snrDetFun.gam)$r.sq,
                        slope =AIC(snrDetFun.gam), c=c_subsample, Pa=pa.gam,  
                      row.names = 'Estimate2'))

results<-rbind(
           results, 
           data.frame(Dc=Dc.scam, SampleSize=n_subsample, A = A,           # Environment
                      NLmean = NLsamp$mean, NLsd = NLsamp$sd,# Noise
                      # Detector params
                        model='scam', intercept=summary(snrDetFun.scam)$r.sq,
                        slope =AIC(snrDetFun.scam), c=c_subsample, Pa=pa.scam,  
                      row.names = 'Estimate3'))



kableExtra::kbl(results, digits = c(4,3,0,1, 2, NA, 3, 3, 3, 4), 
                col.names= c('Dc','(sub)Sample Size', 'A','NL_mean','NL_sd',
                             'Model','R^2','AIC','c','P_a')) %>% 
  kableExtra::kable_classic(full_width=FALSE)

```
