---
title: "callDensity: Two detector simulation"
output: 
  html_document: default
  pdf_document: default
  rmarkdown::html_vignette:
    vignette: >
      %\VignetteEncoding{UTF-8}
      %\VignetteIndexEntry{callDensity: Two detector simulation}
      %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.height = 2, 
  comment = "#>"
)
```

```{r setup}
library(callDensity)
library(ggplot2)
library(scales)
library(kableExtra)
```

# Two-detector test of callDensity R package

Building on the foundation provided by the first vignette, here we attempt to make our simulation even more realistic.

In our first simulation, we simulated a uniform distribution of Antarctic blue whale calls around a hydrophone in a homogenous environment with spherical spreading. We also simulated an automated detector that operated according to a logistic function of SNR. We also simulated false positive detections for the automated detector as an independent normal-distribution in SNR.

We then took a subset of all the calls to simulate annotation of a subset of the data for characterisation of the detectors performance. Specifically, we used the callDensity package to estimate the probability of detecting a single call as a function of SNR from the subset. In our first simulation, we modelled this probabilty of detection using the known ground-truth from the simulation.

In the work of @castro2024 an expert human observer provides the ground truth annotations. However, not every expert human observers can be "practically perfect in every way," and some analysts might fall slightly short of that Mary Poppins level of perfection.

In reality, we cannot expect to obtain perfect ground truth data. A more realistic expectation is that we will have detections from more than one observer. Here the term observer includes both algorithms and human observers.

Our goal in this simulation is to explore and understand how imperfect estimation of the ground truth affects our ability to estimate call density.

Again, we start with the density equation:

\begin{equation}
D_c = \frac{N_c(1-c)}{kTp_aA}
\end{equation}

where: $D_c$ is call density, $N_c$ is number of calls, $c$ is false discovery rate, $k$ is number of sensors (here always 1), $T$ is the duration of data analysed (in hours), $p_a$ is the probability of detection in the study area, and $A$ is the study area (in km\^2).

## Generate known call distribution in space and time.

Start with *n* calls within radius *R* of the hydrophone. This should produce a known call density with a uniform random distribution in a circle around the hydrophone. Assuming the recorder is located at centre of study area then distances are the magnitude of the x & y location. A uniform density and increasing area with larger distances should yield a triangular distance distribution.

```{r call.spatial.distribution, warning=FALSE, message=FALSE}

# 1) Generate a distribution of n calls within radius R of the hydrophone with a
# known call density Uniform random in circle

n = 1e6; # number of simulated calls (regardless of whether detected or not); 
R = 1e6; # radius in m (i.e. 1e6=1000km)
k <- 1   # number of sensors (always 1 for single hydrophone moorings)
minDate <- as.POSIXct("2025-01-01")
maxDate <- as.POSIXct("2026-01-01")
Time <- as.numeric(difftime(maxDate,minDate,unit="days"))/(365) # in years

# Call density, D_c is n/A 
A = studyArea(R/1e3) # Circular study area in km^2
TrueCallDensity = n/(A*Time) # calls/km^2/time

set.seed(1)
# 1) Generate n uniformly distributed calls within radius R and time period Time
sim <- simCallLocation(n=n, R=R, minDate=minDate, maxDate=maxDate)
cat('Study area (km^2): ', A, '\n')
cat('Number of calls: ', n ,'\n');
cat('Time units:', Time, '\n')
cat('True call density (calls per km^2 per unit time): ',TrueCallDensity, '\n')


```

## Assign sonar equation parameters to each call (SL,NL,TL)

### a) Source Level (SL)

Assume source level, SL, is that of Antarctic blue whale song (as in @castro2024), a mean of 190 dB re 1 uPa.However, to keep things clearer in this simple example, we use a standard deviation of 4 dB, which is lower than that used by @castro2024. These values match reasonably well with those that have been estimated for this animal mean of 189 dB and standard deviation of 3-8 dB. Realistic values should help make our simulation better match reality.

### b) Noise Level (NL)

We assign NL to each call using a similarly realistic parameters for the distribution of NL.

Our NL distribution resembles the same real-world measurements used by @castro2024 that were derived from noise samples in the same band, but adjacent in time to the Antarctic blue whale calls they detected. Again, this helps ensure our acoustic environment is somewhat grounded in reality and enhances the plausibility and realism of our simulation.

### c) Transmission Loss (TL)

We assign TL to each call using a simple spherical spreading propagation model:

\begin{equation}
TL = 20  log_{10}  d
\end{equation}

This TL model is a commonly used analytical expression for transmission loss. It is a gross oversimplification compared to real-world transmission losses. However, it has a sound theoretical basis that is readily interpretable (i.e. it's an inverse square law, which is common in physics). 

There has already been some investigations of how uncertainty in TL affects call density (e.g. Helble et al 2013), and TL variability and mismatch is not the focus of this vignette, so there's reasonable justification for keeping this part simple. 

Finally, with our acoustic properties assigned, we apply the passive sonar equation (in dB): \begin{equation}sim
SNR = SL-TL-NL
\end{equation}

```{r sonar.eq}

SL <- data.frame(mean=190, sd=4, sampleSize=350) # True SL distribution for sim
NL = data.frame(mean=84, sd = 4, sampleSize = n)  # True NL distribution for sim
tlFunc <- function(r)20*log10(r)

# Simulate the true acoustic properties of each call and false positive 
sim <- simCallAcoustics(sim, SL, NL, TL = tlFunc)

# Per-transect estimate of TL (used by callDensity package to estimate pDet)
TL <- simTLradials_20logR(maxRange=R, rangeStep=100, numTransects=4)

```

Here's the code from the callDensity package used to simulate sonar eq parameters for calls

`simCallAcoustics<-` 
```{r, code=simTLradials_20logR, eval=FALSE}
```
And the code for the fucntion to simulate TL

`simTLradials_20logR <- ` 
```{r, code=simTLradials_20logR, eval=FALSE}
```

## Simulate detection process

To simulate the detection process we need to make some assumptions about the behaviour of the detector.

Generally, if not by definition, a detector has a higher probability of detection at high SNR, and a lower probability of detection at low SNR. In quantitative sonar performance modelling sometimes a step-function or logistic curve is used when an analytical expression is required. If ground truth data on the detectors performance are available, these models can be fit to the detection data.

Assume probability of detection, p_det, for both detectors follows a logistic curve with SNR as the only independent variable. We can assign a location/intercept for this curve (i.e. location along x-axis where p_det=0.5). We can also assign a scale/slope of the curve to represent the steepness of the transition between low proability and high probability.

### Simulate false positives

In addition to the probability of detection, there is also a probability of false alarm (false positive predictions) that we would like to simulate as well.

There can be multiple mechanisms that can produce false positives. For example, they could be triggered by intense high-SNR sound in the same band as the signal, but from a different source. For some signal types, false positives can also arise from small deviations in ambient noise (producing false positive detections that contain very low SNR). Again, if ground truth data are available then the relationship between false positive detections and false-positive power to noise ratio (SNR of false positives) can be modeled in a similar manner to that of probability of detection.

Here we assume false positive detections are distributed around some mean low SNR. This means that false positives in the 'ground truth' detector can affect the pDet\~SNR curve/function.

### Parameterise detectors

```{=html}
<style type="text/css">
  .mytable {width: 50%;}
</style>
```

Summary of detectors:

::: mytable
| Detector | Description | Location/SNR 'threshold' | Scale/slope | False positive rate |
|--------------|-----------------|--------------|--------------|--------------|
| Detector 1 | Reliable and good detector, meant to emulate a human analyst. | Lower | Same | Lower |
| Detector 2 | Automated detector, meant to emulate a signal processing algorithm. | Higher | Same | Higher |
:::

```{r parameterise.detectors, warning=FALSE, message=FALSE}

#Specify parameters for detector 1
# Good detector with low threshold (location) and low false positive rate.
det1params = data.frame( 
  location=1,    # AKA intercept?
  scale=2,       # AKA slope?
  func='plogis', # logistic function
  c=0.1,         # False discovery rate
  fpMean=0,      # Mean of distribution of false positives (in dB SNR)
  fpSD = 2      # Standard deviation false positive distribution (in dB SNR)
)

# Detector 2 has a higher threshold (location), scale, and false positive rate
# than detector 1.
det2params = data.frame( 
  location=2,    # AKA intercept?
  scale=4,       # AKA slope?
  func='plogis', # logistic function
  c=0.3,         # False discovery rate
  fpMean=0,      # Mean of distribution of false positives (in dB SNR)
  fpSD = 2      # Standard deviation false positive distribution (in dB SNR)
)
```

Now we create a function that will take the detector parameters above and simulate the detection process. We then call this function with each set of parameters to create a simulation for each detector.

```{r simulate.detectors, warning=FALSE, message=FALSE}



# Simulate two detectors each with the different parameters
simDet1<- simulateDetector(det1params,sim)
simDet2<- simulateDetector(det2params,sim)

```

Code used to simulate the detection process:
`simulateDetector <- `
```{r, code=simulateDetector, eval=FALSE}
```


View the spatial detection density (which is not the same as spatial call density since it accounts for neither detection probability nor time).

```{r detector.plots, fig.height=2, fig.width=6, message=FALSE, warning=FALSE}

plotSpatialDetections <- function(sim){
  
  # Spatial distribution (excludes false positives)
  ggplot(data=sim, aes(x=x/1e3, y=y/1e3, weight=detect_table) )+
    geom_bin_2d(alpha=1,binwidth=c(10,10))+
    coord_equal()+
    xlab("X location (km)")+
    ylab("Y location (km)")
}

sp1 <- plotSpatialDetections(simDet1)+ggtitle('Detector 1')
sp2 <- plotSpatialDetections(simDet2)+ggtitle('Detector 2')
gridExtra::grid.arrange(sp1,sp2,nrow=1)

dist1 <- plotDetectionDistribution(simDet1)+ggtitle('Detector 1')
dist2 <- plotDetectionDistribution(simDet2)+ggtitle('Detector 2')
gridExtra::grid.arrange(dist1,dist2,nrow=1)

```

## Subsample data for detector characterisation

@castro2024 subsample approximately 200 hours evenly spaced throughout the year to create detector characterisation curves (detection vs SNR). Here we subsample 1 hour in every 41 hours yielding 214 subsamples that are 1 hour in duration.

```{r subsample.for.detector.characterisation}

subsampleDet1 <- subsampleSimInTime(simDet1,interval = "41 hour")
subsampleDet2 <- subsampleSimInTime(simDet2,interval = "41 hour")

# Plot to validate that we've subsetted the data sensibly by time.
ggplot(subsampleDet1, aes(x=datetime, weights=detect_table) )+
  # geom_point(size=1,alpha=0.1)
  geom_histogram(breaks=seq(from=minDate,to=maxDate,by='12 h'))

ggplot(subsampleDet2, aes(x=datetime, weights=detect_table) )+
  # geom_point(size=1,alpha=0.1)
  geom_histogram(breaks=seq(from=minDate,to=maxDate,by='12 h'))

```

Code used to implement subsampling:
`subsampleSimInTime <- `
```{r, code=subsampleSimInTime, eval=FALSE}
```

## Detection matching between detector 1 and 2

Both detectors (simulated subsets) operate on the same set of calls (i.e. they are different subsets of the same underlying simulation). So we can find detections from one detector that match those of the other detector by merging the detections into a capture history table. We accoomplish this with by a full-outer join of the tables with the timestamp/datetime, as the key/matching-variable.

NB: Real world detections from independent observers rarely have timestamps that match exactly, so real data would require matching criteria that can accomodate some amount of error in detection times, locations, and other parameters.

Presently, false positives have randomly generated date-times, so have a negligible chance of matching between the two detectors (i.e. false positives on one detector are independent from the other). In reality false-positives are unlikely to be fully independent, so there is an opportunity to improve the realism of the simulation by estimating correlation in false positives for real datasets, and then simulating that relationship here.  

*TODO: Create false positives in a way that better approximates their real-world occurrence (i.e. in a way that allows them to be matched between detectors).*

```{r match detections from detector1 and detector2}

capHistTab <- simsTocaptureHistoryTable(subsampleDet1, subsampleDet2)

```

Code used to merge simulated detections into a capture history table:
`simsTocaptureHistoryTable <- `
```{r, code=simsTocaptureHistoryTable, eval=FALSE}
```

Convert this capture history table into an SNRinfo table, and treat detector1 as ground truth.

```{r format.output.for.callDensity.package}

# SNRinfo is used to calculate snrDetFun for detector2 assuming detector1 is the
# ground truth. This is an observer ground (OG) so includes false positives from
# detector1, but does not include detections from detector2 not detected on
# detector1
SNRinfo <- capHistTosnrInfo(capHistTab)

# Positive on detector 2, but not on 1
n_fp_subsample <- sum(capHistTab$detect_table2 & !capHistTab$detect_table1)
n_p_subsampleDet2 <- sum(capHistTab$detect_table2) 

# False discovery rate is number.false.positives/number.predicted.positive
c_subsample <- n_fp_subsample/n_p_subsampleDet2
n_subsample <- dim(SNRinfo)[1]  # Number positive detections in SUBset

truncationDistances <- R 

# Number positive detections in FULL set
Nc <- sum(simDet2$detect_table)       

# Noise level from subsampled Observer ground-truth
NLsamp <- SNRinfo %>% dplyr::summarise(mean=mean(NoiseRL,na.rm = TRUE),
                             sd=sd(NoiseRL,na.rm = TRUE),
                             sampleSize=dplyr::n()-sum(is.na(NoiseRL)))

NLtable <- rbind(NL,NLsamp)
rownames(NLtable)<- c('Actual','Estimated from subsample')
kable(NLtable,caption = 'Noise level distributions')%>% 
  kableExtra::kable_classic(full_width=FALSE)

```

## Calculate call densities using callDensity package

### GLM fit to SNR detection function

The detector was a logistic function, so fitting a GLM should provide good results.

A GAM or SCAM should also be able to provide a good fit to a logistic function, but to keep this concise we leave fitting those as an exercise for the reader (see callDensity vignette).

The callDensity package provides a convenience function fitSNRdetectionFunc to fit GLM, GAM, and SCAM models:
`fitSNRdetectionFunc <- `
```{r, code=fitSNRdetectionFunc, eval=FALSE}
```

```{r GLM p.det, warning=FALSE, message=FALSE}

snrDetFun.glm <- callDensity::fitSNRdetectionFunc(SNRinfo,
                                              modelType = "glm")

results.glm <- cde(Nc=Nc, capHistTab = capHistTab, snrDetFun = snrDetFun.glm,
                   SL = SL, TL = TL, NL=NLsamp, A = A, modelType = 'glm')


```

## Results

Table showing how the the results of the estimated call density compare to the true values. 

```{r Call.Density.Results, echo=TRUE, warning=FALSE}

results.true = data.frame(season='year',    siteCode='',  Nc=n, c=det2params$c,
                          k=1, T=Time, A=A, pa=mean(simDet2$p_det,na.rm=TRUE),
                          SLmean=SL$mean, SLsd=SL$sd, NLmean=NL$mean, NLsd=NL$sd,
                          modelType='true', CV.Nc=0, CV.c=0, CV.pa=0, 
                          Dc=TrueCallDensity,  CV.Dc=0)
# results.true = data.frame(Dc = TrueCallDensity, SampleSize = n, A = A,# Environment 
#                      NLmean = NL$mean, NLsd = NL$sd,            # Noise
#                      model=det2params$func, c = det2params$c, # Detector params
#                      Pa = mean(simDet2$p_det,na.rm=TRUE), 
#                      row.names='Truth')

results<-rbind( results.true, results.glm)
res <- subset(results,select=c('Dc','Nc','A','NLmean','NLsd','modelType','c',
                               'pa') )
kableExtra::kbl(res, digits = c(4,3,0,1, 2, NA, 3, 4), 
                col.names= c('Dc','(sub)Sample Size', 'A','NL_mean','NL_sd',
                             'Model','c','P_a')) %>% 
  kableExtra::kable_classic(full_width=FALSE)

```

In this scenario, our estimate of Dc is far from the true value.

It appears that the estimates of mean noise level are all lower than the actual value. Using the lower mean for the NL distribution yields a higher value for P_a, which creates a bias in the estimate of Dc.

This suggests that we should consider estimating the noise level distribution independently from the positive detections/annotations.

Additionally, the estimates of false discovery rate also appear biased. This arises from true positive detections from detector2 being incorrectly classified as false positive detections because they were missed by detector1.

# Revisit GLM model using improved estimate of NL

Addressing the bias in our NL should improve our estimates of call density. This bias arises because our NL distribution came from the NL around our detections. However, detections with low NL will be over-represented because they are more likely to have high SNR, thus are more likely to be detected. 

To account for the over-representation of detections with lower NL in our sample, we can use our fitted detection function (the GLM from above). Specifically, we can add to our estimated mean NL, the SNR at which the probability of detection is 50%. [NOTE: I can't fully explain how I arrived at this, other than it intuitively made sense to me to 'add back in the difference between detected NL and average SNR,' and it seems to work reasonably well. There is possibly some well-known principle or statistical relationship here that might be recognizable to those more knowledgeable about such things. If so, would be good to update this with that better more mathematical explanation.]

```{r GLM p.det with independently estimated NL, warning=FALSE, message=FALSE}


# 'Independent' estimate of NL 
NLind <- callDensity::nlFromSnrInfo(SNRinfo, snrDetFun.glm)

results.NLind <- cde(Nc=Nc, capHistTab = capHistTab, snrDetFun = snrDetFun.glm,
                   SL = SL, TL = TL, NL=NLind, A = A, modelType = 'glm')

```

CallDensity package code for correcting NL distribution by adding back in the SNR for 50% probability of detection:
`nlFromSnrInfo <- `
```{r, code=nlFromSnrInfo, eval=FALSE}
```


```{r}

results<-rbind( results.true, results.glm, results.NLind)

res <- subset(results,select=c('Dc','Nc','A','NLmean','NLsd','modelType','c',
                               'pa') )
kableExtra::kbl(res, digits = c(4,3,0,1, 2, NA, 3, 4), 
                col.names= c('Dc','(sub)Sample Size', 'A','NL_mean','NL_sd',
                             'Model','c','P_a')) %>% 
  kableExtra::kable_classic(full_width=FALSE)

```

Using the correct mean value for the NL distribution yeilds estimates of Dc and P_a that are closer to the true value.

However, the estimate of the false discovery rate is still incorrect leading to a bias in Dc and P_a.

## Adjudicated capture recapture/VGLM fit to SNR detection function

Adjudicated capture-recapture models can make use of all information from multiple detectors to better estimate false discovery rate and probability of detection [@miller2022].

Here we simulate perfect adjudication by removing all false positives from both detectors and retaining all true positives from both detectors in our capture history table.

We fit a capture-recapture model to these adjudicated data using the VGAM package with the function 'vglm' [@yee2015].

```{r VGLM p.det, warning=FALSE, message=FALSE}
library(VGAM)

#VGLMs work a bit differently. We still pass in a capture history table. But now
#the ground-truth column is in a column separate from detect_table1 (here
#groundTruth2). we want to include in our CH table all predicted positive
#detections from either detector, not just the ground-truth. We fit the vglm
#detection function outside of the call to cde, and pass it in as a parameter.
#For the vglm fit, we want only adjudicated positive detections (no false
#positives). Then we rename all the   columns from detector1 (i.e. that end in
#1) to something that won't conflict. Finally, we rename our ground-truth column
#to detect_table1, since that is what the cde function expects to be
#ground-truth. 
#
#TODO: Change the column name in the capture history table to groundtruth or
#make this an explicit parameter.

ch <- subset(capHistTab,
                      (capHistTab$detect_table1 | capHistTab$detect_table2))
# Rename all detector 1 to something that won't conflict e.g. i
names(ch) <- gsub('1','i',names(capHistTab))

# Rename adjudicated to detect_table1
names(ch)[names(ch)=='groundTruth2']<-'detect_table1'


### Adjudicated positive for VGLM capture recapture model
adjudicated <- subset(capHistTab,capHistTab$groundTruth1 &
                (capHistTab$detect_table1 | capHistTab$detect_table2))

summary(adjudicated$group1)
summary(adjudicated$group2)

# Number of adjudicated positive detections
n_adj <- dim(adjudicated)[1]

# Subset used for calculating false discovery rate of detector 2. Actually, we
# use the fact that the false discovery rate is equal to (1-precision). 
pp2 = sum(capHistTab$detect_table2==TRUE,na.rm=TRUE) # Predicted positive det2
fp2 = sum(capHistTab$detect_table2==TRUE & capHistTab$groundTruth2==FALSE,
          na.rm=TRUE) # false positives from det2

adj_c <- fp2/pp2


observerNames = c("detect_table1", "detect_table2")
snrDetFun.vglm <- fitSNRvglm(adjudicated, observerNames, 
                             whichObserver = "detect_table2")
# Best estimate of NL 
# NLadj <- nlFromSnrInfo(adjudicated, snrDetFun.vglm)

results.vglm <- cde(Nc=Nc, capHistTab = ch, snrDetFun = snrDetFun.vglm,
                   SL = SL, TL = TL, A = A, modelType = 'vglm')

```

```{r}

results<-rbind( results.true, results.glm, results.NLind, results.vglm)

res <- subset(results,select=c('Dc','Nc','A','NLmean','NLsd','modelType','c',
                               'pa') )
kableExtra::kbl(res, digits = c(4,3,0,1, 2, NA, 3, 4), 
                col.names= c('Dc','(sub)Sample Size', 'A','NL_mean','NL_sd',
                             'Model','c','P_a')) %>% 
  kableExtra::kable_classic(full_width=FALSE)

```

```{r eval=TRUE, include=TRUE}

# short version of results for paper
res <- subset(results, select =c(NLmean, c, pa, Dc))
res$error <- (res$Dc[1]-res$Dc)/res$Dc[1]*100
rownames(res)<- c('Truth','OG incorrect NL','OG corrected NL','CR')

kableExtra::kbl(res, digits = c(1,3,4, 4, 1), 
                col.names= c('$N_c$', '$\\overline{\\small{NL}}$',
                             '$\\hat{c}$','$\\hat{p}_a$', '$D_c$','% Error')) %>% 
  kableExtra::kable_classic(full_width=FALSE)


```

The estimate of Dc derived from the vglm model is very close to the true value for this simulation.

Adjudicated capture-recapture along with independently estimated NL yields more accurate estimates of c and P_a.

## Inspect detection functions for insights

```{r compare snr detection functions, warning=FALSE}
snrs <- data.frame(SNR=seq(-20,20,0.5))

cols = colorRampPalette(c('#EDB120','#D95319','#0072BD','#7E2F8E','#77AC30',
                        '#4DBEEE','#A2142F'))
preds.glm= predict(snrDetFun.glm, type='response',newdata=snrs)
preds.vglm=predict(snrDetFun.vglm,type='response',newdata=snrs)
preds.vglm0=predict(snrDetFun.vglm,type='response',newdata=snrs,
                    type.fitted='onempall0')# One minus probability of all zeros
pr1 =apply(cbind(preds.vglm[,1],preds.vglm0),1,prod)
pr2 =apply(cbind(preds.vglm[,2],preds.vglm0),1,prod)



par(cex=0.8,mar=c(3.1,3.1,0.25,0),mgp=c(2.1,1,0))
plot(snrs$SNR, xlim= c(-20,20),  ylim= c(0,1),
     xlab="SNR", ylab = "P(Detection)", type='n')
grid(nx = NULL, ny = NULL, lty = 2, col = "gray", lwd = 2)
lines(snrs$SNR,preds.glm,col=cols(7)[4],lwd=2)
# lines(snrs$SNR,preds.vglm0,col=cols(7)[2],lwd=2) # 1-
lines(snrs$SNR,pr2,col=cols(7)[1],lwd=2)    # vglm 1-p(all zero)
points(simDet2$snr,simDet2$p_det,pch='.')     # Actual probability of detection
lines(snrs$SNR,pr1,col=cols(7)[3],lwd=2)    # vglm 1-p(all zero)
points(capHistTab$snr1,capHistTab$p_det1,pch='.')
# lines(snrs$SNR,preds.vglm[,1],col=cols(7)[5],lwd=2) # Response vglm detector1
# lines(snrs$SNR,preds.vglm[,2],col=cols(7)[6],lwd=2) # Response vglm detector2

legend( "bottomright", 
        legend =c("Detection function TRUE",
                  "Detection function GLM",
                  "Detection function VGLM2",
                  "Response VGLM 1",
                  "Response VGLM 2",
                  "Detection function vglm1"
                  ),
        col=c('black', cols(7)[4], cols(7)[1], cols(7)[5], cols(7)[6],cols(7)[3]),
              #, cols(7)[2]), 
        lty=rep('solid',6),
        lwd=rep(2,6)
)
```

# References
