---
title: "callDensity: Where does the noise level correction come from?"
output: 
  html_document: default
  pdf_document: default
  rmarkdown::html_vignette:
    vignette: >
      %\VignetteEncoding{UTF-8}
      %\VignetteIndexEntry{callDensity: Where does the noise level correction come from?}
      %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.height = 3, 
  comment = "#>"
)
```

```{r setup}
library(callDensity)
library(ggplot2)
library(kableExtra)
```

# The problem

The second vignette found that noise levels measured at detections are biased
low. Detections with low noise are over-represented, because low noise means
high SNR, and high SNR means the call is more likely to be detected.

That vignette also proposed a correction. Add the SNR at which the probability
of detection is 50% back onto the estimated mean noise level. The correction
worked. It took the estimate from 81.2 dB to 83.8 dB against a true value of
84.0 dB.

The vignette was honest that the correction had no derivation behind it. This
vignette asks whether it generalises.

The answer is that it does not. The bias is real and the direction of the
correction is right. The quantity used to size the correction is wrong. It
happens to be close to the right size in that one simulation.

# What sets the size of the bias

Start with the geometry. Calls are uniformly distributed in two dimensions, so
the number of calls within range $r$ is proportional to $r^2$. With spherical
spreading the received level is $RL = SL - 20\log_{10}r$, so a call has
$RL > x$ when it lies inside $r = 10^{(SL-x)/20}$. The area inside that radius
is proportional to $10^{(SL-x)/10}$. Therefore

\begin{equation}
P(RL > x) \propto 10^{-x/10} = e^{-\lambda x}, \qquad \lambda = \frac{\ln 10}{10} \approx 0.23 \text{ per dB}
\end{equation}

Received level is exponentially distributed in dB. The reason is worth stating
plainly. Raise the noise floor by 10 dB and you need 10 dB more received level,
so your detection radius shrinks by a factor of 3.16, so your detection area
shrinks by a factor of 10. Ten times fewer detections for every ten decibels.
That factor is the whole mechanism.

Now write the probability that a call is detected, given that it happened to
occur at a particular noise level $nl$. Let $p(\cdot)$ be the detection function.

\begin{equation}
P(\text{det} \mid nl) = \int f_{RL}(x)\, p(x - nl)\, dx
\end{equation}

Substitute $u = x - nl$ and separate the exponential.

\begin{equation}
P(\text{det} \mid nl) = e^{-\lambda \, nl} \cdot \lambda \int e^{-\lambda u} p(u)\, du = C e^{-\lambda \, nl}
\end{equation}

The integral is a constant. It contains no $nl$. This is the key step. The shape
and threshold of the detector set $C$, which is the overall detection rate. They
do not touch $\lambda$. The rate at which detections fall away with rising noise
is fixed by the spreading law and by how the calls are spread out in space.
Nothing else.

Sampling noise only at detections therefore weights the true distribution by
$e^{-\lambda\,nl}$. For a normal noise distribution this is exponential tilting,
also called the Esscher transform. A normal distribution tilted by an
exponential is still normal.

\begin{equation}
NL \mid \text{detected} \sim \mathcal{N}(\mu - \lambda\sigma^2, \; \sigma^2)
\end{equation}

The mean shifts down by $\lambda\sigma^2$. The variance does not change. That
last part is checkable against the second vignette, where the sampled standard
deviation was 3.88 against a true value of 4.00.

So the bias is $\lambda\sigma^2$. It is a property of the propagation geometry
and the noise variance. The SNR at 50% detection is a property of the detector.
These are unrelated quantities. In the second vignette they were both close to
2.6 dB, which is why the correction appeared to work.

The rest of this vignette tests that claim by simulation.

# Simulation setup

The setup is the same as the second vignette. One million calls uniformly
distributed within 1000 km of the hydrophone, blue whale source levels, a
noise distribution, and spherical spreading.

The scale is set to match the real dataset behind Miller et al. (2026). That
study annotated 200 h of a year of recording at Casey2019 and found between
2246 and 9054 detections per observer. Subsampling one hour in every 41 gives
214 h here, and at the detector settings of the second vignette that yields
about 2300 detections. Adjudicating every 8th detection, as that study did,
takes it to roughly 280. Those are the numbers the correction has to work
with.

One change from the second vignette is deliberate. False positives are
adjudicated away before the noise levels are estimated. The second vignette
does this for its capture-recapture models and not for its observer ground
truth models, so both paths already exist in that vignette. Adjudication is
used here because it is the recommended method of Miller et al. (2026), and
because false positives carry noise levels that were never conditioned on
detection, which would dilute the very thing being measured.

The detection function is fitted with a GLM on the adjudicated detections. The
real method uses a capture-recapture model. That difference does not matter
here, because this vignette is about the noise level estimate rather than about
the capture-recapture machinery, and `nlFromSnrInfo` only ever uses the
detection function to look up one number.

```{r simulation.setup, warning=FALSE, message=FALSE}

n <- 1e6         # number of simulated calls
R <- 1e6         # radius in m
minDate <- as.POSIXct("2025-01-01")
maxDate <- as.POSIXct("2026-01-01")

set.seed(1)
simLoc <- simCallLocation(n = n, R = R, minDate = minDate, maxDate = maxDate)

SL     <- data.frame(mean = 190, sd = 4, sampleSize = 350)
NLtrue <- data.frame(mean = 84,  sd = 4, sampleSize = n)
tlFunc <- function(r) 20 * log10(r)

# WORKAROUND, not style. simulateDetector() draws noise levels for its false
# positives using a variable named NL that it reads out of the global
# environment. NL is not one of its arguments. This assignment exists only to
# satisfy that lookup. Sweep 2 below has to keep it in step with the noise
# distribution being tested.
NL <- NLtrue

sim <- simCallAcoustics(simLoc, SL, NL, TL = tlFunc)

```

Detector parameters follow the second vignette. Detector 1 is the better
detector and supplies the ground truth. Detector 2 is the automated detector
whose performance we characterise.

```{r detector.setup}

# False discovery rates as in the second vignette. These span the range found
# at Casey2019, where f ran from 0.02 to 0.43 across observers and detectors.
det1params <- data.frame(location = 1, scale = 2, func = 'plogis',
                         c = 0.1, fpMean = 0, fpSD = 2)

det2params <- data.frame(location = 2, scale = 4, func = 'plogis',
                         c = 0.3, fpMean = 0, fpSD = 2)

subsampleInterval <- "41 hour"   # 214 h, matching the 200 h annotated at Casey2019
adjudicateEvery   <- 8           # every 8th detection judged, as at Casey2019

```

## A helper for one sweep point

This function runs the existing pipeline once and returns a single row. It
introduces no new method. It calls `simulateDetector`, `subsampleSimInTime`,
`simsTocaptureHistoryTable`, `capHistTosnrInfo`, `fitSNRdetectionFunc` and
`nlFromSnrInfo` in the same order as the second vignette.

Adjudication is simulated by keeping only the detections that the simulation
knows to be true positives, then keeping every 8th of those. In a real dataset
a judge does this work, and the assumption is that the judged subset is
representative of the period.

The SNR offset applied by `nlFromSnrInfo` is recovered by difference rather
than recomputed, so that this vignette tests the shipped function rather than a
copy of it.

```{r sweep.helper}

nlBiasTrial <- function(sim, det1params, det2params,
                        subsampleInterval = "41 hour",
                        adjudicateEvery   = 8) {

  simDet1 <- simulateDetector(det1params, sim)
  simDet2 <- simulateDetector(det2params, sim)

  sub1 <- subsampleSimInTime(simDet1, interval = subsampleInterval)
  sub2 <- subsampleSimInTime(simDet2, interval = subsampleInterval)

  cht     <- simsTocaptureHistoryTable(sub1, sub2)

  # Simulated adjudication. Keep true positives, then judge every nth one.
  # groundTruth1 is 0 for false positives, so this is a perfect judge.
  adj <- subset(cht, as.logical(cht$groundTruth1))
  adj <- adj[seq(1, nrow(cht), by = adjudicateEvery), ]

  SNRinfo <- capHistTosnrInfo(adj)
  
  detFun <- fitSNRdetectionFunc(SNRinfo, modelType = "glm")

  # Uncorrected noise level estimate. This duplicates the first three lines of
  # nlFromSnrInfo. See the note at the end of this vignette.
  NLsampMean <- mean(SNRinfo$NoiseRL, na.rm = TRUE)
  NLsampSd   <- sd(SNRinfo$NoiseRL, na.rm = TRUE)

  NLind <- nlFromSnrInfo(SNRinfo, detFun)

  data.frame(
    nDetected    = nrow(cht),
    nAdjudicated = nrow(adj),
    NLsampMean   = NLsampMean,
    NLsampSd     = NLsampSd,
    NLindMean    = NLind$mean,
    snrOffset    = NLind$mean - NLsampMean
  )
}

```

# Experiment 1: move the detector

If the correction is sound, the size of the bias should track the SNR at 50%
detection. If the derivation above is right, the bias should barely move at all
while the correction swings across the full range.

Both detectors are shifted together, keeping the 1 dB gap between them, so that
the detection function fitted to detector 2 shifts with the detector that
supplies the ground truth.

```{r sweep.detector, warning=FALSE, message=FALSE}

snr50 <- seq(from = -5, to = 10, by = 3)

sweepA <- do.call(rbind, lapply(snr50, function(s) {
  d1 <- det1params; d1$location <- s
  d2 <- det2params; d2$location <- s + 1
  cbind(data.frame(det1Location = s),
        nlBiasTrial(sim, d1, d2, subsampleInterval, adjudicateEvery))
}))

sweepA$biasNeeded <- NL$mean - sweepA$NLsampMean
sweepA$residual   <- sweepA$biasNeeded - sweepA$snrOffset

kbl(subset(sweepA, select = c(det1Location, nDetected, nAdjudicated,
                              NLsampMean, biasNeeded, snrOffset, residual)),
    digits = c(1, 0, 0, 2, 3, 3, 3),
    col.names = c('Detector 1 location', 'Detections', 'Adjudicated',
                  'Sampled NL mean', 'Bias needed', 'Correction applied',
                  'Residual')) %>%
  kable_classic(full_width = FALSE)

```

The detection counts are worth pausing on. At Casey2019 the five observers and
detectors found between 2246 and 9054 detections in 200 h, of which 681 were
adjudicated true positives. The counts above sit in the same range. The
sampled noise mean is an average of a quantity with a standard deviation near
4 dB, so even the sparsest row here estimates it to better than a quarter of a
decibel. Sample size is not what limits this correction.

```{r sweep.detector.plot, warning=FALSE, fig.height=3}

plotA <- data.frame(
  snr50 = rep(sweepA$det1Location, 2),
  dB    = c(sweepA$biasNeeded, sweepA$snrOffset),
  what  = rep(c("Bias needed", "Correction applied"), each = nrow(sweepA))
)

# Detections at Casey2019 were modal at 1-2 dB SNR and spanned roughly -5 to
# +15 dB, so a detector with a location much above about 5 dB would find almost
# nothing. Shade the range a real detector would plausibly occupy.
ggplot(plotA, aes(x = snr50, y = dB, colour = what)) +
  annotate("rect", xmin = -5, xmax = 5, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = 0.5) +
  annotate("text", x = 0, y = 9, label = "plausible detectors",
           colour = "grey30", size = 3) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  xlab("Detector 1 location (dB SNR)") +
  ylab("dB") +
  labs(colour = NULL) +
  theme(legend.position = "top")

```

The bias needed stays flat. The correction applied is a straight line through
the origin, because it is the detector threshold and nothing else. The two
curves cross once. The second vignette sits at that crossing.

The right hand end of this sweep is a stress test rather than a survey of real
detectors. Nobody would deploy a detector that only finds calls above 10 dB
SNR. The shaded band marks where a real detector would sit, and the finding
does not depend on the unshaded part. Within the plausible band alone, the
residual runs from about +7 dB to about -2 dB. That is nine decibels of error
from nothing more than the choice of detector, and the detection functions in
figure 6 of Miller et al. (2026) span roughly that much across the observers
and algorithms of a single real dataset.

# Experiment 2: change the noise variance

The derivation says the bias is $\lambda\sigma^2$, so doubling the standard
deviation of the noise should quadruple the bias. The correction applied knows
nothing about the noise distribution, so it should not move at all.

Unlike the detector sweep, this one stays inside reality throughout. A noise
standard deviation of 4 dB follows Castro et al. (2024). Southern Ocean band
levels swing considerably more than that over a year as wind and sea ice
change, so 8 dB is not an exotic choice.

The acoustics are regenerated for each noise distribution. The call locations
are not, so the geometry is held fixed.

```{r sweep.noise, warning=FALSE, message=FALSE}

nlSd <- c(1, 2, 4, 6, 8, 10)

# A for loop rather than lapply. simulateDetector reads NL from the global
# environment, and a knitr chunk evaluates in the global environment, so
# assigning NL here reaches it. An lapply function body would not.
sweepB <- NULL
for (s in nlSd) {

  NL   <- data.frame(mean = NLtrue$mean, sd = s, sampleSize = n)
  simS <- simCallAcoustics(simLoc, SL, NL, TL = tlFunc)

  sweepB <- rbind(sweepB,
                  cbind(data.frame(nlSd = s),
                        nlBiasTrial(simS, det1params, det2params,
                                    subsampleInterval, adjudicateEvery)))
}

NL <- NLtrue   # restore the global before anything else reads it

sweepB$biasNeeded <- NLtrue$mean - sweepB$NLsampMean

# Analytic prediction from exponential tilting. lambda follows from the
# spreading law and the two-dimensional uniform call distribution.
lambda <- log(10) / 10
sweepB$tiltPredicted <- lambda * sweepB$NLsampSd^2

kbl(subset(sweepB, select = c(nlSd, nAdjudicated, NLsampSd, biasNeeded,
                              snrOffset, tiltPredicted)),
    digits = c(1, 0, 2, 3, 3, 3),
    col.names = c('True sd(NL)', 'Adjudicated', 'Sampled sd(NL)',
                  'Bias needed', 'Correction applied',
                  'Tilt prediction')) %>%
  kable_classic(full_width = FALSE)

```

```{r sweep.noise.plot, warning=FALSE, fig.height=3}

plotB <- data.frame(
  sd   = rep(sweepB$nlSd, 3),
  dB   = c(sweepB$biasNeeded, sweepB$snrOffset, sweepB$tiltPredicted),
  what = rep(c("Bias needed", "Correction applied", "Tilt prediction"),
             each = nrow(sweepB))
)

ggplot(plotB, aes(x = sd, y = dB, colour = what)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  xlab("True sd(NL) (dB)") +
  ylab("dB") +
  labs(colour = NULL) +
  theme(legend.position = "top")

```

Three things to read off this.

The bias grows with the square of the noise standard deviation, as the tilting
result requires. The correction applied is flat, because the detector never
changed.

The sampled standard deviation tracks the true standard deviation closely. This
is the tilting result again. Tilting a normal distribution by an exponential
shifts the mean and leaves the variance alone. It means that in practice
$\sigma$ can be estimated from the detections even though $\mu$ cannot.

The tilt prediction is right at small noise variance and overshoots at large
noise variance. That is not a failure of the derivation. It is the derivation
running out of road, and it is the subject of the next section.

# Where the theory stops

The derivation assumed that the exponential in $P(RL > x)$ runs forever. It does
not. Beyond the edge of the study area there is no more ocean to draw calls
from, so lowering the noise floor stops buying extra detection area. The tilt
weight flattens at low noise, the downward pull on the mean weakens, and the
true bias falls below $\lambda\sigma^2$.

This matters. At the parameters of the second vignette the tilt prediction is
about 3.7 dB against a bias of about 3.1 dB. It overcorrects. At a noise
standard deviation of 10 dB it is wrong by roughly a factor of two.

So $\lambda\sigma^2$ explains the bias. It should not be shipped as the
correction.

The fix is to stop assuming the exponential and compute $P(\text{det} \mid nl)$
directly:

\begin{equation}
P(\text{det} \mid nl) = \frac{\sum_\theta \sum_r w(r) \, \mathrm{E}_{SL}\left[p\left(SL - TL(r,\theta) - nl\right)\right]}{\sum_\theta \sum_r w(r)}
\end{equation}

where $w(r) \propto r$ is the area weight, set to zero beyond the truncation
distance. This is the same weighted sum that `pDetInArea` already forms, taken
as a function of noise level rather than evaluated at a drawn noise level. It
needs the transmission loss lookup table, the source level distribution, the
detection function, and the truncation distances. Every one of those is already
an argument to `cde`.

With $P(\text{det} \mid nl)$ in hand the sampled mean follows directly:

\begin{equation}
\mathrm{E}[NL \mid \text{det}] = \frac{\int nl \, f_N(nl) \, P(\text{det} \mid nl) \, dnl}{\int f_N(nl) \, P(\text{det} \mid nl) \, dnl}
\end{equation}

and the true mean is recovered by solving for the $\mu$ that reproduces the
observed `mean(NoiseRL)`. That is a one dimensional root find.

This version never mentions $20\log_{10}r$. It works with any transmission loss
model, including range dependent output from a parabolic equation model over
variable bathymetry and sound speed. Building and testing it is the next step.

# Notes

`nlBiasTrial` is scaffolding for this vignette. It introduces no method and it
is not proposed for the package.

It does compute `mean(NoiseRL)` and `sd(NoiseRL)` inline, which duplicates the
first three lines of `nlFromSnrInfo`. That duplication is a hint that
`nlFromSnrInfo` is doing two separable jobs, estimating the sampled noise
distribution and then correcting it. Splitting them would let this vignette
call the shipped estimator for both.

## A bug found while writing this

`simulateDetector` draws noise levels for its false positives from a variable
named `NL` that it reads out of the global environment. `NL` is not one of its
arguments and it cannot be derived from the other arguments. The second
vignette happens to define a top level object with that name, so the lookup
lands on the right thing and the function appears to work.

It should take `NL` as an argument. Until it does, any script that calls
`simulateDetector` without a suitably named object in scope will either fail or,
worse, silently draw false positive noise levels from an unrelated
distribution.

Two assumptions in the derivation deserve stating, because both are limitations
of the single sensor method rather than of this vignette.

Calls are assumed uniformly distributed in two dimensions. The exponent
$\lambda$ comes directly from that assumption together with the spreading law.
A clustered or depth structured call distribution changes it.

Noise is assumed independent of call presence and of range. Whether that holds
for a real dataset is a separate question, and it is the reason the second
vignette's suggestion to measure noise independently of the detections is not
as simple as it looks.

# References
