---
title: "callDensity: Where does the noise level correction come from?"
output: 
  html_document: default
  pdf_document: default
  rmarkdown::html_vignette:
    vignette: >
      %\VignetteEncoding{UTF-8}
      %\VignetteIndexEntry{callDensity: Where does the noise level correction come from?}
      %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.height = 3, 
  comment = "#>"
)
```

```{r setup}
library(callDensity)
library(ggplot2)
library(kableExtra)
```

# The problem

The second vignette found that noise levels measured at detections are biased
low. Detections with low noise are over-represented, because low noise means
high SNR, and high SNR means the call is more likely to be detected.

That vignette also proposed a correction. Add the SNR at which the probability
of detection is 50% back onto the estimated mean noise level. The correction
worked. It took the estimate from 81.2 dB to 83.8 dB against a true value of
84.0 dB.

The vignette was honest that the correction had no derivation behind it. This
vignette asks whether it generalises.

The answer is that it does not. The bias is real and the direction of the
correction is right. The quantity used to size the correction is wrong. It
happens to be close to the right size in that one simulation.

# What sets the size of the bias

Start with the geometry. Calls are uniformly distributed in two dimensions, so
the number of calls within range $r$ is proportional to $r^2$. With spherical
spreading the received level is $RL = SL - 20\log_{10}r$, so a call has
$RL > x$ when it lies inside $r = 10^{(SL-x)/20}$. The area inside that radius
is proportional to $10^{(SL-x)/10}$. Therefore

\begin{equation}
P(RL > x) \propto 10^{-x/10} = e^{-\lambda x}, \qquad \lambda = \frac{\ln 10}{10} \approx 0.23 \text{ per dB}
\end{equation}

Received level is exponentially distributed in dB. The reason is worth stating
plainly. Raise the noise floor by 10 dB and you need 10 dB more received level,
so your detection radius shrinks by a factor of 3.16, so your detection area
shrinks by a factor of 10. Ten times fewer detections for every ten decibels.
That factor is the whole mechanism.

Now write the probability that a call is detected, given that it happened to
occur at a particular noise level $nl$. Let $p(\cdot)$ be the detection function.

\begin{equation}
P(\text{det} \mid nl) = \int f_{RL}(x)\, p(x - nl)\, dx
\end{equation}

Substitute $u = x - nl$ and separate the exponential.

\begin{equation}
P(\text{det} \mid nl) = e^{-\lambda \, nl} \cdot \lambda \int e^{-\lambda u} p(u)\, du = C e^{-\lambda \, nl}
\end{equation}

The integral is a constant. It contains no $nl$. This is the key step. The shape
and threshold of the detector set $C$, which is the overall detection rate. They
do not touch $\lambda$. The rate at which detections fall away with rising noise
is fixed by the spreading law and by how the calls are spread out in space.
Nothing else.

Sampling noise only at detections therefore weights the true distribution by
$e^{-\lambda\,nl}$. For a normal noise distribution this is exponential tilting,
also called the Esscher transform. A normal distribution tilted by an
exponential is still normal.

\begin{equation}
NL \mid \text{detected} \sim \mathcal{N}(\mu - \lambda\sigma^2, \; \sigma^2)
\end{equation}

The mean shifts down by $\lambda\sigma^2$. The variance does not change. That
last part is checkable against the second vignette, where the sampled standard
deviation was 3.88 against a true value of 4.00.

So the bias is $\lambda\sigma^2$. It is a property of the propagation geometry
and the noise variance. The SNR at 50% detection is a property of the detector.
These are unrelated quantities. In the second vignette they were both close to
2.6 dB, which is why the correction appeared to work.

The rest of this vignette tests that claim by simulation.

# Simulation setup

The setup is the same as the second vignette. One million calls uniformly
distributed within 1000 km of the hydrophone, blue whale source levels, a
noise distribution, and spherical spreading.

Two changes are deliberate.

False positives are switched off. False positives carry noise levels that are
not conditioned on detection, so they dilute the bias. The dilution gets
stronger as the detection rate falls, which would put a spurious trend into the
first sweep. Switching them off isolates the mechanism being tested.

The subsample interval is shortened from 41 hours to 4 hours. The bias is a
property of the detection process, not of the subsample. Subsample size only
controls precision. The sweeps below reach detection rates near 1%, where 41
hours would leave too few detections to fit a detection function.

```{r simulation.setup, warning=FALSE, message=FALSE}

n <- 1e6         # number of simulated calls
R <- 1e6         # radius in m
minDate <- as.POSIXct("2025-01-01")
maxDate <- as.POSIXct("2026-01-01")

set.seed(1)
simLoc <- simCallLocation(n = n, R = R, minDate = minDate, maxDate = maxDate)

SL     <- data.frame(mean = 190, sd = 4, sampleSize = 350)
NLtrue <- data.frame(mean = 84,  sd = 4, sampleSize = n)
tlFunc <- function(r) 20 * log10(r)

sim <- simCallAcoustics(simLoc, SL, NLtrue, TL = tlFunc)

```

Detector parameters follow the second vignette. Detector 1 is the better
detector and supplies the ground truth. Detector 2 is the automated detector
whose performance we characterise.

```{r detector.setup}

fpRate <- 0.1   # false positives switched off to isolate the mechanism

det1params <- data.frame(location = 1, scale = 2, func = 'plogis',
                         c = fpRate, fpMean = 0, fpSD = 2)

det2params <- data.frame(location = 2, scale = 4, func = 'plogis',
                         c = fpRate, fpMean = 0, fpSD = 2)

subsampleInterval <- "4 hour"

```

## A helper for one sweep point

This function runs the existing pipeline once and returns a single row. It
introduces no new method. It calls `simulateDetector`, `subsampleSimInTime`,
`simsTocaptureHistoryTable`, `capHistTosnrInfo`, `fitSNRdetectionFunc` and
`nlFromSnrInfo` in the same order as the second vignette.

The SNR offset applied by `nlFromSnrInfo` is recovered by difference rather
than recomputed, so that this vignette tests the shipped function rather than a
copy of it.

```{r sweep.helper}

nlBiasTrial <- function(sim, det1params, det2params,
                        subsampleInterval = "4 hour") {

  simDet1 <- simulateDetector(det1params, sim)
  simDet2 <- simulateDetector(det2params, sim)

  sub1 <- subsampleSimInTime(simDet1, interval = subsampleInterval)
  sub2 <- subsampleSimInTime(simDet2, interval = subsampleInterval)

  cht     <- simsTocaptureHistoryTable(sub1, sub2)
  SNRinfo <- capHistTosnrInfo(cht)

  detFun <- fitSNRdetectionFunc(SNRinfo, modelType = "glm")

  # Uncorrected noise level estimate. This duplicates the first three lines of
  # nlFromSnrInfo. See the note at the end of this vignette.
  NLsampMean <- mean(SNRinfo$NoiseRL, na.rm = TRUE)
  NLsampSd   <- sd(SNRinfo$NoiseRL, na.rm = TRUE)

  NLind <- nlFromSnrInfo(SNRinfo, detFun)

  data.frame(
    nSnrInfo   = nrow(SNRinfo),
    detRate    = mean(simDet2$detect_table, na.rm = TRUE),
    NLsampMean = NLsampMean,
    NLsampSd   = NLsampSd,
    NLindMean  = NLind$mean,
    snrOffset  = NLind$mean - NLsampMean
  )
}

```

# Experiment 1: move the detector

If the correction is sound, the size of the bias should track the SNR at 50%
detection. If the derivation above is right, the bias should barely move at all
while the correction swings across the full range.

Both detectors are shifted together, keeping the 1 dB gap between them, so that
the detection function fitted to detector 2 shifts with the detector that
supplies the ground truth.

```{r sweep.detector, warning=FALSE, message=FALSE}

snr50 <- seq(from = -5, to = 10, by = 3)

sweepA <- do.call(rbind, lapply(snr50, function(s) {
  d1 <- det1params; d1$location <- s
  d2 <- det2params; d2$location <- s + 1
  cbind(data.frame(det1Location = s), nlBiasTrial(sim, d1, d2, subsampleInterval))
}))

sweepA$biasNeeded <- NLtrue$mean - sweepA$NLsampMean
sweepA$residual   <- sweepA$biasNeeded - sweepA$snrOffset

kbl(subset(sweepA, select = c(det1Location, detRate, NLsampMean,
                              biasNeeded, snrOffset, residual)),
    digits = c(1, 4, 2, 3, 3, 3),
    col.names = c('Detector 1 location', 'Detection rate', 'Sampled NL mean',
                  'Bias needed', 'Correction applied', 'Residual')) %>%
  kable_classic(full_width = FALSE)

```

```{r sweep.detector.plot, warning=FALSE, fig.height=3}

plotA <- data.frame(
  snr50 = rep(sweepA$det1Location, 2),
  dB    = c(sweepA$biasNeeded, sweepA$snrOffset),
  what  = rep(c("Bias needed", "Correction applied"), each = nrow(sweepA))
)

ggplot(plotA, aes(x = snr50, y = dB, colour = what)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  xlab("Detector 1 location (dB SNR)") +
  ylab("dB") +
  labs(colour = NULL) +
  theme(legend.position = "top")

```

The bias needed stays flat. The correction applied is a straight line through
the origin, because it is the detector threshold and nothing else. The two
curves cross once. The second vignette sits at that crossing.

# Experiment 2: change the noise variance

The derivation says the bias is $\lambda\sigma^2$, so doubling the standard
deviation of the noise should quadruple the bias. The correction applied knows
nothing about the noise distribution, so it should not move at all.

The acoustics are regenerated for each noise distribution. The call locations
are not, so the geometry is held fixed.

```{r sweep.noise, warning=FALSE, message=FALSE}

nlSd <- c(1, 2, 4, 6, 8, 10)

sweepB <- do.call(rbind, lapply(nlSd, function(s) {
  NLs   <- data.frame(mean = 84, sd = s, sampleSize = n)
  simS  <- simCallAcoustics(simLoc, SL, NLs, TL = tlFunc)
  cbind(data.frame(nlSd = s), nlBiasTrial(simS, det1params, det2params,
                                          subsampleInterval))
}))

sweepB$biasNeeded <- 84 - sweepB$NLsampMean

# Analytic prediction from exponential tilting. lambda follows from the
# spreading law and the two-dimensional uniform call distribution.
lambda <- log(10) / 10
sweepB$tiltPredicted <- lambda * sweepB$NLsampSd^2

kbl(subset(sweepB, select = c(nlSd, detRate, NLsampSd, biasNeeded,
                              snrOffset, tiltPredicted)),
    digits = c(1, 4, 2, 3, 3, 3),
    col.names = c('True sd(NL)', 'Detection rate', 'Sampled sd(NL)',
                  'Bias needed', 'Correction applied',
                  'Tilt prediction')) %>%
  kable_classic(full_width = FALSE)

```

```{r sweep.noise.plot, warning=FALSE, fig.height=3}

plotB <- data.frame(
  sd   = rep(sweepB$nlSd, 3),
  dB   = c(sweepB$biasNeeded, sweepB$snrOffset, sweepB$tiltPredicted),
  what = rep(c("Bias needed", "Correction applied", "Tilt prediction"),
             each = nrow(sweepB))
)

ggplot(plotB, aes(x = sd, y = dB, colour = what)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  xlab("True sd(NL) (dB)") +
  ylab("dB") +
  labs(colour = NULL) +
  theme(legend.position = "top")

```

Three things to read off this.

The bias grows with the square of the noise standard deviation, as the tilting
result requires. The correction applied is flat, because the detector never
changed.

The sampled standard deviation tracks the true standard deviation closely. This
is the tilting result again. Tilting a normal distribution by an exponential
shifts the mean and leaves the variance alone. It means that in practice
$\sigma$ can be estimated from the detections even though $\mu$ cannot.

The tilt prediction is right at small noise variance and overshoots at large
noise variance. That is not a failure of the derivation. It is the derivation
running out of road, and it is the subject of the next section.

# Where the theory stops

The derivation assumed that the exponential in $P(RL > x)$ runs forever. It does
not. Beyond the edge of the study area there is no more ocean to draw calls
from, so lowering the noise floor stops buying extra detection area. The tilt
weight flattens at low noise, the downward pull on the mean weakens, and the
true bias falls below $\lambda\sigma^2$.

This matters. At the parameters of the second vignette the tilt prediction is
about 3.7 dB against a bias of about 3.1 dB. It overcorrects. At a noise
standard deviation of 10 dB it is wrong by roughly a factor of two.

So $\lambda\sigma^2$ explains the bias. It should not be shipped as the
correction.

The fix is to stop assuming the exponential and compute $P(\text{det} \mid nl)$
directly:

\begin{equation}
P(\text{det} \mid nl) = \frac{\sum_\theta \sum_r w(r) \, \mathrm{E}_{SL}\left[p\left(SL - TL(r,\theta) - nl\right)\right]}{\sum_\theta \sum_r w(r)}
\end{equation}

where $w(r) \propto r$ is the area weight, set to zero beyond the truncation
distance. This is the same weighted sum that `pDetInArea` already forms, taken
as a function of noise level rather than evaluated at a drawn noise level. It
needs the transmission loss lookup table, the source level distribution, the
detection function, and the truncation distances. Every one of those is already
an argument to `cde`.

With $P(\text{det} \mid nl)$ in hand the sampled mean follows directly:

\begin{equation}
\mathrm{E}[NL \mid \text{det}] = \frac{\int nl \, f_N(nl) \, P(\text{det} \mid nl) \, dnl}{\int f_N(nl) \, P(\text{det} \mid nl) \, dnl}
\end{equation}

and the true mean is recovered by solving for the $\mu$ that reproduces the
observed `mean(NoiseRL)`. That is a one dimensional root find.

This version never mentions $20\log_{10}r$. It works with any transmission loss
model, including range dependent output from a parabolic equation model over
variable bathymetry and sound speed. Building and testing it is the next step.

# Notes

`nlBiasTrial` is scaffolding for this vignette. It introduces no method and it
is not proposed for the package.

It does compute `mean(NoiseRL)` and `sd(NoiseRL)` inline, which duplicates the
first three lines of `nlFromSnrInfo`. That duplication is a hint that
`nlFromSnrInfo` is doing two separable jobs, estimating the sampled noise
distribution and then correcting it. Splitting them would let this vignette
call the shipped estimator for both.

Two assumptions in the derivation deserve stating, because both are limitations
of the single sensor method rather than of this vignette.

Calls are assumed uniformly distributed in two dimensions. The exponent
$\lambda$ comes directly from that assumption together with the spreading law.
A clustered or depth structured call distribution changes it.

Noise is assumed independent of call presence and of range. Whether that holds
for a real dataset is a separate question, and it is the reason the second
vignette's suggestion to measure noise independently of the detections is not
as simple as it looks.

# References
