---
title: "callDensity: Estimating noise levels from detections"
output: 
  html_document: default
  pdf_document: default
  rmarkdown::html_vignette:
    vignette: >
      %\VignetteEncoding{UTF-8}
      %\VignetteIndexEntry{callDensity: Estimating noise levels from detections}
      %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, fig.width = 6, fig.height = 3.5,
                      comment = "#>")
```

```{r setup}
library(callDensity)
library(ggplot2)
library(kableExtra)
```

# The problem in one paragraph

For single sensor call density estimates, we rely on a Monte Carlo simulation to
estimate the probability of detection in the area. This simulation,
`pDetInArea`, requires a distribution of noise levels as an input parameter. An
obvious place to get one is the noise measured at each detection, since those
measurements are already in the capture history table (because they're required
to estimate SNR to fit a detection function). However, those measurements are
biased low. A quiet period gives you a high SNR, and a high SNR gives you a
detection, so your detections oversample the quiet periods. Measure noise only
where you detected something and you will conclude the ocean is quieter than it
is.

This vignette shows how big that bias is, where it comes from, and how to
correct for it. The original Common Ground implementation estimated NL's mean
and sd directly from detections, with no correction and no comment on why one
might be needed -- a reasonable thing to do without a reason to suspect
otherwise, but exactly the assumption this vignette exists to check.

# From an average to a curve

`pDetInArea` already computes something close to what we need. It answers one
question: averaged over a whole distribution of noise levels, what fraction of
the calls in the study area get detected? That fraction is $\hat{p}_a$, a
single number, and `pDetInArea` gets it by drawing noise levels from a normal
distribution and averaging the detection probability over the draws.

We can change one thing about that calculation: instead of drawing noise from a
distribution, hold it fixed at one value. The
output is no longer a single averaged number, it is the detection probability
*at that noise level*. Do this at 70 dB, then 74, then 78, and so on, and the
outputs trace out a curve: probability of detection against noise level. Call
it `pDetGivenNL`.

This curve is the key to the whole solution: Detection probability at a fixed
noise level is exactly the quantity that governs the bias. A quiet spell means
more of the study area is audible, which means more calls detected; a loud spell
means less area is audible and fewer calls detected. The curve is not a new idea
bolted onto `pDetInArea`, it is $\hat{p}_a$ with the average taken back out, so
the noise dependence it was hiding is visible again. That dependence is the
mechanism, and the rest of this vignette is about using it, to understand how
big the bias is, and ultimately to correct for it.

# Building the curve

`pDetGivenNL` takes the same transmission loss table and source level
distribution as `pDetInArea` and `cde`, plus a vector of noise levels to
evaluate at instead of a distribution to draw from.

The transmission loss table and source levels below are the ones the rest of
this vignette uses throughout, and the detector is a plain logistic so this
section does not depend on anything else.

```{r build.curve}

R  <- 1e6
SL <- data.frame(mean = 190, sd = 4, sampleSize = 350)

TL <- simTLradials_20logR(maxRange = R, rangeStep = 1000, numTransects = 4)
TL <- TL[is.finite(rowSums(TL)) & TL[, 1] > 0, ]     # drop r = 0 if present

detector <- function(snr) plogis(snr, location = 1, scale = 2)

nlGrid <- seq(from = 55, to = 135, by = 1)
pCurve <- pDetGivenNL(nlGrid, detector, SL, TL)

curveDf <- data.frame(nl = nlGrid, pDet = pCurve)

```

## Noise level versus audible area

Before plotting it, be clear about what is on the axes, because there is no measured
data on this plot. Every point along the line asks *if the noise were
always exactly this loud, and never varied, what fraction of the study area
would you be hearing calls in?*

That fraction is just detection area divided by study area. So this plot is
really about detection range, which is the quantity you think in anyway. Add
the range at a few points and it becomes readable.

```{r curve.plot, warning=FALSE, fig.height=4}

# Range at which a call of mean source level sits at the detector's 50% point.
# Approximate. Ignores the spread in SL and the smoothness of the detector, so
# it is a label, not a calculation the curve depends on.
detRadius_km <- function(nl, SLmean = 190, loc = 1) {
  10^((SLmean - nl - loc) / 20) / 1000
}

marks <- data.frame(nl = c(74, 84, 94, 104))
marks$pDet  <- pDetGivenNL(marks$nl, detector, SL, TL)
marks$label <- sprintf("%.0f km", detRadius_km(marks$nl))

# Reference line: one decade per 10 dB, anchored at 90 dB. This slope is a
# direct consequence of spherical spreading (TL = 20 log10(r)) plus a
# step-function detector: to hold SNR constant, range must shrink by
# 10^(dNL/20) for each dB of extra noise, so area -- and therefore detection
# probability -- shrinks by 10^(dNL/10). It is tied to this TL model
# specifically, which is exactly why it is used as a reference line rather
# than the curve itself, and why the comparison against pyRAM later in this
# vignette anchors to the same line. It is a bare exponential with no ceiling,
# so it is clamped at 1 -- probability cannot exceed that, even if the formula
# it comes from does not know that.
anchor <- pDetGivenNL(90, detector, SL, TL)
refDf  <- data.frame(nl = nlGrid,
                     pDet = pmin(anchor * 10^(-(nlGrid - 90) / 10), 1))

ggplot(curveDf, aes(x = nl, y = pDet)) +
  geom_line(data = refDf, linetype = "dashed", colour = "grey50",
            linewidth = 0.6) +
  geom_line(linewidth = 0.9, colour = "steelblue") +
  geom_point(data = marks, size = 2, colour = "grey20") +
  geom_text(data = marks, aes(label = label), hjust = -0.25, vjust = -0.4,
            size = 3, colour = "grey20") +
  geom_vline(xintercept = 84, colour = "firebrick", linetype = "dotted") +
  annotate("text", x = 84.5, y = 1e-4, label = "true mean NL",
           colour = "firebrick", hjust = 0, size = 3) +
  scale_y_log10() +
  xlab("Noise level (dB)") +
  ylab("Fraction of the study area you can hear into") +
  labs(caption = "Labels: detection radius. Dashed grey: one decade per 10 dB, spherical spreading")

```

Work through two of those labelled points by hand and the plot stops being
abstract.

**At 84 dB.** A 190 dB call reaches the detector's 50% point when SNR is 1 dB,
so when TL is $190 - 84 - 1 = 105$ dB. Under spherical spreading that is
$10^{105/20}$ metres, or 178 km. So you are hearing calls out to roughly 178 km.
The area you can hear into is $\pi \times 178^2 \approx 99{,}500$ km². The study
area is $\pi \times 1000^2 \approx 3{,}140{,}000$ km². You are hearing into
about 3% of it.

**At 94 dB.** Every call now needs 10 dB more received level. Ten decibels of
spherical spreading is a factor of 3.16 in range. So 178 km becomes 56 km. The
area is $\pi \times 56^2 \approx 9{,}950$ km², one tenth of what it was. You are
hearing into 0.3% of the study area.

Ten decibels louder, ten times less ocean, ten times fewer calls. Each 10 dB
step to the right divides the curve by 10. On a log axis that is a straight
line, and that is all the dashed reference line is saying.

**The detector never entered that.** Move its threshold and the 178 km becomes
200 km or 150 km. The factor of ten per ten decibels does not budge, because it
came from the geometry of a circle, not from the detector. That is the single
most important thing on this plot, and it is why correcting the noise using a
property of the detector cannot be right.

**The shoulder on the left.** Below about 69 dB the labelled radius would exceed
1000 km. There is no ocean out there. It is past the edge of the study area, so
it is not yours to hear into. The curve flattens at 1 and making the sea
quieter buys nothing more. That shoulder is exactly where the tidy algebra
stops working, and it is why this vignette computes a curve rather than fitting
a slope to it.

**Where your data sits.** The red line marks the true mean noise level, out on
the straight portion, where a decibel of noise really does cost you ocean.

If this plot still feels abstract, skip ahead. The next figure is the same
curve on a linear axis, zoomed into the range your noise actually occupies, and
it is doing the real work.

# Illustrating why NL from detected calls is biased

Now use the curve. Nature draws a noise level for each call from a normal
distribution. The detection process then throws most of them away, and it
throws away loud ones far more eagerly than quiet ones. The curve says exactly
how much more eagerly.

```{r filter.plot, warning=FALSE, fig.height=4}

nlTrue  <- 84
nlSd    <- 4

fine  <- seq(nlTrue - 5 * nlSd, nlTrue + 5 * nlSd, length.out = 400)
fN    <- dnorm(fine, nlTrue, nlSd)
pd    <- pDetGivenNL(fine, detector, SL, TL)
wt    <- fN * pd

# Scale each curve to a common height so the shift is visible
filterDf <- rbind(
  data.frame(nl = fine, y = fN / max(fN),  what = "1. All calls (the truth)"),
  data.frame(nl = fine, y = pd / max(pd),  what = "2. Chance of detection"),
  data.frame(nl = fine, y = wt / max(wt),  what = "3. What you measure")
)

meanObs <- sum(fine * wt) / sum(wt)

ggplot(filterDf, aes(x = nl, y = y, colour = what)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = c("1. All calls (the truth)"   = "firebrick",
                                 "2. Chance of detection"      = "seagreen",
                                 "3. What you measure"         = "steelblue")) +
  geom_vline(xintercept = nlTrue,  colour = "firebrick", linetype = "dotted",
             linewidth = 1.1) +
  geom_vline(xintercept = meanObs, colour = "steelblue", linetype = "dashed",
             linewidth = 1.1) +
  annotate("segment", x = meanObs, xend = nlTrue, y = 1.05, yend = 1.05,
           arrow = arrow(length = unit(2, "mm"), ends = "both")) +
  annotate("text", x = (meanObs + nlTrue) / 2, y = 1.12,
           label = sprintf("bias = %.2f dB", nlTrue - meanObs), size = 3) +
  xlab("Noise level (dB)") +
  ylab("Scaled to a common height") +
  labs(colour = NULL) +
  theme(legend.position = "top")

```

Curve 1 is the truth. Curve 2 is the detection curve from the previous plot,
now drawn on a linear axis over the range the noise actually occupies. Curve 3
is curve 1 multiplied by curve 2, and it is what your detections show you.

Multiplying by a curve that leans left drags the peak left. That is the entire
bias. Nothing subtle is happening.

Two things worth noticing, because both get used later.

Curve 3 is narrower than curve 1 by very little. The width survives the
filtering almost intact, which means the standard deviation of your measured
noise is close to the truth even though the mean is not. That leaves only one
unknown to solve for.

The size of the shift depends on how steeply curve 2 leans, and on how wide
curve 1 is. It does not depend on where the detector sits. Move the detector
threshold and curve 2 changes height, not slope.

# Correcting the bias

The previous plot is a prediction machine. Give it a candidate true mean, and
it tells you what mean you should have measured.

We already know what we measured, and we have already estimated the
detection function (because it is required by `pDetInArea`), so we can run this
prediction machine backwards.

```{r rootfind.plot, warning=FALSE, fig.height=3.5}

muTry <- seq(80, 88, by = 1)
muObs <- vapply(muTry, function(m)
  predictSampledNL(mu = m, sigma = nlSd, detFun = detector, SL = SL, TL = TL),
  numeric(1))

rootDf <- data.frame(muTrue = muTry, muMeasured = muObs)

# The lookup itself: read across from the known measured value (y-axis) to
# the curve, then down to the candidate true mean that produced it. uniroot
# does this numerically; the arrows below do it visually, landing exactly on
# uniroot's own answer rather than on a separately-marked "true" position.
recovered <- uniroot(
  function(mu) predictSampledNL(mu = mu, sigma = nlSd, detFun = detector,
                                SL = SL, TL = TL) - meanObs,
  interval = c(meanObs, meanObs + 25))$root

xFloor <- min(muTry)
yFloor <- min(c(rootDf$muMeasured, meanObs))

ggplot(rootDf, aes(x = muTrue, y = muMeasured)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey70", linetype = "dashed") +
  geom_line(linewidth = 0.9, colour = "steelblue") +
  # Step 1: across, from what you measured to the point on the curve
  annotate("segment", x = xFloor, xend = recovered, y = meanObs, yend = meanObs,
           colour = "steelblue", linewidth = 1.1,
           arrow = arrow(length = unit(3, "mm"), type = "closed")) +
  # Step 2: down, from that point to the candidate true mean it came from
  annotate("segment", x = recovered, xend = recovered, y = meanObs, yend = yFloor,
           colour = "firebrick", linewidth = 1.1,
           arrow = arrow(length = unit(3, "mm"), type = "closed")) +
  annotate("point", x = recovered, y = meanObs, size = 2.5, colour = "black") +
  annotate("text", x = xFloor + 0.2, y = meanObs, hjust = 0, vjust = -0.7,
           size = 3, colour = "steelblue", label = "measured") +
  annotate("text", x = recovered, y = yFloor, hjust = -0.15, vjust = 1.3,
           size = 3, colour = "firebrick", label = "recovered NL") +
  coord_equal() +
  xlab("Candidate true mean NL (dB)") +
  ylab("Measured NL") +
  labs(caption = "Dashed grey: the 1:1 line you would get with no bias")

```

The blue line sits below the 1:1 line by the size of the bias. It rises
steadily, so there is exactly one candidate that reproduces your measurement.
The two arrows are the lookup, drawn: across from what you measured (blue) to
the point on the curve, then down (red) to read off the candidate that
produced it. That point and that landing spot are not illustrative
approximations -- they are `uniroot`'s own answer, plotted.

`nlFromDetections` is the package function that does exactly this (`uniroot`
over `predictSampledNL`, same shape as above). It also defaults to reading noise
measurements from a `NoiseRL` column, so it accepts an `snrInfo` table without
modification:

```{r nlfromdetections-demo}

nlFromDetections(snrInfo = data.frame(NoiseRL = rnorm(500, meanObs, nlSd)),
                 snrDetFun = detector, SL = SL, TL = TL)

```

That's a synthetic `snrInfo` built from the biased mean this vignette already
computed, just to show the call shape. `mean` should land close to `nlTrue`
(84 dB); `sd` is passed through from the input essentially unchanged, per the
note two paragraphs up about curve 3's width.

# Where this leaves the old function (pre callDensity v1.1.1)

Prior to callDensity version 1.1.1, `nlFromSnrInfo` was used to correct this
bias by adding the SNR at which the detection function reaches 0.5. However,
that quantity is a property of the detector, and the plots above say the bias is
a property of the geometry and the noise variance, and the two are close only by
coincidence for some detectors.

`nlFromSnrInfo` was always a stopgap. It corrects the bias using a property of
the detector (the SNR at its 50% point). This derived from the example scenario,
where it arrived at the correct answer coincidentally,
**not** because it was the thing driving the bias. `nlFromDetections` is derived
from the geometry that actually produces the bias, and is consistently less
biased across every detector and site this has been tested against.
`nlFromSnrInfo` is retained as a deprecated wrapper for old scripts, but there
is no situation where it should be preferred over `nlFromDetections`. `cde`
already defaults to `nlFromDetections` when `NL` is not supplied.

# Solution built from a lookup table, not a formula

`pDetGivenNL` was never built around spherical spreading, or any other
propagation formula. It reads transmission loss straight out of whatever
table you hand it, and evaluates the sonar equation against that table, cell
by cell. Swap the table for pyRAM output, AcTUP output, or anything else with
the same range and TL columns, and the function does not need to change,
because it never encoded an assumption about what the physics look like in
the first place. The dashed reference line on the first plot is not part of
that machinery: it is one candidate formula ($20\log_{10}r$), included there
as a comparison so any departure from it is visible, not as anything
`pDetGivenNL` relies on.

Spherical spreading is not a serious model for a baleen whale call at 25 Hz in
2000 m of water at 62°S. So below we use a real one. The file below is pyRAM
output computed on GLORYS12 sound speed profiles along 24 radial transects out
to 1000 km, for the South Kerguelen Plateau mooring in summer, at 25 Hz.

## First, look at the data

```{r pyram.inspect}

dim(kerguelen2015TL)
names(kerguelen2015TL)
range(kerguelen2015TL$range_m)

kerguelen2015TLmeta[, c("azimuths", "TruncationDistance_m")]

```

`kerguelen2015TL` and `kerguelen2015TLmeta` are TL data that ship with the
callDensity package (`?kerguelen2015TL`, `?kerguelen2015TLmeta`).
`kerguelen2015TLmeta` records where each of the 24 transects actually runs onto
land, using the pyRAM pipeline's bathymetry source
(`cmems_mod_glo_phy_my_0.083deg_static`).

The layout is already what `pDetGivenNL` wants: `range_m` first, then 24
columns named by bearing from `tl000` to `tl345`. Range runs 100 m to 1000 km
in 100 m steps, so 10,000 rows.

```{r pyram.prepare}

# Drop any non-finite cells. Range is already in metres and starts above zero.
TLpyram <- kerguelen2015TL[is.finite(rowSums(kerguelen2015TL)), ]

sprintf("%d transects, range step %.0f m, TL at 1000 km: %.1f to %.1f dB",
        ncol(TLpyram) - 1,
        diff(TLpyram$range_m)[1],
        min(as.numeric(TLpyram[nrow(TLpyram), -1])),
        max(as.numeric(TLpyram[nrow(TLpyram), -1])))

# Match truncation distances to transects by azimuth, not by row position --
# two objects agreeing on row order is an assumption, not a guarantee, and
# this is exactly the kind of silent misalignment that would be invisible in
# the output. tl015 gets the metadata row where azimuths == 15, full stop.
transectNames <- names(TLpyram)[-1]                      # tl000 .. tl345
transectAz    <- as.numeric(sub("^tl", "", transectNames))
pyramTrunc    <- kerguelen2015TLmeta$TruncationDistance_m[
                   match(transectAz, kerguelen2015TLmeta$azimuths)]
stopifnot("Every transect column must have a matching azimuth in kerguelen2015TLmeta" =
            !anyNA(pyramTrunc))

truncated <- pyramTrunc < max(TLpyram$range_m)
sprintf("%d of %d transects hit land before %.0f km: %s km (bearings %s)",
        sum(truncated), length(pyramTrunc), max(TLpyram$range_m) / 1000,
        paste(round(pyramTrunc[truncated] / 1000), collapse = ", "),
        paste(transectAz[truncated], collapse = ", "))

```

The spread in TL is very large, but that is because it is computed from every
bearing's row at 1000 km regardless of whether that bearing is still in water
there. Eight of the 24 bearings hit land between 410 and 950 km (bearings 120°
to 225°, roughly south through the plateau), so their rows beyond that point
describe whatever pyRAM did past the coastline, not propagation through open
water. `TLpyram[nrow(TLpyram), -1]` mixes those in with the sixteen genuinely
open-water bearings. It is a fair description of what the raw file contains, and
not yet a fair description of the site. For this reason, pDetGivenNL and
pDetInArea can take a truncationDistance as an input, and this can be used to mask (i.e. remove) land from the study area when estimating `NL` and `p_a`.

## The same curve, real physics

```{r pyram.curve, warning=FALSE, fig.height=4}

pPyram <- pDetGivenNL(nlGrid, detector, SL, TLpyram,
                      truncationDistance = pyramTrunc)

bothDf <- rbind(
  data.frame(nl = nlGrid, pDet = pCurve, what = "Spherical spreading"),
  data.frame(nl = nlGrid, pDet = pPyram, what = "pyRAM, SKP summer, 25 Hz")
)

anchorP <- pDetGivenNL(90, detector, SL, TLpyram, truncationDistance = pyramTrunc)
refBoth <- data.frame(nl = nlGrid,
                      pDet = pmin(anchorP * 10^(-(nlGrid - 90) / 10), 1))

ggplot(bothDf, aes(x = nl, y = pDet, colour = what)) +
  geom_line(data = refBoth, aes(x = nl, y = pDet), inherit.aes = FALSE,
            linetype = "dashed", colour = "grey50", linewidth = 0.6) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = 84, colour = "firebrick", linetype = "dotted") +
  scale_y_log10() +
  xlab("Noise level (dB)") +
  ylab("Fraction of the study area you can hear into") +
  labs(colour = NULL,
       caption = "Dashed grey: one decade per 10 dB, anchored on the pyRAM curve") +
  theme(legend.position = "top")

```

`truncationDistance = pyramTrunc` changes what "the study area" on the y-axis
means. Without it, the denominator is the full 1000 km circle, including the
land eight of these bearings never reach. With it, the denominator is the
actual water area `cde()` would use, via the same `td^2`-per-sector logic
`studyArea()` applies. This curve is now the one that matches what `p_a`
inside `cde()` actually estimates for this site, not a hypothetical circle.

What to look for, and what it means.

**The slope should not be one decade per 10 dB.** At 25 Hz in 2000 m of water
the sound is in a waveguide, not a free field. Beyond a few water depths it
spreads more like a cylinder than a sphere, which makes TL climb more slowly
with range. Slower TL means a decibel of extra noise costs you a wider annulus,
so the curve should fall *faster* than the dashed line, not slower.

If it does, that is the whole argument in one picture. The number
$\lambda = \ln(10)/10$ from the spherical derivation is simply wrong here, and
anything built on it inherits that error.

**The curve should be bumpy.** Convergence zones and bottom interaction add and
remove audible ocean at particular ranges. Wiggles are physics, not noise.

**The shoulder should be further left, or gone.** Flatter TL means longer
detection range, so it takes a quieter sea to run out of study area.

**One thing worth checking before trusting any of this.** `pDetGivenNL`'s only
approximation is the TL bin width, and a bumpy curve is exactly the case where
too coarse a bin could blur real structure away rather than just smooth noise.

```{r pyram.binwidth, warning=FALSE}

binCheck <- data.frame(binWidth = c(1, 0.5, 0.25, 0.1))
binCheck$pDet84 <- vapply(binCheck$binWidth, function(b)
  pDetGivenNL(84, detector, SL, TLpyram, truncationDistance = pyramTrunc,
             binWidth = b), numeric(1))
binCheck$nBins <- vapply(binCheck$binWidth, function(b)
  length(seq(40, 140, by = b)), numeric(1))

binCheck

```

Stable from 0.5 dB down, so `pDetGivenNL`'s default of 0.25 dB is fine here.
Nothing about using a real, bumpy curve instead of a synthetic one changes
that.

## Why your measurements would be biased here

Same mechanism as the spherical-spreading case, same three curves. The only
difference is which detection-probability curve does the filtering.

```{r pyram.filter, warning=FALSE, fig.height=4}

finePyram <- seq(nlTrue - 5 * nlSd, nlTrue + 5 * nlSd, length.out = 400)
fNPyram   <- dnorm(finePyram, nlTrue, nlSd)
pdPyram   <- pDetGivenNL(finePyram, detector, SL, TLpyram,
                        truncationDistance = pyramTrunc)
wtPyram   <- fNPyram * pdPyram

filterPyramDf <- rbind(
  data.frame(nl = finePyram, y = fNPyram / max(fNPyram),
            what = "1. All calls (the truth)"),
  data.frame(nl = finePyram, y = pdPyram / max(pdPyram),
            what = "2. Chance of detection"),
  data.frame(nl = finePyram, y = wtPyram / max(wtPyram),
            what = "3. What you measure")
)

measuredMasked <- sum(finePyram * wtPyram) / sum(wtPyram)

ggplot(filterPyramDf, aes(x = nl, y = y, colour = what)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = c("1. All calls (the truth)"   = "firebrick",
                                 "2. Chance of detection"      = "darkorange3",
                                 "3. What you measure"         = "steelblue")) +
  geom_vline(xintercept = nlTrue,       colour = "firebrick", linetype = "dotted",
            linewidth = 1.1) +
  geom_vline(xintercept = measuredMasked, colour = "steelblue", linetype = "dashed",
            linewidth = 1.1) +
  annotate("segment", x = measuredMasked, xend = nlTrue, y = 1.05, yend = 1.05,
           arrow = arrow(length = unit(2, "mm"), ends = "both")) +
  annotate("text", x = (measuredMasked + nlTrue) / 2, y = 1.12,
           label = sprintf("bias = %.2f dB", nlTrue - measuredMasked), size = 3) +
  xlab("Noise level (dB)") +
  ylab("Scaled to a common height") +
  labs(colour = NULL) +
  theme(legend.position = "top")

```

The bias at this real site is `r round(nlTrue - measuredMasked, 2)` dB.
Leaving the land mask out entirely -- `pDetGivenNL` with its default full
1000 km circle -- gives almost exactly the same number. That is not a
coincidence, and it does not generalise to `Dc` or `p_a`, so it is worth
being clear about why.

Adding the land mask back in only changes the bias if the water it adds back
is actually contributing to the weighted average that produces `measuredMasked`.
The near-identical numbers say it mostly isn't: whatever lies beyond each
bearing's truncation distance is already indistinguishable from zero
detection probability at every noise level this NL distribution reaches, so
restoring it inflates `pDetGivenNL`'s denominator (more nominal area) without
moving its numerator (that area was never contributing detections). A change
that dilutes a weighted mean by the same amount at every noise level cancels
out of `predictSampledNL`, because `predictSampledNL` only depends on the
curve's *shape*, not its absolute height. This is an inference from the two
numbers matching, not something independently confirmed here, but it fits the
picture from "The same curve, real physics" above: TL climbs quickly in this
shallow, low-frequency waveguide, so detection range is short relative to
1000 km, and the truncated bearings' far ends were already irrelevant.

`Dc` and `p_a` do depend on absolute height, not just shape, which is exactly
why the land mask matters for them and barely registers here. Land masking
being a big deal for `Dc`/`p_a` and a non-issue for NL are both true, and
both consequences of the same shape-versus-height distinction.

```{r pyram.seasons, eval=FALSE}

# WORTH DOING, NOT DONE. The bias is a property of the propagation, so it
# should vary by season as the sound speed profile changes, and by site. Loop
# the four seasons and the seven AWR sites and tabulate the bias. If it moves
# by more than a decibel between seasons, that is an argument for estimating NL
# seasonally, which is what Castro et al. (2024) recommend for f anyway.

```

# When does the correction still hold?

`nlFromDetections` substitutes the sample standard deviation of your
detections for the true standard deviation of the whole noise population, and
that substitution is where its residual bias comes from. It is a small
residual in every example so far, because in every example so far
`log(pDetGivenNL(nl))` has stayed close to straight over the width of the
noise distribution. A straight line in log-probability has no curvature, and
zero curvature means the substitution costs nothing: multiplying a Gaussian by
a perfect exponential shifts the mean and leaves the variance untouched.

Three things control how straight that line stays: how steep the detector is,
how far from a power law the propagation is, and how wide the noise
distribution is relative to both. None of the three is expressible as "the
propagation's fault" or "the detector's fault" in isolation, because
`pDetGivenNL` is a single curve built from both, and curvature is a property
of that combined curve, not of either ingredient separately. The only way to
find out which combinations stay safely straight is to sweep all three at
once.

## A sweep, computed exactly

The residual can be computed without simulating a single call. `meanObs` and
its standard deviation are both weighted moments of the same
`fN * pDetGivenNL(nl)` product already used to build the three-curve plots,
so both come from the same deterministic integral, no repeated sampling and
no Monte Carlo noise to average out.

```{r residual.function}

#' Residual bias in nlFromDetections's recovered mean, computed exactly.
#'
#' Builds the same weighted mean and weighted variance the three-curve plots
#' use, then inverts predictSampledNL twice: once with the true sd (which
#' should recover nlTrue exactly, as a check), and once with the sample sd of
#' the detections (which is what nlFromDetections actually receives). The
#' difference between the second recovery and nlTrue is the residual.
residualNL <- function(detFun, TLtab, truncDist, sd, nlTrue = 84) {

  fine <- seq(nlTrue - 6 * sd, nlTrue + 6 * sd, length.out = 600)
  fN   <- dnorm(fine, nlTrue, sd)
  pd   <- pDetGivenNL(fine, detFun, SL, TLtab, truncationDistance = truncDist)
  wt   <- fN * pd

  if (sum(wt) <= 0 || !all(is.finite(wt))) return(NA_real_)

  meanObs  <- sum(fine * wt) / sum(wt)
  sigmaObs <- sqrt(sum((fine - meanObs)^2 * wt) / sum(wt))

  recoveredSample <- tryCatch(
    uniroot(function(mu) predictSampledNL(mu = mu, sigma = sigmaObs,
                                          detFun = detFun, SL = SL,
                                          TL = TLtab,
                                          truncationDistance = truncDist) - meanObs,
           interval = c(meanObs, meanObs + 40))$root,
    error = function(e) NA_real_)

  recoveredSample - nlTrue
}

```

```{r residual.sweep}

# Detector steepness: plogis's scale parameter, smaller is steeper.
detScaleGrid <- c(0.5, 1, 2, 4, 8)

# Width of the true NL distribution, dB.
sdGrid <- c(1, 2, 3, 4, 6, 8)

# Mean NL -- the same four landmark values already used to label detection
# radius in "Noise level versus audible area", so this reuses points already
# seen rather than introducing new ones. For spherical spreading, moving this
# only matters near the shoulder (a ceiling-at-1 effect, unrelated to
# propagation); for pyRAM, it also matters because it slides the Gaussian
# window along a curve with real wiggles in it, and whether a given nlSd
# catches a smooth stretch or a kink depends on where the window sits.
nlTrueGrid <- c(74, 84, 94, 104)

propModels <- list(
  spherical = list(TL = TL,      truncationDistance = max(TL[[1]])),
  pyRAM     = list(TL = TLpyram, truncationDistance = pyramTrunc)
)

sweepGrid <- expand.grid(detScale    = detScaleGrid,
                         nlSd        = sdGrid,
                         nlTrue      = nlTrueGrid,
                         propagation = names(propModels),
                         stringsAsFactors = FALSE)

# This runs residualNL() once per row. Every call re-bins the TL table (that
# is how pDetGivenNL works), and uniroot calls predictSampledNL, which calls
# pDetGivenNL again, several dozen times per row while it converges. Every row
# is independent of every other row, so this is embarrassingly parallel --
# same future.callr backend callDensity's own parallel functions use
# elsewhere. future.seed is deliberately omitted: residualNL never touches the
# RNG, so there is no seed for future to manage, and no need to force
# RNGkind() the way an actual simulation-based sweep would require.
library(future)
library(future.apply)

future::plan(future.callr::callr, workers = future::availableCores(omit = 2))

sweepGrid$residual <- future_mapply(function(scale, sd, nlT, prop) {
  detFun <- function(snr) plogis(snr, location = 1, scale = scale)
  p <- propModels[[prop]]
  residualNL(detFun, p$TL, p$truncationDistance, sd, nlTrue = nlT)
}, sweepGrid$detScale, sweepGrid$nlSd, sweepGrid$nlTrue, sweepGrid$propagation,
future.packages = "callDensity")

future::plan(future::sequential)

```

```{r residual.plot, warning=FALSE, fig.height=5.5}

ggplot(sweepGrid, aes(x = nlSd, y = residual, colour = factor(detScale))) +
  geom_hline(yintercept = 0, colour = "grey60", linetype = "dashed") +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  facet_grid(propagation ~ nlTrue,
            labeller = labeller(nlTrue = as_labeller(function(x) paste0(x, " dB")))) +
  scale_colour_viridis_d(name = "Detector scale\n(smaller = steeper)") +
  xlab("True NL standard deviation (dB)") +
  ylab("Residual bias in nlFromDetections (dB)") +
  theme(legend.position = "right")

```

Two of the four things predicted here held up once this was actually run, and
two did not:

**Confirmed: the residual grows as the detector gets steeper (smaller
`detScale`).** A steeper transition packs the same total change in detection
probability into a narrower band of NL, which means more curvature in
log-probability at fixed `nlSd`.

**Confirmed: the residual grows with `nlSd`.** A wider distribution samples
further into whatever part of the curve has bent away from locally
exponential, regardless of why it bent.

**Wrong: "pyRAM should sit above spherical wherever it is visibly bumpier."**
At 74 and 84 dB pyRAM is *less* biased than spherical, the opposite of that
one-directional prediction. At 94 and especially 104 dB it is far worse,
including the one genuinely unexpected feature in this sweep -- a positive
residual (`nlFromDetections` overshooting `nlTrue`) around `nlSd` 4-6 at
`nlTrue` = 104, the only sign flip anywhere in the grid. Every other cell
undercorrects; that one family of cells overcorrects, which "more curvature"
alone cannot explain, since curvature magnitude does not change the sign.

**Wrong: "spherical should be flat except near the 74 dB shoulder."** The
knitted figure shows the spherical row peaking in magnitude around 84-94 dB
and getting *smaller* again at both 74 and 104, not flat except at one end.
A shoulder confined to the low-NL end cannot produce a pattern that improves
at both ends, so whatever is actually happening involves more of the curve's
shape than the shoulder argument alone accounts for. Both wrong predictions
are left here rather than quietly corrected, because the mechanism behind
either is not yet established -- see "Looking at the curve directly" below
for the follow-up built to chase it.

```{r residual.extremes}

sweepGrid[order(-abs(sweepGrid$residual)), ][1:5, ]
sweepGrid[order(abs(sweepGrid$residual)), ][1:5, ]

```

The largest and smallest rows say which regime this vignette's own examples
sit in. The spherical, `scale = 2`, `nlSd = 4`, `nlTrue = 84` case used
throughout the rest of this document is one specific point on this grid, not
necessarily a representative one -- worth checking where it actually falls
once this runs.

## Looking at the curve directly

The aggregate residual says something odd happens for pyRAM around `nlTrue` =
104, `nlSd` 4-6. It cannot say what, because it is a single number per cell,
already collapsed from the curve that produced it. "The same curve, real
physics" earlier in this document already plots `pDetGivenNL` on a log axis
over the full 55-135 dB range, which implicitly shows the same
`log(pDetGivenNL)` this section's curvature argument is about -- slope on
that plot is exactly the quantity that matters. But reading curvature by eye
off a full-range line plot is exactly the wrong tool for finding one
localised bend, so this zooms into the specific window the anomaly lives in
and plots the slope itself, rather than a curve to eyeball for slope changes.

```{r pyram.slope.diagnostic, warning=FALSE, fig.height=4.5}

# Zoom to roughly the union of the nlSd = 4 and nlSd = 6 windows around
# nlTrue = 104, where the sign flip in the sweep shows up. Faceted across
# three of the detector scales that showed the sign flip in the extremes
# table (0.5, 1, 2), since the single fixed detector used everywhere else in
# this vignette does not establish whether this shape is specific to that
# one detector or holds more generally.
nlFine <- seq(65, 145, by = 0.5)
slopeDetScales <- c(0.5, 1, 2)

slopeDf <- do.call(rbind, lapply(slopeDetScales, function(sc) {
  detFunSc <- function(snr) plogis(snr, location = 1, scale = sc)

  pPyramFine <- pDetGivenNL(nlFine, detFunSc, SL, TLpyram,
                            truncationDistance = pyramTrunc)
  pSpherFine <- pDetGivenNL(nlFine, detFunSc, SL, TL)

  rbind(
    data.frame(nl    = nlFine[-1] - diff(nlFine) / 2,
              slope = diff(log(pPyramFine)) / diff(nlFine),
              what  = "pyRAM, SKP summer, 25 Hz",
              detScale = sc),
    data.frame(nl    = nlFine[-1] - diff(nlFine) / 2,
              slope = diff(log(pSpherFine)) / diff(nlFine),
              what  = "Spherical spreading",
              detScale = sc)
  )
}))

ggplot(slopeDf, aes(x = nl, y = slope, colour = what)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = 104, colour = "firebrick", linetype = "dotted") +
  annotate("rect", xmin = 104 - 4, xmax = 104 + 4, ymin = -Inf, ymax = Inf,
           fill = "grey50", alpha = 0.15) +
  annotate("rect", xmin = 104 - 6, xmax = 104 + 6, ymin = -Inf, ymax = Inf,
           fill = "grey50", alpha = 0.10) +
  facet_wrap(~ detScale, ncol = 3,
            labeller = labeller(detScale = as_labeller(function(x) paste0("scale = ", x)))) +
  xlab("Noise level (dB)") +
  ylab("Local slope of log(pDetGivenNL) per dB") +
  labs(colour = NULL,
       caption = "Shaded bands: the nlSd = 4 and nlSd = 6 integration windows around nlTrue = 104") +
  theme(legend.position = "top")

```

What to look for, stated as questions rather than conclusions, since the
answer is whatever this plot actually shows:

Does the pyRAM reversal near 100-110 dB get sharper, shift position, or
disappear as the detector steepens from `scale` = 2 down to 0.5? The extremes
table showed the sign flip at all three of these scales, so the reversal
itself should survive across all three panels in some form -- if it does not,
that is worth understanding on its own.

Does the spherical curve's own mid-range dip and high-NL bend (visible at
`scale` = 2) look similar at `scale` = 0.5, or does a steeper detector wash
it out? A detector steep enough approaches a step function, which is the
scenario the original spherical-spreading derivation assumed -- if the bend
shrinks as the detector steepens, that would tie this directly back to the
detector's smoothness rather than the propagation model, for the spherical
case specifically.

The sharp upturn at the very right edge of every panel (~145 dB) is very
likely a numerical artefact of this diagnostic, not a real feature of either
propagation model: right at the edge of the `nlFine` grid, probabilities are
very small and the finite-difference derivative is more sensitive to
`binWidth` and floating point precision than to physics. Worth confirming
this by extending `nlFine` further and checking whether the upturn moves with
the edge of the grid, but I would not build any conclusion on that part of
the plot as it stands.

# Does the 2D solver actually recover the truth?

Before this solver ever sees real data, it is worth checking whether it can
recover a truth it is handed on a plate. The residual sweep above already has
ground truth built in -- every cell starts from a known `nlTrue` and `nlSd`
-- so running the same two-parameter inversion on those same synthetic cells,
and checking whether it gets back to the values that generated them, is the
natural validation. It is also a fair test of the solver's own machinery
(Nelder-Mead on an unconstrained least-squares objective, no bracket
guarantee the way `uniroot` has) on its own terms, separate from any question
about detector realism or real data.

```{r moment.functions}

#' Forward map: candidate true (mu, sigma) to the (mean, sd) you would
#' actually measure at detections. Same weighted-moment integral used
#' throughout this vignette, just returning both moments instead of one.
predictSampledMoments <- function(mu, sigma, detFun, SL, TL, truncDist,
                                  nNodes = 600) {
  fine <- seq(mu - 6 * sigma, mu + 6 * sigma, length.out = nNodes)
  fN   <- dnorm(fine, mu, sigma)
  pd   <- pDetGivenNL(fine, detFun, SL, TL, truncationDistance = truncDist)
  wt   <- fN * pd
  meanPred <- sum(fine * wt) / sum(wt)
  sdPred   <- sqrt(sum((fine - meanPred)^2 * wt) / sum(wt))
  c(mean = meanPred, sd = sdPred)
}

#' Inverse map: observed (mean, sd) to the candidate true (mu, sigma) that
#' would have produced them, via unconstrained least squares. sigma is
#' optimised on the log scale so no boundary hack is needed to keep it
#' positive. Returns the achieved objective alongside convergence, because
#' optim reporting convergence means its own stopping rule fired, not that
#' the fit is actually good -- a small residual is the thing that says that.
#' A proper package function would still want a dedicated 2D root-finder with
#' bracket guarantees, the way uniroot has for the 1D case; this does not.
recoverTrueMoments <- function(obsMean, obsSd, detFun, SL, TL, truncDist) {
  obj <- function(par) {
    pred <- predictSampledMoments(par[1], exp(par[2]), detFun, SL, TL, truncDist)
    sum((pred - c(obsMean, obsSd))^2)
  }
  fit <- optim(c(obsMean + 2, log(obsSd)), obj, method = "Nelder-Mead",
              control = list(reltol = 1e-12, maxit = 2000))
  c(muTrue = fit$par[1], sigmaTrue = exp(fit$par[2]),
    convergence = fit$convergence, objective = fit$value)
}

```

```{r solver.validation}

# Deliberately a much smaller grid than the residual sweep -- each cell here
# costs an optim() run on top of what residualNL() already costs, so the full
# 5 x 6 x 4 x 2 grid would be prohibitively slow. Three detector scales
# (steep, middling, shallow), three true sd values, and both nlTrue = 84 and
# 104 (a "normal" cell and the one where the sign flip lives), across both
# propagation models.
recover2D <- function(detFun, TLtab, truncDist, sd, nlTrue = 84) {

  fine <- seq(nlTrue - 6 * sd, nlTrue + 6 * sd, length.out = 600)
  fN   <- dnorm(fine, nlTrue, sd)
  pd   <- pDetGivenNL(fine, detFun, SL, TLtab, truncationDistance = truncDist)
  wt   <- fN * pd

  if (sum(wt) <= 0 || !all(is.finite(wt))) {
    return(c(muRecovered = NA_real_, sigmaRecovered = NA_real_,
            muError = NA_real_, sigmaError = NA_real_,
            convergence = NA_integer_, fitRMSE = NA_real_))
  }

  meanObs  <- sum(fine * wt) / sum(wt)
  sigmaObs <- sqrt(sum((fine - meanObs)^2 * wt) / sum(wt))

  fit <- tryCatch(
    recoverTrueMoments(meanObs, sigmaObs, detFun, SL, TLtab, truncDist),
    error = function(e) c(muTrue = NA_real_, sigmaTrue = NA_real_,
                          convergence = NA_integer_, objective = NA_real_))

  c(muRecovered    = unname(fit["muTrue"]),
    sigmaRecovered = unname(fit["sigmaTrue"]),
    muError        = unname(fit["muTrue"]) - nlTrue,
    sigmaError     = unname(fit["sigmaTrue"]) - sd,
    convergence    = unname(fit["convergence"]),
    fitRMSE        = sqrt(unname(fit["objective"]) / 2))
}

detScaleGrid2D <- c(0.5, 2, 8)
sdGrid2D       <- c(2, 4, 8)
nlTrueGrid2D   <- c(84, 104)

validationGrid <- expand.grid(detScale    = detScaleGrid2D,
                              nlSd        = sdGrid2D,
                              nlTrue      = nlTrueGrid2D,
                              propagation = names(propModels),
                              stringsAsFactors = FALSE)

library(future)
library(future.apply)
future::plan(future.callr::callr, workers = future::availableCores(omit = 2))

validationFits <- future_mapply(function(scale, sd, nlT, prop) {
  detFun <- function(snr) plogis(snr, location = 1, scale = scale)
  p <- propModels[[prop]]
  recover2D(detFun, p$TL, p$truncationDistance, sd, nlTrue = nlT)
}, validationGrid$detScale, validationGrid$nlSd, validationGrid$nlTrue,
validationGrid$propagation, future.packages = "callDensity")

future::plan(future::sequential)

validationGrid <- cbind(validationGrid, t(validationFits))

```

```{r solver.validation.plot, warning=FALSE, fig.height=4}

ggplot(validationGrid, aes(x = nlSd, y = muError, colour = factor(detScale))) +
  geom_hline(yintercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(size = 2) +
  geom_line(linewidth = 0.7) +
  facet_grid(propagation ~ nlTrue,
            labeller = labeller(nlTrue = as_labeller(function(x) paste0("nlTrue = ", x, " dB")))) +
  scale_colour_viridis_d(name = "Detector scale") +
  xlab("True NL standard deviation (dB)") +
  ylab("Recovered mu error (dB)") +
  labs(caption = "Should sit at zero everywhere, if the solver reliably recovers the truth")

```

```{r solver.validation.table}

validationGrid[order(-abs(validationGrid$muError)), 
              c("detScale", "nlSd", "nlTrue", "propagation",
                "muError", "sigmaError", "convergence", "fitRMSE")][1:10, ]

```

`muError` and `sigmaError` sit at roughly 1e-5 to 1e-6 dB across this whole
grid, with `convergence == 0` everywhere. That is a real validation of the
solver, not just a demonstration that it runs: whenever the truth actually is
Gaussian, this recovers it to a precision far tighter than anything else in
this vignette needs.

**On making this the default in `nlFromDetections`:** closer than it was,
but still not yet. This validation removes one of the concerns raised when
this was first built -- within the grid actually tested, the solver is
trustworthy, not just convergent. That grid is 36 cells, not the full
residual sweep's 240, so it has not been checked at every corner of detector,
propagation, `nlTrue` and `nlSd` this vignette has explored elsewhere; worth
widening before leaning on this claim at the true extremes of the parameter
space. Two things still stand between this and production, and neither is
about the solver:

The Gaussian-mismatch risk demonstrated below is not something a better
detector or a better solver fixes. It is a property of whichever real time
period the estimate is built over, and needs its own check before any
2D-corrected estimate is trusted, regardless of how good the solver itself
is.

The performance cost is real and matters more in production than in a
vignette: every evaluation here costs several hundred `pDetGivenNL` calls
worth of work, repeated inside `optim`'s search, versus one bracketed
`uniroot` search over a single pre-built curve for the current 1D method. If
`cde()` calls `nlFromDetections` inside a bootstrap or Monte Carlo loop, that
cost is paid on every replicate. The residual sweep above exists precisely to
show this cost is only worth paying in specific corners of parameter space --
an opt-in path, gated by that same sweep's criterion, protects the common
case's speed while still being available where the 1D residual is actually
large.

# What happens when this meets real seasonal data

The validation above used synthetic cells with a known, exactly Gaussian
`nlTrue`/`nlSd` behind them. Real detections carry no such guarantee. The
table below is real per-season data from Kerguelen2015 (@castro2024, Beyond
Counting Calls); `NLmean`/`NLsd` are exactly the *observed*, biased
moments `nlFromDetections` receives as input in practice -- the same
`meanObs`/`sigmaObs` this vignette has been computing throughout, just from
real detections instead of a synthetic distribution.

```{r seasonal.table}

seasonalNL <- data.frame(
  season = c("summer", "autumn", "winter", "spring", "year"),
  nc     = c(13248, 19123, 32631, 12903, 77905),
  f      = c(0.464, 0.243, 0.265, 0.388, 0.314),
  T      = c(2075.4, 2199.7, 2197.3, 2167.8, 8640.2),
  pa     = c(0.0658, 0.0192, 0.0701, 0.1836, 0.0789),
  NLmean = c(92.4, 96.8, 87.5, 81.0, 88.6),
  NLsd   = c(2.6, 3.7, 3.5, 2.4, 5.9)
)

seasonalNL

```

**This still uses the vignette's placeholder detector, not the actual fitted
detection function for Kerguelen2015** -- illustrative of how the solver
behaves on real-shaped moments, not corrected values for the manuscript.

`sdGrid`'s 1 to 8 dB range already comfortably covers every `NLsd` value in
this table, and the validation above already showed the solver recovers the
truth reliably across that whole range whenever the truth is actually
Gaussian, so there is no separate sdGrid-coverage question left to answer
here. What is worth actually running is `year`, since it is the row most
likely to expose what happens when the Gaussian assumption itself is wrong;
the four individual seasons would mostly reconfirm what the synthetic sweep
already established.

```{r seasonal.year}

yearFit <- recoverTrueMoments(88.6, 5.9, detector, SL, TLpyram, pyramTrunc)

data.frame(
  NLmean      = 88.6,
  muTrue      = yearFit["muTrue"],
  meanBias    = yearFit["muTrue"] - 88.6,
  NLsd        = 5.9,
  sigmaTrue   = yearFit["sigmaTrue"],
  sdBias      = yearFit["sigmaTrue"] - 5.9,
  convergence = yearFit["convergence"],
  fitRMSE     = sqrt(yearFit["objective"] / 2)
)

```

`year`'s recovered mean sits about 10.9 dB above the observed mean, and its
recovered sd (about 7.2 dB) is nearly double the largest individual season's
sd reported in the table above (3.7 dB, autumn). Both numbers came from a
solver just shown, on synthetic data, to recover the truth to within about
1e-5 dB. That combination is the point: the solver is not the source of this
discrepancy, so whatever is left has to be a property of the real data it
was asked to fit a single Gaussian to.

`fitRMSE` cannot catch this, and not because it happens to be small here by
chance. `recoverTrueMoments` has exactly two free parameters matching
exactly two targets -- observed mean and observed sd -- so for essentially
any finite pair of target moments, some `(mu, sigma)` exists that reproduces
them almost exactly. `year`'s `fitRMSE` (about 1.1e-6) is barely larger than
what a well-behaved season's `fitRMSE` would look like, despite `year` being
the case there is every reason to distrust. `fitRMSE` checks whether the
solver found *a* fixed point, not whether fitting a single Gaussian was the
right thing to do -- with only two moments feeding two parameters, it has no
way to see that.

**Hypothesis, not an established mechanism:** `year` aggregates four seasons
whose own observed means span roughly 81 to 97 dB. A single Gaussian forced
to stand in for a mixture of differently-centred Gaussians would generally
need to inflate its own variance to absorb the *between-season* spread on top
of each season's *within-season* variance -- the same decomposition behind
the law of total variance. `year`'s recovered sd coming out well above every
individual season's is consistent with that story, and it would be a more
direct signature of this failure mode than anything `fitRMSE` can offer. But
this vignette has shown one data point, not a validated diagnostic: nothing
here has checked that this pattern reliably shows up for genuine non-Gaussian
mixtures in general, or that it stays quiet for nested windows that really
are Gaussian throughout. Treat it as a plausible explanation worth testing,
not a mechanism this vignette has established.

If the hypothesis holds up, it would point to a real constraint on `cde`'s
`T`, on top of the ones already known: choosing a time period over which to
treat TL and detection probability as constant would not be enough by
itself, since the noise level distribution within that same period would
also need to be close enough to Gaussian for `nlFromDetections`, 1D or 2D, to
mean what it claims to mean. `year` looks like a `T` chosen too coarse for
that. Whether the individual seasons are themselves coarse enough to show
some of the same problem in miniature is a separate open question this table
cannot settle by itself -- their own biases (0.8 to 4.1 dB) are compatible
with some residual non-Gaussianity, not only with the small solver/detector
residual the synthetic validation already bounds. Checking whether recovered
sd inflates when a candidate `T` is built by aggregating finer sub-periods is
a plausible diagnostic for that question, contingent on the hypothesis above
actually validating.

Non-Gaussian and non-parametric noise level distributions are out of scope
for this vignette. Every function here -- `pDetGivenNL`, `predictSampledNL`,
`nlFromDetections`, and the exploratory 2D solver above -- assumes a Gaussian
`fN`. Real seasonal NL distributions from Kerguelen2015 are available and not
obviously Gaussian by eye (winter in particular looks bimodal, and `year`
looks like a clear mixture, consistent with the hypothesis above), which
would make them a natural next step if this ever gets pursued further, but
fitting other parametric families or a non-parametric density is a
substantially harder problem than anything solved here, and deliberately not
attempted in this vignette.

# What is still open

`pDetGivenNL`, `predictSampledNL` and `nlFromDetections` are all in the
package now, and `cde` calls `nlFromDetections` by default whenever `NL` is
not supplied, so `nlFromSnrInfo` is a deprecated wrapper rather than the
active path. Two things noted while building this are still genuinely open.

`pDetGivenNL` still duplicates work that `pDetInArea` already does. Both thin
the transmission loss table, broadcast the truncation distance across
transects, mask cells beyond truncation, and form area weights. That
preparation should be pulled out of `pDetInArea` into a shared helper and used
by both. Doing it any other way means the two drift apart.

The standard deviation is taken straight from the detections, and "When does
the correction still hold?" above is the sweep that decides whether that
matters. A first attempt at solving for the mean and standard deviation
jointly now exists ("Does the 2D solver actually recover the truth?" and
"What happens when this meets real seasonal data?" above), built with
`optim` rather than a proper bracketed 2D root-finder. Validated against
synthetic ground truth, it recovers the truth to within about 1e-5 dB --
detector data was never the limiting factor. Real seasonal data exposed a
different, more fundamental problem instead: the correction assumes the true
NL distribution is Gaussian, and whether that holds depends on how the time
period `T` was chosen, not on the solver or the detector. Promoting this into
the package, and deciding whether it becomes a default or an opt-in path,
should wait on that question having its own answer, not just on fixing the
solver's remaining robustness gaps.

# What this does not fix

Two assumptions run through every plot above and neither is touched by any of
this.

Calls are assumed spread evenly in two dimensions. The area weight
` r * (r <= td)` in `pDetGivenNL` is that assumption written down. Clustered
calling, or calls concentrated at particular depths, changes it.

Noise is assumed independent of whether whales are calling. If loud hours are
also whale hours, the detection curve is not the only thing shaping curve 3,
and the correction is solving the wrong problem accurately.

# References
