---
title: "Using all available information: a union call density estimate"
output: 
  html_document: default
  pdf_document: default
  rmarkdown::html_vignette:
    vignette: >
      %\VignetteEncoding{UTF-8}
      %\VignetteIndexEntry{Using all available information: a union call density estimate}
      %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.height = 4,
  comment = "#>"
)
```

```{r setup}
library(callDensity)
library(VGAM)
library(dplyr)
library(ggplot2)
library(kableExtra)
```

# Left on the table

@miller2026 introduced adjudicated capture-recapture (CR) as a way to characterise detectors without assuming any one observer is ground truth, and argued that the method "readily accommodates all available information from multiple observers and/or detectors, if more than one of either is available." What the paper actually demonstrated, though, was CR producing a call density estimate *for* each detector -- CR1S, CR2S, CRAS and so on in its own terminology -- never a single estimate that draws on more than one detector's own detections at once. The claim about using all available information was true of the underlying capture-recapture model (which is exactly what lets it estimate a false discovery rate without a trusted observer in the first place), but the density estimates built on top of it stayed one-detector-at-a-time throughout.

This vignette closes that gap directly. The capture-recapture model already estimates, as a normal step in fitting it, the probability that *at least one* of several detectors would catch a given call -- this is a standard quantity in double-observer distance sampling, and callDensity's own `vglmDetectionProb()` already computes it internally as an intermediate step toward each individual detector's own detection probability. Surfacing it directly, rather than discarding it, gives a union detection function: fit one capture-recapture model jointly on two (or more) detectors, and read off not just "detector 1's own probability of detecting a call" or "detector 2's own probability," but "the probability that detector 1 *or* detector 2 detects it." Paired with the union of their own raw detection counts -- how many calls did *either* detector flag, across the full dataset -- this gives a call density estimate that uses both detectors' information at once, rather than two separate, redundant analyses of the same underlying calls.

The simulation below follows exactly the same setup as `vignette("callDensity_CommonGround")` -- the same call population, the same two detectors, the same adjudicated-subsample workflow -- so the two vignettes can be read side by side. What's new here is entirely in how the existing per-detector pieces (`Nc`, false discovery rate, detection probability) get combined into one union estimate, not in how any of them are individually derived.

# Simulate a shared call population and two detectors

The call population, source level, noise level, and transmission loss are exactly as in `callDensity_CommonGround`: a uniform distribution of calls in space and time, with acoustic properties following the passive sonar equation.

```{r simulate.calls, warning=FALSE, message=FALSE}
n <- 2e5        # number of simulated calls, whether detected or not
R <- 1e6        # radius in m
k <- 1          # number of sensors
minDate <- as.POSIXct("2025-01-01")
maxDate <- as.POSIXct("2026-01-01")
Time <- as.numeric(difftime(maxDate, minDate, unit = "days")) / 365

A <- studyArea(R / 1e3)
TrueCallDensity <- n / (A * Time)

set.seed(1)
sim <- simCallLocation(n = n, R = R, minDate = minDate, maxDate = maxDate)

SL <- data.frame(mean = 190, sd = 4, sampleSize = 350)
NL <- data.frame(mean = 84, sd = 4, sampleSize = n)
tlFunc <- function(r) 20 * log10(r)
sim <- simCallAcoustics(sim, SL, NL, TL = tlFunc)

TL <- simTLradials_20logR(maxRange = R, rangeStep = 100, numTransects = 8)

cat("Study area (km^2):", A, "\n")
cat("True call density (calls / km^2 / year):", TrueCallDensity, "\n")
```

Two detectors, with the same parameters as `callDensity_CommonGround`: detector 1 is the more sensitive, more reliable of the two (lower threshold, lower false discovery rate); detector 2 is noisier on both counts.

```{r parameterise.detectors}
det1params <- data.frame(location = 1, scale = 2, func = 'plogis',
                         c = 0.1, fpMean = 0, fpSD = 2)
det2params <- data.frame(location = 2, scale = 4, func = 'plogis',
                         c = 0.3, fpMean = 0, fpSD = 2)

simDet1 <- simulateDetector(det1params, sim)
simDet2 <- simulateDetector(det2params, sim)
```

# Two capture history tables: full year, and adjudicated subsample

Two different tables are needed here, matching a distinction that already exists for a single detector -- `Nc` always comes from the full dataset, while the capture history table used to fit a detection function and estimate a false discovery rate comes from a small adjudicated subsample. The union case needs both, in the same shapes as before: a full-year table across both detectors for the union `Nc`, and an adjudicated subsample across both detectors for the joint capture-recapture fit.

```{r match.full.year}
# Full year, both detectors -- this is what the union Nc comes from. Both
# detectors examined the exact same simulated year, so unionDetections()'s
# fullCoverageConfirmed requirement is satisfied here by construction (see
# ?unionDetections for why this can't be verified automatically from the
# table alone, and has to be confirmed explicitly instead).
capHistTabFull <- simsTocaptureHistoryTable(simDet1, simDet2)

observerCols <- c("detect_observer1", "detect_observer2")

Nc1     <- sum(simDet1$detect_table)
Nc2     <- sum(simDet2$detect_table)
NcUnion <- unionDetections(capHistTabFull, observerCols, fullCoverageConfirmed = TRUE)

cat("Nc, detector 1 alone: ", Nc1, "\n")
cat("Nc, detector 2 alone: ", Nc2, "\n")
cat("Nc, union of both:    ", NcUnion,
    " (between max(Nc1,Nc2) and Nc1+Nc2, as it must be)\n")
```

```{r match.subsample}
# Adjudicated subsample, both detectors -- matches callDensity_CommonGround's
# own 1-hour-in-41 subsampling. groundTruth here is the simulated ground
# truth standing in for a judge's verdict, exactly as in that vignette.
subsampleDet1 <- subsampleSimInTime(simDet1, interval = "41 hour")
subsampleDet2 <- subsampleSimInTime(simDet2, interval = "41 hour")

capHistTab <- simsTocaptureHistoryTable(subsampleDet1, subsampleDet2)
# cde()'s own false-discovery-rate handling needs a single consolidated SNR
# column -- a separate mechanism from chtToSNRinfo()'s signalCol/noiseCol
# averaging used below.
capHistTab$SNR <- rowMeans(capHistTab[, c("snr_observer1", "snr_observer2")], na.rm = TRUE)

adjudicated <- subset(capHistTab, capHistTab$groundTruth &
                        (capHistTab$detect_observer1 | capHistTab$detect_observer2))

cat("Adjudicated subsample:", nrow(adjudicated), "true-positive events\n")
```

# One joint model, three ways to read it

Fitting a capture-recapture model with both detectors' columns as the two occasions estimates their detection probabilities jointly -- this part doesn't change at all from a normal two-detector CR analysis. What's new is asking for three different things back from the same fit, via `whichObserver`: detector 1's own probability, detector 2's own probability, and `"any"` for the probability that at least one of them detects the call.

```{r fit.joint.model, warning=FALSE, message=FALSE}
fitObs1  <- fitDetFun(adjudicated, modelType = "vglm", yColNames = observerCols,
                      whichObserver = "detect_observer1")
fitObs2  <- fitDetFun(adjudicated, modelType = "vglm", yColNames = observerCols,
                      whichObserver = "detect_observer2")
fitUnion <- fitDetFun(adjudicated, modelType = "vglm", yColNames = observerCols,
                      whichObserver = "any")
```

```{r plot.detection.functions, warning=FALSE, fig.height=4}
models <- list("Detector 1" = fitObs1, "Detector 2" = fitObs2, "Union (either)" = fitUnion)

# distribution="none": a union model's own "missed" set is empty by
# construction here (adjudicated is already filtered to events flagged by
# at least one detector, so nothing is left to be "missed by the union"),
# which the density panels below have no data for. The three curves
# themselves make the point without needing that panel.
showDetFun(models, distribution = "none", rug = FALSE)
```

The union curve sits above both individual curves at every SNR, as it must: the probability that *either* detector catches a call can never be lower than either detector's own probability on its own.

# Three call density estimates

`cde()` itself needs no new arguments for any of this beyond what already exists: `observerCol` becomes the vector `c("detect_observer1", "detect_observer2")` for the union case, matching the same detectors the model was fit with `whichObserver = "any"` for -- `cde()` checks that these two agree and refuses to silently combine a union false discovery rate with a single detector's own detection probability, or the reverse. `NL` is left for `cde()` to estimate internally via `nlFromDetections()`, exactly as it would for a single detector -- no special handling needed for the union case.

```{r cde.individual, warning=FALSE, message=FALSE}
res1 <- cde(Nc = Nc1, capHistTab = capHistTab, snrDetFun = fitObs1,
           SL = SL, TL = TL, A = A, T = Time, k = k,
           groundTruthCol = "groundTruth", observerCol = "detect_observer1",
           signalCol = c("signalRMSdB_observer1", "signalRMSdB_observer2"),
           noiseCol  = c("noiseRMSdB_observer1",  "noiseRMSdB_observer2"))

res2 <- cde(Nc = Nc2, capHistTab = capHistTab, snrDetFun = fitObs2,
           SL = SL, TL = TL, A = A, T = Time, k = k,
           groundTruthCol = "groundTruth", observerCol = "detect_observer2",
           signalCol = c("signalRMSdB_observer1", "signalRMSdB_observer2"),
           noiseCol  = c("noiseRMSdB_observer1",  "noiseRMSdB_observer2"))
```

```{r cde.union, warning=FALSE, message=FALSE}
resUnion <- cde(Nc = NcUnion, capHistTab = capHistTab, snrDetFun = fitUnion,
                SL = SL, TL = TL, A = A, T = Time, k = k,
                groundTruthCol = "groundTruth", observerCol = observerCols,
                signalCol = c("signalRMSdB_observer1", "signalRMSdB_observer2"),
                noiseCol  = c("noiseRMSdB_observer1",  "noiseRMSdB_observer2"))
```

# Results

```{r results.table, echo=TRUE, warning=FALSE}
resultsTrue <- data.frame(model = "Truth", Nc = n, c = NA, pa = NA,
                          Dc = TrueCallDensity, CV.Dc = 0)

results <- rbind(
  resultsTrue,
  data.frame(model = "Detector 1 alone", Nc = Nc1, c = res1$c, pa = res1$pa,
            Dc = res1$Dc, CV.Dc = res1$CV.Dc),
  data.frame(model = "Detector 2 alone", Nc = Nc2, c = res2$c, pa = res2$pa,
            Dc = res2$Dc, CV.Dc = res2$CV.Dc),
  data.frame(model = "Union (both)", Nc = NcUnion, c = resUnion$c, pa = resUnion$pa,
            Dc = resUnion$Dc, CV.Dc = resUnion$CV.Dc)
)
results$DcFraction <- results$Dc / TrueCallDensity

kableExtra::kbl(results, digits = c(NA, 0, 3, 4, 4, 3, 2), row.names = FALSE,
                col.names = c("Model", "$N_c$", "$\\hat{c}$", "$\\hat{p}_a$",
                             "$D_c$", "$CV.D_c$", "$D_c$ / true $D_c$")) %>%
  kableExtra::kable_classic(full_width = FALSE)
```

```{r results.plot, fig.height=3.5}
plotData <- subset(results, model != "Truth")
plotData$model <- factor(plotData$model, levels = plotData$model)

ggplot2::ggplot(plotData, ggplot2::aes(x = model, y = Dc, fill = model)) +
  ggplot2::geom_col(width = 0.6) +
  ggplot2::geom_errorbar(ggplot2::aes(ymin = Dc * (1 - CV.Dc), ymax = Dc * (1 + CV.Dc)),
                        width = 0.15) +
  ggplot2::geom_hline(yintercept = TrueCallDensity, linetype = "dashed", colour = "grey30") +
  ggplot2::annotate("text", x = 0.6, y = TrueCallDensity, label = "Truth",
                    vjust = -0.5, hjust = 0, colour = "grey30", size = 3) +
  ggplot2::labs(x = NULL, y = expression(D[c]~(calls~km^-2~year^-1)), fill = NULL) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "none")
```

The union estimate's error bars ($\pm 1$ CV) are visibly narrower than either individual detector's own -- the direct consequence of the union false discovery rate being estimated from more adjudicated events than either detector's own count alone (every event either detector flagged, rather than just one detector's own flags), which is the one piece of this that improves unconditionally regardless of what the two detectors' own probability curves happen to look like.

# When this helps, and when it might not

The false discovery rate's own coefficient of variation is smaller for the union than for either individual detector, always -- it only depends on how many adjudicated events went into estimating it, and the union pools strictly more of them (every event either detector flagged) than either detector's own count alone does. That part of the improvement is guaranteed by construction, not a property of this particular simulation.

The detection probability's own uncertainty is a different matter, and it's the one that actually decides whether the union wins overall. The union curve is a genuinely new quantity, not simply a better-sampled version of either individual curve, and its own precision depends on how well the joint model's coefficients happen to be estimated in a given dataset -- which is not guaranteed to be tighter than an individual detector's own curve when that detector is already well characterised on its own. Running this same simulation across several random seeds makes the pattern concrete rather than hypothetical: the union's $CV.D_c$ came out below detector 1's every time, but only sometimes below detector 2's -- detector 2 (the higher-threshold, noisier detector by construction here) happens to have a well-identified curve of its own in some replicates, tight enough that its own $CV.D_c$ beats the union's despite the union's false discovery rate still being the more precisely estimated of the two. Detector 1's own estimate, by contrast, never won outright in any replicate tried.

So the honest summary is: the union's false discovery rate is unconditionally tighter, and the union is never worse than the *worse* of the two detectors here -- but it is not a guarantee of beating the *better* one, since that comes down to how well that detector's own curve happens to be identified in a given dataset, a separate question from the union's own construction.

What the union approach always buys, regardless of how that particular contest goes, is using detections that a single-detector analysis discards outright: a call caught only by detector 2 contributes nothing to a "detector 1 alone" estimate's own count, adjudicated sample, or detection function, even though it is exactly the kind of information adjudicated capture-recapture was introduced to make use of.

# Scaling to more detectors

Automated detectors are becoming a commodity for some species and call types -- running ten mediocre detectors costs little more than running two, unlike the human observer effort adjudicated capture-recapture was originally built around. If that's the actual situation, the two-detector case above is really a lower bound on what's available. This section repeats the same union analysis for 3, 6, and 10 detectors of varying quality, sharing the same underlying call population, to see how the false discovery rate's guaranteed improvement (more adjudicated positives to estimate it from) plays out as the pool grows.

The ten detectors span a plausible "commodity" range -- threshold, steepness, and false discovery rate all spread from fairly good to fairly mediocre -- and the 3/6/10-detector comparisons below use the first 3, first 6, and all 10 of the same pool, so adding detectors here means literally adding to the same set rather than swapping in a different one.

```{r ndetectors.setup}
nMax <- 10
locations <- seq(1, 4, length.out = nMax)
scales    <- seq(2, 4, length.out = nMax)
fdrs      <- seq(0.1, 0.4, length.out = nMax)

detParamsN <- lapply(seq_len(nMax), function(i) {
  data.frame(location = locations[i], scale = scales[i], func = 'plogis',
            c = fdrs[i], fpMean = 0, fpSD = 2)
})

simDetsN <- lapply(seq_len(nMax), function(i) simulateDetector(detParamsN[[i]], sim))
```

```{r ndetectors.analysis, warning=FALSE, message=FALSE}
runUnionAnalysis <- function(N) {
  detsSubset <- simDetsN[1:N]
  subsampled <- lapply(detsSubset, function(d) subsampleSimInTime(d, interval = "41 hour"))

  capHistTabFullN <- do.call(simsTocaptureHistoryTable, detsSubset)
  capHistTabN <- do.call(simsTocaptureHistoryTable, subsampled)
  capHistTabN$SNR <- rowMeans(capHistTabN[, paste0("snr_observer", seq_len(N))], na.rm = TRUE)

  observerColsN <- paste0("detect_observer", seq_len(N))
  signalColsN   <- paste0("signalRMSdB_observer", seq_len(N))
  noiseColsN    <- paste0("noiseRMSdB_observer", seq_len(N))

  NcUnionN <- unionDetections(capHistTabFullN, observerColsN, fullCoverageConfirmed = TRUE)

  adjudicatedN <- subset(capHistTabN, capHistTabN$groundTruth &
                           Reduce(`|`, as.list(capHistTabN[observerColsN])))

  fitUnionN <- fitDetFun(adjudicatedN, modelType = "vglm", yColNames = observerColsN,
                         whichObserver = "any")

  resUnionN <- cde(Nc = NcUnionN, capHistTab = capHistTabN, snrDetFun = fitUnionN,
                   SL = SL, TL = TL, A = A, T = Time, k = k,
                   groundTruthCol = "groundTruth", observerCol = observerColsN,
                   signalCol = signalColsN, noiseCol = noiseColsN)

  data.frame(N = N, Nc = NcUnionN, c = resUnionN$c, pa = resUnionN$pa,
            Dc = resUnionN$Dc, CV.Dc = resUnionN$CV.Dc,
            DcFraction = resUnionN$Dc / TrueCallDensity)
}

scalingResults <- do.call(rbind, lapply(c(2, 3, 6, 10), runUnionAnalysis))

kableExtra::kbl(scalingResults, digits = c(0, 0, 3, 4, 4, 3, 2), row.names = FALSE,
                col.names = c("N detectors", "$N_c$", "$\\hat{c}$", "$\\hat{p}_a$",
                             "$D_c$", "$CV.D_c$", "$D_c$ / true $D_c$")) %>%
  kableExtra::kable_classic(full_width = FALSE)
```

```{r ndetectors.plot, fig.height=3.5}
ggplot2::ggplot(scalingResults, ggplot2::aes(x = N, y = CV.Dc)) +
  ggplot2::geom_line() +
  ggplot2::geom_point(size = 2) +
  ggplot2::scale_x_continuous(breaks = scalingResults$N) +
  ggplot2::labs(x = "Number of detectors in the union", y = expression(CV.D[c])) +
  ggplot2::theme_bw()
```

$CV.D_c$ falls overall as the detector pool grows -- confirmed as a general pattern across several random seeds during development, not just the one run shown here -- though not perfectly monotonically at every single step (the two- to three-detector step above is a small, genuine wobble, not smoothed over). The reason is the same one from the two-detector case, just compounding: each additional detector adds more events to the union's adjudicated positive set, which is what the false discovery rate's precision depends on directly. Ten mediocre commodity detectors pooled this way substantially outprecision two decent ones -- an argument for breadth over quality when detectors are genuinely cheap to run in parallel.

# Detector agreement and false positive rate

@miller2026 noted, without demonstrating it directly, an inverse relationship between how many detectors flag a given event and how likely that event is to be a false positive. The ten-detector pool above gives enough range in "how many agreed" (0 through 10) to check this directly: for every event flagged by at least one of the ten, how many detectors flagged it, and was it actually real?

```{r agreement.analysis}
# runUnionAnalysis() above builds its own capture history table internally
# and doesn't return it -- rebuilding the N=10 case explicitly here, rather
# than threading it out as an extra return value for a one-off use.
detsSubset10 <- simDetsN[1:10]
subsampled10 <- lapply(detsSubset10, function(d) subsampleSimInTime(d, interval = "41 hour"))
capHistTab10 <- do.call(simsTocaptureHistoryTable, subsampled10)
observerCols10 <- paste0("detect_observer", 1:10)

unionPositive <- subset(capHistTab10, Reduce(`|`, as.list(capHistTab10[observerCols10])))
unionPositive$nAgree <- rowSums(unionPositive[, observerCols10])

agreementTable <- unionPositive %>%
  dplyr::group_by(nAgree) %>%
  dplyr::summarise(nEvents = dplyr::n(),
                   fpRate = mean(!groundTruth))

ggplot2::ggplot(agreementTable, ggplot2::aes(x = nAgree, y = fpRate)) +
  ggplot2::geom_col(fill = "steelblue") +
  ggplot2::geom_text(ggplot2::aes(label = nEvents), vjust = -0.3, size = 3) +
  ggplot2::scale_x_continuous(breaks = 1:10) +
  ggplot2::labs(x = "Number of detectors agreeing", y = "Empirical false positive rate",
               caption = "Label above each bar: number of events at that agreement count") +
  ggplot2::theme_bw()
```

The direction is exactly as expected -- agreement concentrated entirely among single-detector events, and the false positive rate drops from most of that single-detector bucket to essentially none of it once two or more detectors agree. But the shape here is sharper than real data is likely to show, and it's worth being explicit about why: in this simulation, each detector's own false positives are independently, randomly timed, so two detectors' false alarms coinciding by chance is vanishingly unlikely by construction -- any event two or more detectors agree on here is, definitionally, almost certain to be a real call. Real false alarms are not always like that. A shared, correlated source -- another species' overlapping calls, a consistent noise transient, ice or vessel noise with its own predictable signature -- could plausibly fool several detectors at once, especially ones built on similar underlying features. The direction of the relationship (more agreement, fewer false positives) should still hold, since it follows from genuine calls being the one thing every detector is actually built to find, but the real-data version of this curve likely declines more gradually than the near step-function shown here, rather than dropping to exactly zero after the first additional detector agrees.

# References
