---
title: "callDensity_snrThreshold"
output: 
  html_document: default
  pdf_document: default
  rmarkdown::html_vignette:
    vignette: >
      %\VignetteEncoding{UTF-8}
      %\VignetteIndexEntry{callDensity_snrThreshold}
      %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.height = 3.75, 
  comment = "#>"
)
```

```{r setup}
library(callDensity)
library(ggplot2)
library(scales)
library(kableExtra)
```

# Further exploration and topics in callDensity package (TODO)

1)  Spatial distribution of calls
    A)  Effect of land, ice, and other zero-probability areas
    B)  Effect of non-uniform distribution of calls in area (clustered/concentrated)
2)  **Truncation in SNR**
3)  **Detectors with different slopes**
4)  Minimum number of recaptures/effort required to get a reliable estimate
5)  More realistic Transmission Loss models and effects of mismatch


As usual, we start with the density equation:

\begin{equation}
D_c = \frac{N_c(1-c)}{kTp_aA}
\end{equation}

where: $D_c$ is call density, $N_c$ is number of calls, $c$ is false discovery rate, $k$ is number of sensors (here always 1), $T$ is the duration of data analysed (in hours), $p_a$ is the probability of detection in the study area, and $A$ is the study area (in km\^2).

## Effects of truncation in SNR

Scenario: Antarctic blue whales in a quiet ocean with spherical spreading of sounds. 

In this example, we investigate the effects of excluding detections with an SNR that are below a threshold value.  

```{r call.spatial.distribution, warning=FALSE, message=FALSE}

n = 1e6; # number of simulated calls (regardless of whether detected or not); 
R = 1e6; # radius of 1000 km (but in m)
k <- 1   # number of sensors
minDate <- as.POSIXct("2025-01-01")
maxDate <- as.POSIXct("2026-01-01")
Time <- as.numeric(difftime(maxDate,minDate,unit="days"))/365 # in years

# Call density, D_c is n/A 
A = studyArea(R/1e3) # Circular study area in km^2
TrueCallDensity = n/(A*Time) # calls/km^2/time

# 1) Generate n uniformly distributed calls within radius R and time period Time
sim <- simCallLocation(n=n, R=R, minDate=minDate, maxDate=maxDate)

```
Sonar equation parameters for this scenario:

\begin{equation}
TL = 20  log_{10}  d
\end{equation}
\begin{equation}
SNR = SL-TL-NL
\end{equation}


```{r sonar.eq}

SL <- data.frame(mean=190, sd=4, sampleSize=350) # True SL distribution for sim
NL = data.frame(mean=84, sd = 4, sampleSize = n)  # True NL distribution for sim
tlFunc <- function(r)20*log10(r)

# Simulate the true acoustic properties of each call and false positive 
sim <- simCallAcoustics(sim, SL, NL, TL = tlFunc)

# Per-transect estimate of TL (used by callDensity package to estimate pDet)
TL <- simTLradials_20logR(maxRange=R, rangeStep=100, numTransects=8)

```

### Simulate detectors:

```{r parameterise.detectors, warning=FALSE, message=FALSE}

#Specify parameters for detector 1
# Good detector with low threshold (location) and low false positive rate.
det1params = data.frame( 
  location=1,    # AKA intercept?
  scale=1,       # AKA slope?
  func='plogis', # logistic function
  c=0.1,         # False discovery rate
  fpMean=0,      # Mean of distribution of false positives (in dB SNR)
  fpSD = 2      # Standard deviation false positive distribution (in dB SNR)
)

# Detector 2 has a higher threshold (location), scale, and false positive rate
# than detector 1.
det2params = data.frame( 
  location=3,    # AKA intercept?
  scale=0.5,       # AKA slope?
  func='plogis', # logistic function
  c=0.3,         # False discovery rate
  fpMean=0,      # Mean of distribution of false positives (in dB SNR)
  fpSD = 2      # Standard deviation false positive distribution (in dB SNR)
)
```

Now we create a function that will take the detector parameters above and simulate the detection process. We then call this function with each set of parameters to create a simulation for each detector.

```{r simulate.detectors, warning=FALSE, message=FALSE}

# Simulate two detectors each with the different parameters
simDet1<- simulateDetector(det1params,sim)
simDet2<- simulateDetector(det2params,sim)

```

## SNR Truncation

Apply SNR truncation to the detectors by removing all detections that were less than the threshold.

```{r}
snrTruncationThreshold <- 0; # dB

# True number of calls above threshold
nAboveThreshold <- sum(sim$snr>snrTruncationThreshold) 

# Variables from sims that were truncated in SNR get a 't' as a suffix
simDet1t <- subset(simDet1, simDet1$snr >= snrTruncationThreshold)
simDet2t <- subset(simDet2, simDet2$snr >= snrTruncationThreshold)

# Number of detections above threshold for each detector
Nc1 <- sum(simDet1t$detect_table)
Nc2 <- sum(simDet2t$detect_table)

```

View the spatial detection density (which is not the same as spatial call density since it  accounts for neither detection probability nor time).

```{r detector.plots, warning=FALSE, message=FALSE}

p1  <- plotSpatialDetections(simDet1)
p2  <- plotSpatialDetections(simDet2)
p1t <- plotSpatialDetections(simDet1t)+ylab("Truncated\nY (kM)")
p2t <- plotSpatialDetections(simDet2t)+ylab("Truncated\nY (kM)")

gridExtra::grid.arrange(p1,p2,p1t,p2t,nrow=2)

dist1 <- plotDetectionDistribution(simDet1)
dist2 <- plotDetectionDistribution(simDet2)
dist1t<- plotDetectionDistribution(simDet1t)+ylab("Truncated count")
dist2t<- plotDetectionDistribution(simDet2t)+ylab("Truncated count")
gridExtra::grid.arrange(dist1,dist2,dist1t,dist2t,nrow=2)

```

## Subsample data for detector characterisation

@castro2024 subsample approximately 200 hours evenly spaced throughout the year to create detector characterisation curves (detection vs SNR). Here we subsample 1 hour in every 41 hours yielding 214 subsamples that are 1 hour in duration.

```{r subsample.for.detector.characterisation}

subsampleDet1 <- subsampleSimInTime(simDet1,interval = "23 hour")
subsampleDet2 <- subsampleSimInTime(simDet2,interval = "23 hour")

subsampleDet1t <- subsampleSimInTime(simDet1t,interval = "23 hour")
subsampleDet2t <- subsampleSimInTime(simDet2t,interval = "23 hour")

print(paste("True positives in original  subsample of detector 1: ", 
            with(subsampleDet1,sum(groundTruth & detect_table)) ) )
print(paste("True positives in truncated subsample of detector 1: ", 
            with(subsampleDet1t,sum(groundTruth & detect_table)) ) )

print(paste("True positives in original  subsample of detector 2: ", 
            with(subsampleDet2,sum(groundTruth & detect_table)) ) )
print(paste("True positives in truncated subsample of detector 2: ", 
            with(subsampleDet2t,sum(groundTruth & detect_table)) ) )

```

## Detection matching between detector 1 and 2

Both detectors (simulated subsets) operate on the same set of calls (i.e. they are different subsets of the same underlying simulation). So we can find detections from one detector that match those of the other detector by merging the detections into a capture history table.

Presently, false positives have randomly generated date-times, so have a negligible chance of matching between the two detectors.

*TODO: Create false positives in a way that better approximates their real-world occurrence (i.e. in a way that allows them to be matched between detectors).*

```{r match detections from detector1 and detector2}

# Create a capture history table from the two tables of detections (All
# simulated calls in the subset whether detected or not)
capHistTab<- simsTocaptureHistoryTable(subsampleDet1,subsampleDet2)
capHistTabt<- simsTocaptureHistoryTable(subsampleDet1t,subsampleDet2t)

# Total number of true positive detections for detector1 OR detector2
nTrueDetectedSubset <- with(capHistTab,sum(detect_table1 & groundTruth1 |   
                                        detect_table2 & groundTruth2) )
nTrueDetectedSubsett <- with(capHistTabt,sum(detect_table1 & groundTruth1 |   
                                        detect_table2 & groundTruth2) )

# Total number of positive detections for detector1 OR detector2
nPositiveSubset <- with(capHistTab,sum(detect_table1 | detect_table2))
nPositiveSubsett <- with(capHistTabt,sum(detect_table1 | detect_table2))

```

Convert this capture history table into an SNRinfo table, and treat detector1 as ground truth.

```{r format.output.for.callDensity.package}

# Convert capture history into SNRinfo format (removing detections )
SNRinfo <- capHistTosnrInfo(capHistTab)
SNRinfot<- capHistTosnrInfo(capHistTabt)

# Detector 2 true & false positives
n_p_subsampleDet2 <- sum(capHistTab$detect_table2)
n_p_subsampleDet2t <- sum(capHistTabt$detect_table2) 

# Detectir 2 false positives (Positive on detector 2, but not on 1)
n_fp_subsample <- sum(capHistTab$detect_table2 & !capHistTab$detect_table1)
n_fp_subsamplet <- sum(capHistTabt$detect_table2 & !capHistTabt$detect_table1)

# False discovery rate is number.false.positives/number.predicted.positive
c_subsample <- n_fp_subsample/n_p_subsampleDet2
c_subsamplet<- n_fp_subsamplet/n_p_subsampleDet2t

n_subsample <- dim(SNRinfo)[1]  # Number positive detections in SUBset
n_subsamplet <- dim(SNRinfot)[1]  # Number positive detections in SUBset

truncationDistances <- R 

# Number positive detections in FULL set
Nc <- sum(simDet2$detect_table)
Nct <- sum(simDet2t$detect_table)       

```

## Calculate call densities
### SNR Truncation ONLY for SNR Detection Function

First we calculate call density using the original subset (without SNR truncation). These results are given the suffix .glm. 

We compare the call density and probability of detection in the area of original subset to those derived using SNR truncatiot but only truncating in SNR to derive the SNR-detection function, and NOT applying SNR truncation to the Monte-Carlo model nor to derivation of false positive rate, c.  

```{r GLM p.det with independently estimated NL, warning=FALSE, message=FALSE}

snrDetFun.glm <- callDensity::fitSNRdetectionFunc(SNRinfo,  modelType = "glm")
snrDetFun.glmt<- callDensity::fitSNRdetectionFunc(SNRinfot, modelType = "glm")

snrDetFun.glm$r.sq <- with(summary(snrDetFun.glm),  1 - deviance/null.deviance)
snrDetFun.glmt$r.sq<- with(summary(snrDetFun.glmt), 1 - deviance/null.deviance)

# Estimate of NL distribution. NOTE: Noise levels estimated from mean NL of
# non-truncated detections plus intercept of SNR detection function (truncated
# or not).
NLind <- nlFromDetections(SNRinfo,  snrDetFun.glm, SL = SL, TL = TL)
NLindt<- nlFromDetections(SNRinfo, snrDetFun.glmt, SL = SL, TL = TL)

Pa.glm <- callDensity::pDetInArea(snrDetFun.glm, SL=SL, TLlookup = TL, NL=NLind,
           output.resolution.m=100, outerloop = 100, truncationDistance=R, # m
           snrTruncationThreshold =-Inf)
pa.glm <- Pa.glm$overall

Pa.glmt <- callDensity::pDetInArea(snrDetFun.glmt, SL=SL, TLlookup = TL, NL=NLindt,
           output.resolution.m=100, outerloop = 100, truncationDistance=R, # m
           snrTruncationThreshold =-Inf)
pa.glmt <- Pa.glmt$overall


# Density estimation equation
Dc.glm <- (Nc * (1 - c_subsample) )/( k * A * pa.glm * Time )
Dc.glmt <- (Nc * (1 - c_subsample) )/( k * A * pa.glmt * Time )


results.glm <- data.frame(Dc=Dc.glm, SampleSize=n_subsample, A = A,
                      NLmean = NLind$mean, NLsd = NLind$sd, 
                      model='glm', c=c_subsample, Pa=pa.glm,  
                      row.names = 'All SNR')
results.glmt <- data.frame(Dc=Dc.glmt, SampleSize=n_subsamplet, A = A,# Environment
                      NLmean = NLind$mean, NLsd = NLind$sd,      # Noise
                      model='glm', c=c_subsample, Pa=pa.glmt,  
                      row.names = paste('Truncated',snrTruncationThreshold,'dB'))

```


Compare with the true results

```{r Call.Density.Results, echo=TRUE, warning=FALSE}

resultsTrue = data.frame(Dc = TrueCallDensity, SampleSize = n, A = A,# Environment 
                      NLmean = NL$mean, NLsd = NL$sd,            # Noise
                        # Detector params
                        model=det2params$func, c = det2params$c, 
                        Pa = mean(simDet2$p_det,na.rm=TRUE), 
                      row.names='Truth')

# True density and PA of calls above threshold
Dct <- nAboveThreshold/(A*Time) 
Pat <- mean(simDet2$p_det[simDet2$snr > snrTruncationThreshold ],na.rm=T)

results<-rbind( resultsTrue, results.glm, results.glmt)

kableExtra::kbl(results, digits = c(4,3,0,1, 2, NA, 3, 4), 
                col.names= c('Dc','(sub)Sample Size', 'A','NL_mean','NL_sd',
                             'Model','c','P_a')) %>% 
  kableExtra::kable_classic(full_width=FALSE)


```

In this example, the estimate of call density with SNR truncation is further from the true value than the estimate without. 

## SNR Truncation everywhere (snrDetFun, pDetInArea, c, Nc)

Alternatively, we can apply SNR truncation for every parameter that goes into the call density equation. 

This means applying SNR truncation when tabulating the number of calls, Nc; when estimating the density of calls within the area, pa; as well as applying SNR truncation when estimating the false discovery rate, c. 

Applying SNR truncation in this way returns the density of **all** calls, not the density of above-threshold calls, provided the truncation is applied consistently to every term that counts detections. This is not obvious, and it is worth being explicit about why.

Let $q(\theta)$ be the fraction of calls that arrive at the sensor with an SNR above the threshold $\theta$. Truncation shrinks $N_c$ by a factor of roughly $q(\theta)$, because we are counting fewer detections. It shrinks $p_a$ by the same factor, because `pDetInArea` sets the probability of detection to zero below $\theta$ rather than discarding those parts of the area. The two factors appear on opposite sides of the density equation, so they cancel, and $D_c$ is unchanged.

The cancellation only happens if $p_a$ is the *unconditional* mean $E[p \cdot 1(SNR \ge \theta)]$ rather than the conditional mean $E[p \mid SNR \ge \theta]$. That is the difference between setting below-threshold cells to zero and dropping them. Zeroing keeps them in the average and they contribute nothing; dropping them renormalises the average over the survivors, which would give the density of above-threshold calls instead. `pDetInArea` zeroes them.

Distance truncation is not treated the same way, and must not be. Beyond the truncation distance $w$ the study area is redefined as $A = \pi w^2$, so that area genuinely leaves the question and those cells are dropped. Calls below $\theta$ have not left the question. They are still in the water and still in $A$. We have only declared that we cannot hear them.

```{r}

# True value of false discovery rate for truncated sim
c_truncated_actual = sum(simDet2t$groundTruth==FALSE)/sum(simDet2t$detect_table)

# The truncated estimate targets the density of ALL calls, so that is the
# truth to compare it against. Dct, the density of above-threshold calls, is
# reported alongside only to show the size of q(theta).
#
# Pa here is the UNCONDITIONAL mean of p over every call, with
# below-threshold calls contributing zero. This is what a zeroed pDetInArea
# targets. Pat, the conditional mean over surviving calls, is the quantity
# the old NA behaviour returned, and is much larger.
PaTruncUnconditional <- mean(simDet2$p_det * (simDet2$snr >= snrTruncationThreshold),
                             na.rm = TRUE)
qTheta <- nAboveThreshold / n

resultsTruncated = data.frame(Dc = TrueCallDensity, SampleSize=nAboveThreshold,
                      A = A, NLmean = NL$mean, NLsd = NL$sd,            # Noise
                      model=det2params$func, 
                      c = c_truncated_actual, Pa = PaTruncUnconditional, 
                      row.names='Truth (all calls)' ) 

# Probability of detection in teh area of calls that are above threshold (NOT all calls)
Pa.allt <- callDensity::pDetInArea(snrDetFun.glmt, SL=SL, TLlookup = TL, NL=NLindt,
           output.resolution.m=100, outerloop = 100, truncationDistance=R, # m
           snrTruncationThreshold =snrTruncationThreshold)
pa.allt <- Pa.allt$overall
Dc.allt <- (Nct * (1 - c_subsamplet) )/( k * A * pa.allt * Time )

results.allt <- data.frame(Dc=Dc.allt, SampleSize=n_subsamplet, A = A,# Environment
                      NLmean = NLindt$mean, NLsd = NLindt$sd,      # Noise
                      model='glm', c=c_subsamplet, Pa=pa.allt,  
                      row.names = paste('Truncated',snrTruncationThreshold,'dB'))

resultst<-rbind( resultsTruncated, results.allt)

kableExtra::kbl(resultst, digits = c(4,3,0,1, 2, NA,  3, 4), 
                col.names= c('Dc','(sub)Sample Size', 'A','NL_mean','NL_sd',
                             'Model','c','P_a')) %>% 
  kableExtra::kable_classic(full_width=FALSE)

```

Applying truncation throughout all the terms of the call density equation produces a correct estimate of the density of all calls. The estimate is not of a different quantity from the untruncated one; it is the same quantity, reached by discarding the part of the detection function that the data could never support.

The earlier concern that these densities would only be comparable across sites using identical thresholds does not apply once $q(\theta)$ cancels. Two sites truncated at different thresholds both return all-call density, and are directly comparable. What they do not share is precision: the site with the higher threshold has a smaller $N_c$.

What truncation does do is move an assumption from one place to another. Without truncation, $p_a$ depends on the shape of the detection function at SNRs where no observer ever contributed a detection. That shape is not measurable, even in principle. With truncation, $p_a$ depends instead on $q(\theta)$, which comes from SL, TL and NL. That is the sonar equation, which is already the backbone of the method, and which can be checked against independent measurements. Truncation creates no information about faint calls. It relocates the assumption onto ground that can be inspected.

The cost is precision, and it is not small. $N_c$ shrinks with $q(\theta)$, and the truncated estimate is more sensitive to error in SL, NL and TL than the untruncated one, because $q(\theta)$ now carries weight that the detection function used to carry. Sensitivity to acoustic parameter mismatch is left as an open question and remains on the TODO list.

An earlier draft of this vignette proposed tracking the proportion of calls excluded by truncation, and scaling the above-threshold density back up to recover all-call density. That proportion is $q(\theta)$, and zeroing obtains it for free. No bookkeeping is required, and none was added.

## VGLM and truncation

Placeholder to demonstrate that adjudicated capture recapture produces same results when truncating in SNR. 

```{r VGLM p.det, warning=FALSE, message=FALSE}
library(VGAM)
### Adjudicated capture recapture (capture history table with false positives
### removed by judge)
adjudicated <- subset(capHistTab,
                (capHistTab$groundTruth1 | capHistTab$groundTruth2) &
                (capHistTab$detect_table1 | capHistTab$detect_table2))
adjudicatedt <- subset(capHistTabt,
                (capHistTabt$groundTruth1 | capHistTabt$groundTruth2) &
                (capHistTabt$detect_table1 | capHistTabt$detect_table2))

# Number of adjudicated positive detections
n_adj <- dim(adjudicatedt)[1]

adjudicatedt$SNR <-   rowMeans(
  subset(adjudicatedt,select=c('snr1','snr2')) ,na.rm=T) #dB

#False discovery rate of truncated
tp <- sum(with(capHistTabt,(detect_table2 & groundTruth2) ))
fp <- sum(with(capHistTabt,(detect_table2 & !groundTruth2) ))
# precision <- sum(tp)/length(tp)
adj_c <- fp/(fp+tp)

observerNames = c("detect_table1", "detect_table2")
snrDetFun.vglm <- fitSNRvglm(adjudicatedt, observerNames,
                             whichObserver = "detect_table2")

# Estimate of NL distribution
noiseTab <- capHistTab[capHistTab$detect_table2 | capHistTab$detect_table1,]
noiseTab$NoiseRL <- rowMeans(
  subset(noiseTab,select=c('noiseRMSdB1','noiseRMSdB2')) ,na.rm=T) #dB
NLadj<- nlFromSnrInfo(noiseTab, snrDetFun.vglm)

Pa.vglm <- callDensity::pDetInArea(snrDetFun.vglm, SL=SL, TLlookup = TL, NL=NLadj,
           output.resolution.m=100,
           outerloop = 10,
           truncationDistance=R, # in m
           snrTruncationThreshold = snrTruncationThreshold)
pa.vglm <- Pa.vglm$overall


# Density estimation equation
Dc.vglm <- (Nct * (1 - adj_c) )/( k * A * pa.vglm * Time )

results.vglm <- data.frame(Dc=Dc.vglm, SampleSize=n_adj, A = A,# Environment
                      NLmean = NLadj$mean, NLsd = NLadj$sd,# Noise
                      model='vglm',  c=adj_c, Pa=pa.vglm,  
                      row.names = 'ACR')

```


```{r}

resultst<-rbind( resultsTruncated, results.allt, results.vglm)

kableExtra::kbl(resultst, digits = c(4,3,0,1, 2, NA,  3, 4), 
                col.names= c('Dc','(sub)Sample Size', 'A','NL_mean','NL_sd',
                             'Model','c','P_a')) %>% 
  kableExtra::kable_classic(full_width=FALSE)


```
The adjudicated capture-recapture model seems to produce a correct estimate of density of above-threshold calls. But something might not be quite right since the values for c seem a bit off. Further investigation of the code is warranted. 


## Choosing the threshold

The threshold is not a free parameter to be tuned until the answer looks right. It has a meaning, and the meaning tells you where to put it.

A capture-recapture model identifies detection probability from disagreement between observers. It needs both observers to contribute detections across the SNR range that matters. Where only one observer ever contributes, there are no recaptures, and the model is not estimating anything there. It is extrapolating.

The temptation is to set the threshold where the *sample* runs out. That is the wrong diagnostic. A shallow partner can reach a long way down on its own and fill the sample with low-SNR detections, none of which are recaptures. The sample looks well covered and the model is still extrapolating.

The requirement is a property of the partner's **slope**, not its recall. A partner with a steep detection function behaves like a step function: above its threshold it hears everything, below it nothing, and the faint region has no recaptures at any recall. A mediocre partner that degrades gradually is more useful for capture-recapture than a sharp one, however good the sharp one is. An SNR-insensitive CNN is close to an ideal partner. A hard energy threshold is close to a worst case.

So set $\theta$ where the recaptures run out. That is measurable directly from the capture history table, before any call density is estimated:

```{r recapture.coverage, warning=FALSE, message=FALSE}

# Recapture coverage: the SNR range over which BOTH observers contribute
# detections. Not where the sample reaches; where the agreement reaches.
recaptureCoverage <- function(ch, breaks = seq(-20, 20, by = 2),
                              obs1 = "detect_table1", obs2 = "detect_table2",
                              snrCol = "SNR") {
  # Only calls somebody flagged are in the capture-recapture sample.
  ch  <- ch[ch[[obs1]] | ch[[obs2]], ]
  bin <- cut(ch[[snrCol]], breaks = breaks)
  mid <- head(breaks, -1) + diff(breaks) / 2

  data.frame(
    snr        = mid,
    det1only   = as.vector(tapply(ch[[obs1]] & !ch[[obs2]], bin, sum, default = 0)),
    det2only   = as.vector(tapply(!ch[[obs1]] & ch[[obs2]], bin, sum, default = 0)),
    recaptures = as.vector(tapply(ch[[obs1]] & ch[[obs2]], bin, sum, default = 0))
  )
}

coverage <- recaptureCoverage(capHistTab)
print(subset(coverage, det1only + det2only + recaptures > 0), row.names = FALSE)

```

Read down the `recaptures` column. Where it reaches zero while `det2only` is still large, detector 1 has stopped contributing, and the model has nothing to work with below that point. That is where $\theta$ belongs.

For the steep partner used below, recaptures are zero in every bin below about $-1$ dB, while detector 2 alone contributes hundreds of detections per bin. The sample is well populated down to $-15$ dB and none of it is informative. Recaptures only become substantial at $+1$ dB.

That is worth dwelling on, because it is the argument for the diagnostic. The threshold it points at, near 0 dB, is the same threshold the sweep in the next section shows to remove the bias. The diagnostic reaches it from the capture history table alone, without knowing the true call density and without fitting anything. In a real analysis the truth is not available and the sweep cannot be run, so this is the only version of the question that can actually be asked.

### Does it work?

The failure that motivates truncation is a steep partner. With detector 1 at `scale = 0.5` and detector 2 at `scale = 4`, capture-recapture overestimates $p_a$ by about 60% and therefore underestimates $D_c$ by the reciprocal. Truncating removes the extrapolated region and the bias goes away, while $D_c$ remains the density of all calls:

| partner | $\theta$ | $N_c$ | $p_a$ / true $p_a$ | $D_c$ / true $D_c$ |
|---|---|---|---|---|
| steep (`scale = 0.5`) | none | 6055 | 1.62 | **0.62** |
| steep | 0 dB | 2735 | 1.02 | 0.99 |
| steep | 3 dB | 1741 | 1.00 | 1.01 |
| steep | 6 dB | 994 | 1.02 | 0.99 |
| shallow (`scale = 8`) | none | 6051 | 0.99 | 1.01 |
| shallow | 0 dB | 2699 | 1.00 | 1.00 |
| shallow | 3 dB | 1735 | 1.01 | 1.00 |
| shallow | 6 dB | 1001 | 1.03 | 0.99 |

Mean of three seeds per cell, 50,000 calls, false discovery rate zero, noise supplied. Produced by `notes/truncationSweep.R`.

Three things in that table are worth stating plainly.

$D_c$ stays at the true all-call density while $N_c$ falls six-fold from 6055 to 994. That is the $q(\theta)$ cancellation working. If $p_a$ were renormalised instead of zeroed, $D_c$ would fall with $N_c$.

Truncation does not damage the case that already worked. The shallow partner is unbiased with or without it, so the threshold costs precision and buys nothing there.

The threshold does not need to be tuned. Anywhere from 0 to 6 dB fixes the steep partner. The bias is caused by a region of the curve that carries no data, and any threshold that excludes that region works. This matters in practice, because it means $\theta$ can be chosen from the recapture diagnostic above without reference to the answer it produces.

One practical note. Fitting the untruncated VGLM in the steep-partner case emits convergence warnings from `posbernoulli.t` about working weights and half-step sizes. Those warnings are the symptom, not a separate problem: the model is being asked to fit a region where it has no information. They disappear under truncation.

## Inspect detection functions for insights

```{r compare snr detection functions, warning=FALSE}
SNR <- seq(-20,20,0.5);
snrs <- data.frame(SNR)

cols = colorRampPalette(c('#EDB120','#D95319','#0072BD','#7E2F8E','#77AC30',
                        '#4DBEEE','#A2142F'))
preds.glm= predict(snrDetFun.glm, type='response',newdata=snrs)
preds.glmt= predict(snrDetFun.glmt, type='response',newdata=snrs)
preds.vglm=predict(snrDetFun.vglm,type='response',newdata=snrs)
preds.vglm0=predict(snrDetFun.vglm,type='response',newdata=snrs,
                    type.fitted='onempall0')# One minus probability of all zeros
pr =apply(cbind(preds.vglm[,2],preds.vglm0),1,prod)


par(cex=0.8,mar=c(3.1,3.1,0.25,0),mgp=c(2.1,1,0))
plot(snrs$SNR, xlim= c(-5,10),  ylim= c(0,1),
     xlab="SNR", ylab = "P(Detection)", type='n')
grid(nx = NULL, ny = NULL, lty = 2, col = "gray", lwd = 2)
lines(snrs$SNR,preds.glm,col=cols(7)[4],lwd=2)
lines(snrs$SNR,preds.glmt,col=cols(7)[5],lwd=2)
lines(snrs$SNR,pr,col=cols(7)[1],lwd=2)    # vglm 1-p(all zero)
points(simDet2t$snr,simDet2t$p_det,pch='.')     # Actual probability of detection
legend( "bottomright", 
        legend =c("TRUE",
                  "GLM-No SNR Truncation",
                  "GLM-SNR Truncated",
                  "VGLM-SNR Truncated"
                  # "Perception bias: detector1 (VGLM)",
                  # "Perception bias: detector2 (VGLM)",
                  # "1-p(all zeros) (VGLM)"
                  ),
        col=c('black', cols(7)[4], cols(7)[5], 
              cols(7)[1] #, cols(7)[6],  cols(7)[2]), 
        ),
        lty=rep('solid',6),
        lwd=rep(2,6)
)
```


# References
