---
title: "callDensity: Estimating noise levels from detections"
output: 
  html_document: default
  rmarkdown::html_vignette:
    vignette: >
      %\VignetteEncoding{UTF-8}
      %\VignetteIndexEntry{callDensity: Estimating noise levels from detections}
      %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, fig.width = 6, fig.height = 3.5,
                      comment = "#>")
```

```{r setup}
library(callDensity)
library(ggplot2)
library(kableExtra)
```

# The problem in one paragraph

Call density needs a distribution of noise levels. The obvious place to get one
is the noise measured at each detection, since those measurements are already
in the capture history table. Those measurements are biased low. A quiet period
gives you a high SNR, a high SNR gives you a detection, so your detections
oversample the quiet periods. Measure noise only where you detected something
and you will conclude the ocean is quieter than it is.

This vignette shows how big that bias is, where it comes from, and how to undo
it.

# The one idea

`pDetInArea` already answers a question very close to the one we need. It asks:
given a distribution of noise levels, what fraction of the calls in the study
area get detected? That fraction is $\hat{p}_a$. It draws noise levels from a
normal distribution and averages over them.

Ask a slightly different question. **What would $\hat{p}_a$ be if the noise were
always exactly 84 dB?** Not drawn from a distribution. Fixed.

Same machinery, one line different. Call the answer `pDetGivenNL(84)`. Compute
it for 70 dB, 71 dB, 72 dB and so on, and you have a curve: probability of
detection against noise level.

That curve is the whole solution. It is not a new concept. It is $\hat{p}_a$
with the noise held still.

# Building the curve

`pDetGivenNL` is not in the package yet. Here it is as local code so that this
section runs today.

```{r pdetgivennl}

#' Probability of detecting a call somewhere in the study area, given a fixed
#' noise level.
#'
#' @param nl      Vector of noise levels (dB) to evaluate.
#' @param detFun  Either a detFun object, or a plain function of SNR returning
#'                probability.
#' @param TL      Transmission loss table. First column is range in m, the
#'                remaining columns are TL in dB for each radial transect.
#' @param SL      data.frame with mean and sd of the source level distribution.
#' @param truncationDistance Scalar, or one value per transect.
#' @param nSLnodes Number of quadrature nodes for the source level integral.
#' @return Numeric vector, same length as nl.
pDetGivenNL <- function(nl, detFun, TL, SL,
                        truncationDistance = max(TL[, 1]),
                        nSLnodes = 41) {

  range_m <- TL[, 1]
  tlMat   <- as.matrix(TL[, -1, drop = FALSE])

  # Area weights. Each range step stands for an annulus, and the area of an
  # annulus grows with r, so a step at 200 km carries ten times the weight of
  # a step at 20 km. Cells beyond the truncation distance carry no weight.
  truncationDistance <- rep_len(truncationDistance, ncol(tlMat))
  wMat <- outer(range_m, truncationDistance, function(r, td) r * (r <= td))
  totalWeight <- sum(wMat)

  # Source levels. Calls are not all equally loud, so average over the SL
  # distribution using a weighted grid.
  slNodes <- seq(SL$mean - 4 * SL$sd, SL$mean + 4 * SL$sd, length.out = nSLnodes)
  slWt    <- dnorm(slNodes, SL$mean, SL$sd)
  slWt    <- slWt / sum(slWt)

  # The detection function depends on one number, SNR. So evaluate it once on a
  # grid and interpolate, rather than calling predict() millions of times.
  pLookup <- if (is.function(detFun)) {
    detFun
  } else {
    snrGrid <- seq(-100, 100, by = 0.1)
    pGrid   <- predictDetFun(detFun, newdata = data.frame(SNR = snrGrid))$fit
    approxfun(snrGrid, pGrid, rule = 2)
  }

  vapply(nl, function(thisNL) {
    acc <- 0
    for (i in seq_along(slNodes)) {
      snr <- slNodes[i] - tlMat - thisNL      # sonar equation, every cell
      acc <- acc + slWt[i] * sum(pLookup(snr) * wMat)
    }
    acc / totalWeight
  }, numeric(1))
}

```

Now build a curve. The transmission loss table and source levels follow the
second vignette, and the detector is a plain logistic so this section does not
depend on anything else.

```{r build.curve}

R  <- 1e6
SL <- data.frame(mean = 190, sd = 4, sampleSize = 350)

TL <- simTLradials_20logR(maxRange = R, rangeStep = 1000, numTransects = 4)
TL <- TL[is.finite(rowSums(TL)) & TL[, 1] > 0, ]     # drop r = 0 if present

detector <- function(snr) plogis(snr, location = 1, scale = 2)

nlGrid <- seq(from = 55, to = 135, by = 1)
pCurve <- pDetGivenNL(nlGrid, detector, TL, SL)

curveDf <- data.frame(nl = nlGrid, pDet = pCurve)

```

## The graph that explains everything

Before plotting it, be clear about what is on the axes, because there is no
data on this plot. Every point is a counterfactual. It asks: *if the noise were
always exactly this loud, and never varied, what fraction of the study area
would you be hearing calls in?*

That fraction is just detection area divided by study area. So this plot is
really about detection range, which is the quantity you think in anyway. Add
the range at a few points and it becomes readable.

```{r curve.plot, warning=FALSE, fig.height=4}

# Range at which a call of mean source level sits at the detector's 50% point.
# Approximate. Ignores the spread in SL and the smoothness of the detector, so
# it is a label, not a calculation the curve depends on.
detRadius_km <- function(nl, SLmean = 190, loc = 1) {
  10^((SLmean - nl - loc) / 20) / 1000
}

marks <- data.frame(nl = c(74, 84, 94, 104))
marks$pDet  <- pDetGivenNL(marks$nl, detector, TL, SL)
marks$label <- sprintf("%.0f km", detRadius_km(marks$nl))

# A reference line falling exactly one decade per 10 dB, anchored at 90 dB.
anchor <- pDetGivenNL(90, detector, TL, SL)
refDf  <- data.frame(nl = nlGrid, pDet = anchor * 10^(-(nlGrid - 90) / 10))

ggplot(curveDf, aes(x = nl, y = pDet)) +
  geom_line(data = refDf, linetype = "dashed", colour = "grey50",
            linewidth = 0.6) +
  geom_line(linewidth = 0.9, colour = "steelblue") +
  geom_point(data = marks, size = 2, colour = "grey20") +
  geom_text(data = marks, aes(label = label), hjust = -0.25, vjust = -0.4,
            size = 3, colour = "grey20") +
  geom_vline(xintercept = 84, colour = "firebrick", linetype = "dotted") +
  annotate("text", x = 84.5, y = 1e-4, label = "true mean NL",
           colour = "firebrick", hjust = 0, size = 3) +
  scale_y_log10() +
  xlab("Noise level (dB)") +
  ylab("Fraction of the study area you can hear into") +
  labs(caption = "Labels: detection radius. Dashed grey: one decade per 10 dB")

```

Work through two of those labelled points by hand and the plot stops being
abstract.

**At 84 dB.** A 190 dB call reaches the detector's 50% point when SNR is 1 dB,
so when TL is $190 - 84 - 1 = 105$ dB. Under spherical spreading that is
$10^{105/20}$ metres, or 178 km. So you are hearing calls out to roughly 178 km.
The area you can hear into is $\pi \times 178^2 \approx 99{,}500$ km². The study
area is $\pi \times 1000^2 \approx 3{,}140{,}000$ km². You are hearing into
about 3% of it.

**At 94 dB.** Every call now needs 10 dB more received level. Ten decibels of
spherical spreading is a factor of 3.16 in range. So 178 km becomes 56 km. The
area is $\pi \times 56^2 \approx 9{,}950$ km², one tenth of what it was. You are
hearing into 0.3% of the study area.

Ten decibels louder, ten times less ocean, ten times fewer calls. Each 10 dB
step to the right divides the curve by 10. On a log axis that is a straight
line, and that is all the dashed reference line is saying.

**The detector never entered that.** Move its threshold and the 178 km becomes
200 km or 150 km. The factor of ten per ten decibels does not budge, because it
came from the geometry of a circle, not from the detector. That is the single
most important thing on this plot, and it is why correcting the noise using a
property of the detector cannot be right.

**The shoulder on the left.** Below about 69 dB the labelled radius would exceed
1000 km. There is no ocean out there. It is past the edge of the study area, so
it is not yours to hear into. The curve flattens at 1 and making the sea
quieter buys nothing more. That shoulder is exactly where the tidy algebra
stops working, and it is why this vignette computes a curve rather than fitting
a slope to it.

**Where your data sits.** The red line marks the true mean noise level, out on
the straight portion, where a decibel of noise really does cost you ocean.

If this plot still feels abstract, skip ahead. The next figure is the same
curve on a linear axis, zoomed into the range your noise actually occupies, and
it is doing the real work.

# Why your measurements are biased

Now use the curve. Nature draws a noise level for each call from a normal
distribution. The detection process then throws most of them away, and it
throws away loud ones far more eagerly than quiet ones. The curve says exactly
how much more eagerly.

```{r filter.plot, warning=FALSE, fig.height=4}

nlTrue  <- 84
nlSd    <- 4

fine  <- seq(nlTrue - 5 * nlSd, nlTrue + 5 * nlSd, length.out = 400)
fN    <- dnorm(fine, nlTrue, nlSd)
pd    <- pDetGivenNL(fine, detector, TL, SL)
wt    <- fN * pd

# Scale each curve to a common height so the shift is visible
filterDf <- rbind(
  data.frame(nl = fine, y = fN / max(fN),  what = "1. All calls (the truth)"),
  data.frame(nl = fine, y = pd / max(pd),  what = "2. Chance of detection"),
  data.frame(nl = fine, y = wt / max(wt),  what = "3. What you measure")
)

meanObs <- sum(fine * wt) / sum(wt)

ggplot(filterDf, aes(x = nl, y = y, colour = what)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = nlTrue,  colour = "grey30", linetype = "dotted") +
  geom_vline(xintercept = meanObs, colour = "grey30", linetype = "dashed") +
  annotate("segment", x = meanObs, xend = nlTrue, y = 1.05, yend = 1.05,
           arrow = arrow(length = unit(2, "mm"), ends = "both")) +
  annotate("text", x = (meanObs + nlTrue) / 2, y = 1.12,
           label = sprintf("bias = %.2f dB", nlTrue - meanObs), size = 3) +
  xlab("Noise level (dB)") +
  ylab("Scaled to a common height") +
  labs(colour = NULL) +
  theme(legend.position = "top")

```

Curve 1 is the truth. Curve 2 is the detection curve from the previous plot,
now drawn on a linear axis over the range the noise actually occupies. Curve 3
is curve 1 multiplied by curve 2, and it is what your detections show you.

Multiplying by a curve that leans left drags the peak left. That is the entire
bias. Nothing subtle is happening.

Two things worth noticing, because both get used later.

Curve 3 is narrower than curve 1 by very little. The width survives the
filtering almost intact, which means the standard deviation of your measured
noise is close to the truth even though the mean is not. That leaves only one
unknown to solve for.

The size of the shift depends on how steeply curve 2 leans, and on how wide
curve 1 is. It does not depend on where the detector sits. Move the detector
threshold and curve 2 changes height, not slope.

# Undoing it

The previous plot is a prediction machine. Give it a candidate true mean, and
it tells you what mean you should have measured.

You know what you measured. So run it backwards.

```{r predict.function}

#' Predict the mean noise level you would measure at detections, given a
#' candidate true noise distribution.
predictSampledNL <- function(mu, sigma, detFun, TL, SL,
                             truncationDistance = max(TL[, 1])) {
  nl <- seq(mu - 6 * sigma, mu + 6 * sigma, length.out = 121)
  wt <- dnorm(nl, mu, sigma) *
        pDetGivenNL(nl, detFun, TL, SL, truncationDistance)
  sum(nl * wt) / sum(wt)
}

```

```{r rootfind.plot, warning=FALSE, fig.height=3.5}

muTry <- seq(80, 92, by = 1)
muObs <- vapply(muTry, function(m)
  predictSampledNL(m, nlSd, detector, TL, SL), numeric(1))

rootDf <- data.frame(muTrue = muTry, muMeasured = muObs)

ggplot(rootDf, aes(x = muTrue, y = muMeasured)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey70", linetype = "dashed") +
  geom_line(linewidth = 0.9, colour = "steelblue") +
  geom_hline(yintercept = meanObs, colour = "firebrick", linetype = "dotted") +
  geom_vline(xintercept = nlTrue,  colour = "firebrick", linetype = "dotted") +
  annotate("text", x = 80.2, y = meanObs + 0.3, hjust = 0, size = 3,
           colour = "firebrick", label = "what you measured") +
  xlab("Candidate true mean NL (dB)") +
  ylab("Mean NL you would measure at detections") +
  labs(caption = "Dashed grey: the 1:1 line you would get with no bias")

```

The blue line sits below the 1:1 line by the size of the bias. It rises
steadily, so there is exactly one candidate that reproduces your measurement.
Find it by reading across from the red horizontal line to the blue curve, then
down. That is what `uniroot` does.

```{r nlfromdetections}

#' Estimate the true noise level distribution from noise measured at detections.
#'
#' PLACEHOLDER. This will become nlFromDetections() in the package, replacing
#' nlFromSnrInfo(). See "What is left to build" below.
nlFromDetections <- function(snrInfo, detFun, TL, SL,
                             truncationDistance = max(TL[, 1]),
                             searchWidth = 25) {

  obsMean <- mean(snrInfo$NoiseRL, na.rm = TRUE)
  sigma   <- sd(snrInfo$NoiseRL,   na.rm = TRUE)

  # The bias is always positive, so the true mean is above the measured one.
  gap <- function(mu) {
    predictSampledNL(mu, sigma, detFun, TL, SL, truncationDistance) - obsMean
  }
  mu <- uniroot(gap, interval = c(obsMean, obsMean + searchWidth))$root

  data.frame(mean = mu, sd = sigma,
             sampleSize = sum(!is.na(snrInfo$NoiseRL)))
}

```

# Where this leaves the old function

`nlFromSnrInfo` corrects the same bias by adding the SNR at which the detection
function reaches 0.5. That quantity is a property of the detector. The plots
above say the bias is a property of the geometry and the noise variance. The
two happen to be close for the parameters of the second vignette and they part
company elsewhere.

```{r sweep.placeholder, eval=FALSE}

# PLACEHOLDER - see callDensity3 scratchpad, to be folded in here.
#
# Two sweeps, both at Casey2019 scale (214 h subsample, every 8th detection
# adjudicated):
#
#   A. Move the detector threshold from -5 to +10 dB SNR. The bias needed
#      stays near 3 dB throughout while the old correction tracks the
#      threshold exactly.
#
#   B. Vary sd(NL) from 1 to 10 dB. The bias grows with the square of sd
#      while the old correction does not move.
#
# Add nlFromDetections as a third column to both. Expect it to sit on the
# bias-needed line in both sweeps.

```

# Does it depend on the transmission loss model?

No, and this is the point of computing a curve rather than fitting a slope.
`pDetGivenNL` reads transmission loss out of a lookup table. It never writes
down $20\log_{10}r$. The dashed reference line on the first plot is a
diagnostic, not an assumption.

Spherical spreading is not a serious model for a baleen whale call at 25 Hz in
2000 m of water at 62°S. So use a real one. The file below is pyRAM output
computed on GLORYS12 sound speed profiles along 24 radial transects out to
1000 km, for the South Kerguelen Plateau mooring in summer, at 25 Hz.

## First, look at the file

The layout has to be checked rather than assumed. `pDetGivenNL` needs range in
metres in the first column and one column of TL in dB per transect.

```{r pyram.inspect}

tlFile <- file.path("s:/manuscripts/2026-CopernicusPyRAM/data/output",
                    "TL_Kerguelen2015_combined_25hz_GLORYS12_summer_25Hz.csv")

tlRaw <- read.csv(tlFile)

dim(tlRaw)
names(tlRaw)
tlRaw[1:3, 1:min(5, ncol(tlRaw))]
range(tlRaw[[1]])

```

Expect 101 rows and 25 columns, since the pipeline runs 24 transects at 15°
spacing with 101 range points. If the first column runs to 1e6 it is metres and
nothing needs doing. If it runs to 1000 it is kilometres and needs multiplying.
If the first column is not range at all, stop here and tell me what it is.

```{r pyram.prepare}

TLpyram <- tlRaw

# Range to metres if the file is in km.
if (max(TLpyram[[1]]) < 1e4) TLpyram[[1]] <- TLpyram[[1]] * 1000

# pyRAM returns TL as a positive loss in dB. Drop any non-finite cells and the
# r = 0 row, where TL is undefined.
TLpyram <- TLpyram[is.finite(rowSums(TLpyram)) & TLpyram[[1]] > 0, ]

# 101 points over 1000 km is a 10 km range step, which is coarse next to the
# 1 km step used for the spherical curve above. That matters most at short
# range, where the area weight is small, so it should be a minor effect here.
# Worth checking by halving the step and seeing if the curve moves.
sprintf("%d transects, range step %.1f km, TL at 1000 km: %.1f to %.1f dB",
        ncol(TLpyram) - 1,
        diff(TLpyram[[1]])[1] / 1000,
        min(as.numeric(TLpyram[nrow(TLpyram), -1])),
        max(as.numeric(TLpyram[nrow(TLpyram), -1])))

```

## The same curve, real physics

```{r pyram.curve, warning=FALSE, fig.height=4}

pPyram <- pDetGivenNL(nlGrid, detector, TLpyram, SL)

bothDf <- rbind(
  data.frame(nl = nlGrid, pDet = pCurve, what = "Spherical spreading"),
  data.frame(nl = nlGrid, pDet = pPyram, what = "pyRAM, SKP summer, 25 Hz")
)

anchorP <- pDetGivenNL(90, detector, TLpyram, SL)
refBoth <- data.frame(nl = nlGrid,
                      pDet = anchorP * 10^(-(nlGrid - 90) / 10))

ggplot(bothDf, aes(x = nl, y = pDet, colour = what)) +
  geom_line(data = refBoth, aes(x = nl, y = pDet), inherit.aes = FALSE,
            linetype = "dashed", colour = "grey50", linewidth = 0.6) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = 84, colour = "firebrick", linetype = "dotted") +
  scale_y_log10() +
  xlab("Noise level (dB)") +
  ylab("Fraction of the study area you can hear into") +
  labs(colour = NULL,
       caption = "Dashed grey: one decade per 10 dB, anchored on the pyRAM curve") +
  theme(legend.position = "top")

```

What to look for, and what it means.

**The slope should not be one decade per 10 dB.** At 25 Hz in 2000 m of water
the sound is in a waveguide, not a free field. Beyond a few water depths it
spreads more like a cylinder than a sphere, which makes TL climb more slowly
with range. Slower TL means a decibel of extra noise costs you a wider annulus,
so the curve should fall *faster* than the dashed line, not slower.

If it does, that is the whole argument in one picture. The number
$\lambda = \ln(10)/10$ from the spherical derivation is simply wrong here, and
anything built on it inherits that error.

**The curve should be bumpy.** Convergence zones and bottom interaction add and
remove audible ocean at particular ranges. Wiggles are physics, not noise.

**The shoulder should be further left, or gone.** Flatter TL means longer
detection range, so it takes a quieter sea to run out of study area.

## Does the estimator still work?

The point is that none of the above changes anything. `pDetGivenNL` used the
lookup table. It did not fit a slope to it, it did not assume one, and it never
noticed that the physics changed.

```{r pyram.check, warning=FALSE}

# What a set of detections at this site would show you, and what the estimator
# recovers from that.
truthMean <- 84
truthSd   <- 4

measured <- predictSampledNL(truthMean, truthSd, detector, TLpyram, SL)

recovered <- uniroot(
  function(mu) predictSampledNL(mu, truthSd, detector, TLpyram, SL) - measured,
  interval = c(measured, measured + 25))$root

data.frame(
  site      = "Kerguelen2015 summer 25 Hz",
  truth     = truthMean,
  measured  = round(measured, 3),
  bias      = round(truthMean - measured, 3),
  recovered = round(recovered, 3),
  error     = round(recovered - truthMean, 4)
)

```

`recovered` should land on 84 to several decimals, because this is the same
integral run forwards and then backwards. That is a check that `uniroot` is
bracketing and converging, not evidence that the physics is right.

The interesting column is `bias`. Compare it to the 3 dB or so that spherical
spreading gives. If the pyRAM curve is steeper, the bias here is larger, and
`nlFromSnrInfo` would be further wrong at the real site than in the simulation
it was tuned on.

```{r pyram.seasons, eval=FALSE}

# WORTH DOING, NOT DONE. The bias is a property of the propagation, so it
# should vary by season as the sound speed profile changes, and by site. Loop
# the four seasons and the seven AWR sites and tabulate the bias. If it moves
# by more than a decibel between seasons, that is an argument for estimating NL
# seasonally, which is what Castro et al. (2024) recommend for f anyway.

```

# What is left to build

Everything above runs, but three of the pieces live in this file rather than in
the package.

`pDetGivenNL` duplicates work that `pDetInArea` already does. Both thin the
transmission loss table, broadcast the truncation distance across transects,
mask cells beyond truncation, and form area weights. That preparation should be
pulled out of `pDetInArea` into a shared helper and used by both. Doing it any
other way means the two drift apart.

`nlFromDetections` needs to replace the `is.null(NL)` branch inside `cde`, which
currently calls `nlFromSnrInfo`. Everything it needs is already an argument to
`cde`. `nlFromSnrInfo` then becomes a deprecated wrapper, following the same
pattern as `fitSNRdetectionFunc`.

The standard deviation is taken straight from the detections. The second plot
shows why that is nearly right and not exactly right: filtering narrows curve 3
a little. Solving for the mean and the standard deviation together would remove
that, at the cost of turning a one dimensional root find into a two dimensional
optimisation. Worth measuring the residual first and only fixing it if it
matters.

# What this does not fix

Two assumptions run through every plot above and neither is touched by any of
this.

Calls are assumed spread evenly in two dimensions. The area weight
`#r * (r <= td)` in `pDetGivenNL` is that assumption written down. Clustered
calling, or calls concentrated at particular depths, changes it.

Noise is assumed independent of whether whales are calling. If loud hours are
also whale hours, the detection curve is not the only thing shaping curve 3,
and the correction is solving the wrong problem accurately.

# References
