---
title: "callDensity: Estimating noise levels from detections"
output: 
  html_document: default
  pdf_document: default
  rmarkdown::html_vignette:
    vignette: >
      %\VignetteEncoding{UTF-8}
      %\VignetteIndexEntry{callDensity: Estimating noise levels from detections}
      %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
editor_options: 
  markdown: 
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, fig.width = 6, fig.height = 3.5,
                      comment = "#>")
```

```{r setup}
library(callDensity)
library(ggplot2)
library(kableExtra)
library(future)
library(future.apply)
library(ggside)
```

# The problem in one paragraph

For single sensor call density estimates, we rely on a Monte Carlo simulation to
estimate the probability of detection in the area. This simulation,
`pDetInArea`, requires a distribution of noise levels, NL, as an input
parameter. An obvious place to get such a distribution is from the NL that are
associated with each detection (i.e. NL used in the denominator to estimate SNR
when fitting a detection function). However, those measurements are biased low.
A quiet period gives you a high SNR, and a high SNR gives you a detection, so
your detections oversample the quiet periods. Measure noise only where you
detected something and you will conclude the ocean is quieter than it is.

This vignette shows how big that bias is, where it comes from, and how to
correct for it. In Beyond Counting Calls [@castro2024], the mean and standard
deviation of noise levels, NL, were estimated directly from detections, with no
correction -- a reasonable thing to do at the time without suspecting a
correction was required. By the Common Ground manuscript [@miller2026], the need
for a correction had emerged, and `nlFromSnrInfo` was the result, but it
corrects using a property of the detector only, and did not account for the
geometry that also drives the bias, so it was never a general solution (see
"Where this leaves the old function" below for more on why). This vignette
proposes and investigates methods to correct the bias that arises from reusing
the NL measured while fitting the detection function as the NL distribution
`pDetInArea`'s Monte Carlo simulation draws from.

# From an average to a curve

`pDetInArea` , which is used by `cde` for call density estimation, already
computes something close to what we need to correct the bias in the distribution
of NL from detected calls. This function answers one question: what fraction of
the calls in the study area get detected averaged over a whole distribution of
noise levels and source levels? That fraction is $\hat{p}_a$, a single number,
and `pDetInArea` gets it by Monte Carlo simulation: drawing NL and source
levels, SL, from a normal distribution and averaging the detection probability
over the draws.

We can change one thing about that calculation: instead of drawing NL and SL
from a distribution, hold it fixed at one value. The output is no longer a
single averaged number, it is the detection probability *at that noise level*.
Do this at 70 dB, then 74, then 78, and so on, and the outputs trace out a
curve: probability of detection against noise level. Call it `pDetGivenNL`.

This curve is the key to the whole solution. `pDetInArea` already knows how to
compute detection probability from a distribution of noise levels, it just
averages the answer into a single number, $\hat{p}_a$. Hold the noise level
fixed instead of drawing it from a distribution, and the same calculation
returns detection probability at that one noise level -- exactly the quantity
that determines how strongly detections favour quiet periods over loud ones,
since a quiet spell means more of the study area is audible and a loud spell
means less. `pDetGivenNL` isn't new machinery bolted onto `pDetInArea`, it's the
same machinery run one noise level at a time instead of averaged. What the rest
of this vignette does with it is quantify the bias described in the opening
paragraph, and then correct for it.

# Building a simplified detection range curve

`pDetGivenNL` takes the same transmission loss table and source level
distribution as `pDetInArea` and `cde`, plus a vector of noise levels to
evaluate at instead of a distribution to draw from.

The simple spherical spreading transmission loss table and source levels below
are used throughout the rest of the vignette, though we later demonstrate that
the solution works for more realistic numerical models of TL from more realistic
environments as well. The detector is simulated as a plain logistic, so, this
section does not depend on any other external datasets.

```{r build.curve}

R  <- 1e6
SL <- data.frame(mean = 190, sd = 4, sampleSize = 350)

TL <- simTLradials_20logR(maxRange = R, rangeStep = 1000, numTransects = 4)
TL <- TL[is.finite(rowSums(TL)) & TL[, 1] > 0, ]     # drop r = 0 if present

detector <- function(snr) plogis(snr, location = 1, scale = 2)

nlGrid <- seq(from = 55, to = 135, by = 1)
pCurve <- pDetGivenNL(nlGrid, detector, SL, TL)

curveDf <- data.frame(nl = nlGrid, pDet = pCurve)

```

## Noise level versus audible area (AKA detection range)

In this section we create a plot to illustrate how `pDetGivenNL` works. Before
looking at the plot it is important to understand that noise levels on this plot
are all hypothetical rather than actual measurements. Every point along the line
asks for the given SL and detection function, *if the noise were always exactly
this loud, and never varied, what fraction of the study area would yield
detections?*

That fraction is just detection area divided by study area. So this plot is
really about detection range, which is a more familiar quantity anyway. Adding
these ranges at a few points along the curve provides references that aid
interpretation for those who think more naturally about detection range rather
than about $\hat{p}_a$.

```{r curve.plot, warning=FALSE, fig.height=4}

# Range at which a call of mean source level sits at the detector's 50% point.
# Approximate. Ignores the spread in SL and the smoothness of the detector, so
# it is a label, not a calculation the curve depends on.
detRadius_km <- function(nl, SLmean = 190, loc = 1) {
  10^((SLmean - nl - loc) / 20) / 1000
}

marks <- data.frame(nl = c(74, 84, 94, 104))
marks$pDet  <- pDetGivenNL(marks$nl, detector, SL, TL)
marks$label <- sprintf("%.0f km", detRadius_km(marks$nl))

# Reference line: one decade per 10 dB, anchored at 90 dB. This slope is a
# direct consequence of spherical spreading (TL = 20 log10(r)) plus a
# step-function detector: to hold SNR constant, range must shrink by
# 10^(dNL/20) for each dB of extra noise, so area -- and therefore detection
# probability -- shrinks by 10^(dNL/10). It is tied to this TL model
# specifically, which is exactly why it is used as a reference line rather
# than the curve itself, and why the comparison against pyRAM later in this
# vignette anchors to the same line. It is a bare exponential with no ceiling,
# so it is clamped at 1 -- probability cannot exceed that, even if the formula
# it comes from does not know that.
anchor <- pDetGivenNL(90, detector, SL, TL)
refDf  <- data.frame(nl = nlGrid,
                     pDet = pmin(anchor * 10^(-(nlGrid - 90) / 10), 1))

ggplot(curveDf, aes(x = nl, y = pDet)) +
  geom_line(data = refDf, linetype = "dashed", colour = "grey50",
            linewidth = 0.6) +
  geom_line(linewidth = 0.9, colour = "steelblue") +
  geom_point(data = marks, size = 2, colour = "grey20") +
  geom_text(data = marks, aes(label = label), hjust = -0.25, vjust = -0.4,
            size = 3, colour = "grey20") +
  geom_vline(xintercept = 84, colour = "firebrick", linetype = "dotted") +
  annotate("text", x = 84.5, y = 1e-4, label = "true mean NL",
           colour = "firebrick", hjust = 0, size = 3) +
  scale_y_log10() +
  xlab("Noise level (dB)") +
  ylab("Fraction of the study area you can hear into") +
  labs(caption = "Labels: detection radius. Dashed grey: one decade per 10 dB, spherical spreading")

```

Working through two of those labelled points by hand helps the plot stop being
abstract.

**At 84 dB.** A 190 dB call reaches the detector's 50% point when SNR is 1 dB,
so when TL is $190 - 84 - 1 = 105$ dB. Under spherical spreading that is
$10^{105/20}$ metres, or 178 km. So you are hearing calls out to roughly 178 km.
The area you can hear into is $\pi \times 178^2 \approx 99{,}500$ km². The study
area is $\pi \times 1000^2 \approx 3{,}140{,}000$ km². You are hearing into
about 3% of it.

**At 94 dB.** Every call now needs 10 dB more received level. Ten decibels of
spherical spreading is a factor of 3.16 in range. So 178 km becomes 56 km. The
area is $\pi \times 56^2 \approx 9{,}850$ km², one tenth of what it was. You are
hearing into 0.3% of the study area.

Ten decibels louder, ten times less ocean, ten times fewer calls. Each 10 dB
step to the right divides the curve by 10. On a log axis that is a straight
line, and that is all the dashed reference line is saying.

**The shoulder on the left.** Below about 69 dB the labelled radius would exceed
1000 km. For simple spherical spreading, this is past the edge of the study
area. In this regime, the curve flattens to 1 and making the sea quieter yields
no additional detections *in the defined study area*. That shoulder is exactly
where the tidy algebra stops working, and it is one of the reasons why the
function `pDetGivenNL` numerically computes a curve rather than simply fitting a
slope to it.

# Illustrating why NL from detected calls is biased

**How the detection range curve relates to true NL and measured NL.** So far
everything mentioned about the NL in the above plot has been abstract and
hypothetical. Now it is time to make it more realistic. Let the red line mark
the true mean noise level, out on the variable portion where a decibel of noise
really does decrease detections.

The same curve is now plotted on a linear axis below, zoomed into the range the
noise actually occupies, and it better illustrates the solution.

Now use the curve: nature draws a noise level for each call from a normal
distribution. The detection process then throws most of them away, and it throws
away loud ones far more eagerly than quiet ones. The curves below illustrate
exactly how much more eagerly.

```{r filter.plot, warning=FALSE, fig.height=4}

nlTrue  <- 84
nlSd    <- 4

fine  <- seq(nlTrue - 5 * nlSd, nlTrue + 5 * nlSd, length.out = 400)
fN    <- dnorm(fine, nlTrue, nlSd)
pd    <- pDetGivenNL(fine, detector, SL, TL)
wt    <- fN * pd

# Scale each curve to a common height so the shift is visible
filterDf <- rbind(
  data.frame(nl = fine, y = fN / max(fN),  what = "1. All calls (the truth)"),
  data.frame(nl = fine, y = pd / max(pd),  what = "2. Chance of detection"),
  data.frame(nl = fine, y = wt / max(wt),  what = "3. What you measure")
)

meanObs <- sum(fine * wt) / sum(wt)

ggplot(filterDf, aes(x = nl, y = y, colour = what)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = c("1. All calls (the truth)"   = "firebrick",
                                 "2. Chance of detection"      = "seagreen",
                                 "3. What you measure"         = "steelblue")) +
  geom_vline(xintercept = nlTrue,  colour = "firebrick", linetype = "dotted",
             linewidth = 1.1) +
  geom_vline(xintercept = meanObs, colour = "steelblue", linetype = "dashed",
             linewidth = 1.1) +
  annotate("segment", x = meanObs, xend = nlTrue, y = 1.05, yend = 1.05,
           arrow = arrow(length = unit(2, "mm"), ends = "both")) +
  annotate("text", x = (meanObs + nlTrue) / 2, y = 1.12,
           label = sprintf("bias = %.2f dB", nlTrue - meanObs), size = 3) +
  xlab("Noise level (dB)") +
  ylab("Scaled to a common height") +
  labs(colour = NULL) +
  theme(legend.position = "top")

```

The red curve is the truth. The green curve is the detection curve from the
previous plot, now drawn on a linear axis over the range the noise actually
occupies. The blue curve is the red curve multiplied by the green curve, and it
is also the NL that would be measured from the detections used to fit the
detection function.

Multiplying the true value by a curve that leans left drags the peak left. 

However, there are two additional things worth noticing, because both get used
later.

The blue curve is narrower than the red curve by very little. The width survives
the filtering almost intact, which means the standard deviation of your measured
noise is near the true value even though the mean is not. If we assume that the
measured standard deviation is close enough to the truth to use it as is, then
that leaves only one unknown to solve for: the size of the leftward shift (i.e.
the bias in the mean).

The size of the shift depends on how steeply the detection function declines,
and on how wide the true noise distribution is.

# Correcting the bias

The previous plot is a prediction machine. Give it a candidate true mean, and it
tells you what mean you should have measured. That machine is `predictSampledNL`,
shown here in seagreen -- the same function used throughout this vignette, now
swept across a range of candidate true means instead of evaluated once.

We already know what we measured, and we have already estimated the detection
function (because it is required by `pDetInArea`), so we can run this prediction
machine backwards.

```{r rootfind.plot, warning=FALSE, fig.height=4.5, fig.width=4.5}

library(ggside)

muTry <- seq(75, 92, by = 1)
muObs <- vapply(muTry, function(m)
  predictSampledNL(mu = m, sigma = nlSd, detFun = detector, SL = SL, TL = TL),
  numeric(1))

rootDf <- data.frame(muTrue = muTry, muMeasured = muObs)

# The lookup itself: read across from the known measured value (y-axis) to
# the curve, then down to the candidate true mean that produced it. uniroot
# does this numerically; the arrows below do it visually, landing exactly on
# uniroot's own answer rather than on a separately-marked "true" position.
recovered <- uniroot(
  function(mu) predictSampledNL(mu = mu, sigma = nlSd, detFun = detector,
                                SL = SL, TL = TL) - meanObs,
  interval = c(meanObs, meanObs + 25))$root

xFloor <- min(muTry)
yFloor <- xFloor
xCeil <- 92

# curve 1 and curve 3: reuse fine/fN/wt directly from the filter.plot chunk,
# scaled to a common height the same way that plot already does (divide by
# each curve's own max), rather than recomputing anything fresh here.
keepRange  <- fine >= xFloor & fine <= xCeil
trueDensDf <- data.frame(x = fine[keepRange], dens = fN[keepRange] / max(fN))
measDensDf <- data.frame(y = fine[keepRange], dens = wt[keepRange] / max(wt))

ggplot(rootDf, aes(x = muTrue, y = muMeasured)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey70", linetype = "dashed") +
  geom_line(linewidth = 0.9, colour = "seagreen") +
  # Step 1: across, from what you measured to the point on the curve
  annotate("segment", x = xFloor, xend = recovered, y = meanObs, yend = meanObs,
           colour = "steelblue", linewidth = 1.1,
           arrow = arrow(length = unit(3, "mm"), type = "closed")) +
  # Step 2: down, from that point to the candidate true mean it came from
  annotate("segment", x = recovered, xend = recovered, y = meanObs, yend = yFloor,
           colour = "firebrick", linewidth = 1.1,
           arrow = arrow(length = unit(3, "mm"), type = "closed")) +
  annotate("point", x = recovered, y = meanObs, size = 2.5, colour = "black") +
  annotate("text", x = xFloor + 0.2, y = meanObs, hjust = 0, vjust = -0.7,
           size = 3, colour = "steelblue", label = "measured") +
  annotate("text", x = recovered, y = yFloor, hjust = -0.15, vjust = 0,
           size = 3, colour = "firebrick", label = "recovered NL") +
  annotate("text", x = max(rootDf$muTrue) - 1,
          y = rootDf$muMeasured[which.max(rootDf$muTrue)] + 0.2,
          hjust = 1.1, size = 3, angle=42, colour = "seagreen",
          label = "predictSampledNL(trueNL, nlSd)") +
  geom_xsideline(data = trueDensDf, aes(x = x, y = dens), inherit.aes = FALSE,
                colour = "firebrick", linewidth = 0.8) +
  geom_ysideline(data = measDensDf, aes(x = dens, y = y), inherit.aes = FALSE,
                colour = "steelblue", linewidth = 0.8) +
  ggside(y.pos = "left", x.pos="bottom") +
  coord_fixed(ratio=1, xlim = c(xFloor, xCeil), ylim = c(xFloor, xCeil), expand=FALSE) +
  scale_x_continuous(position='top')+
  scale_y_continuous(position='right')+
  xlab("Candidate true mean NL (dB)") +
  ylab("Measured NL") +
  labs(caption = "Dashed grey: the 1:1 line you would get with no bias") +
  theme_minimal() +
  theme(ggside.panel.scale = 0.2) +
  theme_ggside_void()


```

The green line sits below the 1:1 line by the size of the bias. It rises
steadily, so there is exactly one candidate that reproduces your measurement.
The two arrows are the lookup, drawn: across from what you measured (blue) to
the point on the curve, then down (red) to read off the candidate that produced
it. That point and that landing spot are not illustrative approximations -- they
are `uniroot`'s own answer, plotted.

`nlFromDetections` is the package function that does exactly this (`uniroot`
over `predictSampledNL`, same shape as above). It also defaults to reading noise
measurements from a `NoiseRL` column, so it accepts an `snrInfo` table without
modification:

```{r nlfromdetections-demo}

nlFromDetections(snrInfo = data.frame(NoiseRL = rnorm(500, meanObs, nlSd)),
                 snrDetFun = detector, SL = SL, TL = TL)

```

That's a synthetic `snrInfo` built from the biased mean this vignette already
computed, just to show the call shape. `mean` should land close to `nlTrue` (84
dB); `sd` is passed through from the input essentially unchanged, per the note
two paragraphs up about curve 3's width.

# Where this leaves the old function (pre callDensity v1.1.1)

Prior to callDensity version 1.1.1, `nlFromSnrInfo` was used to correct this
bias by adding the SNR at which the detection function reaches 0.5.
`nlFromSnrInfo` was always a stopgap. It corrects the bias using measured NL and
a property of the detector (the SNR at its 50% point). This derived from the
example scenario, where it arrived at the correct answer coincidentally, **not**
because it was the thing driving the bias.

`nlFromDetections` is derived from the measured NL, detector, and geometry that
actually produces the bias, and is consistently less biased across every
detector and site this has been tested against. `nlFromSnrInfo` is retained as a
deprecated wrapper for old scripts, but there is no situation where it should be
preferred over `nlFromDetections`. `cde` already defaults to `nlFromDetections`
when `NL` is not supplied.

# Solution built from a lookup table, not a formula

`pDetGivenNL` was never built around spherical spreading, or any other
propagation formula. It reads transmission loss straight out of whatever table
you hand it, and evaluates the sonar equation against that table, cell by cell.
Swap the spherical spreading table for a more realistic model (e.g. pyRAM
output, AcTUP output, anything else with the same range and TL columns), and the
function does not need to change, because it never encoded an assumption about
what the physics look like in the first place. The dashed reference line on the
first plot is not part of that machinery: it is one candidate formula
($20\log_{10}r$), included there as a comparison so any departure from it is
visible, not as anything `pDetGivenNL` relies on.

Spherical spreading is not a serious model for a baleen whale call at 25 Hz in
2000 m of water at 62°S. So below we use a real one to demonstrate that
`nlFromDetections` is TL-model agnostic. The file below is pyRAM output computed
on GLORYS12 sound speed profiles along 24 radial transects out to 1000 km, for
the South Kerguelen Plateau mooring in summer, at 25 Hz.

## First, look at the data

```{r pyram.inspect}

dim(kerguelen2015TL)
names(kerguelen2015TL)
range(kerguelen2015TL$range_m)

kerguelen2015TLmeta[, c("azimuths", "TruncationDistance_m")]

```

`kerguelen2015TL` and `kerguelen2015TLmeta` are TL data that ship with the
callDensity package (`?kerguelen2015TL`, `?kerguelen2015TLmeta`).
`kerguelen2015TLmeta` records where each of the 24 transects actually runs onto
land, using the pyRAM pipeline's bathymetry source
(`cmems_mod_glo_phy_my_0.083deg_static`).

The layout is already what `pDetGivenNL` wants: `range_m` first, then 24 columns
named by bearing from `tl000` to `tl345`. Range runs 100 m to 1000 km in 100 m
steps, so 10,000 rows.

```{r pyram.prepare}

# Drop any non-finite cells. Range is already in metres and starts above zero.
TLpyram <- kerguelen2015TL[is.finite(rowSums(kerguelen2015TL)), ]

sprintf("%d transects, range step %.0f m, TL at 1000 km: %.1f to %.1f dB",
        ncol(TLpyram) - 1,
        diff(TLpyram$range_m)[1],
        min(as.numeric(TLpyram[nrow(TLpyram), -1])),
        max(as.numeric(TLpyram[nrow(TLpyram), -1])))

# Match truncation distances to transects by azimuth, not by row position --
# two objects agreeing on row order is an assumption, not a guarantee, and
# this is exactly the kind of silent misalignment that would be invisible in
# the output. tl015 gets the metadata row where azimuths == 15, full stop.
transectNames <- names(TLpyram)[-1]                      # tl000 .. tl345
transectAz    <- as.numeric(sub("^tl", "", transectNames))
pyramTrunc    <- kerguelen2015TLmeta$TruncationDistance_m[
                   match(transectAz, kerguelen2015TLmeta$azimuths)]
stopifnot("Every transect column must have a matching azimuth in kerguelen2015TLmeta" =
            !anyNA(pyramTrunc))

truncated <- pyramTrunc < max(TLpyram$range_m)
sprintf("%d of %d transects hit land before %.0f km: %s km (bearings %s)",
        sum(truncated), length(pyramTrunc), max(TLpyram$range_m) / 1000,
        paste(round(pyramTrunc[truncated] / 1000), collapse = ", "),
        paste(transectAz[truncated], collapse = ", "))

```

The spread in TL is very large, but that is because it is computed from every
bearing's row at 1000 km regardless of whether that bearing is still in water
there. Eight of the 24 bearings hit land between 410 and 950 km (bearings 120°
to 225°, roughly south through the plateau), so their rows beyond that point
describe whatever pyRAM did past the coastline, not propagation through open
water. `TLpyram[nrow(TLpyram), -1]` mixes those in with the sixteen genuinely
open-water bearings. It is a fair description of what the raw file contains, and
not yet a fair description of the site. For this reason, pDetGivenNL and
pDetInArea can take a truncationDistance as an input, and this can be used to
mask (i.e. remove) land from the study area when estimating `NL` and `p_a`.

## The audible-area curve, spherical spreading vs realistic TL

```{r pyram.curve, warning=FALSE, fig.height=4}

pPyram <- pDetGivenNL(nlGrid, detector, SL, TLpyram,
                      truncationDistance = pyramTrunc)

bothDf <- rbind(
  data.frame(nl = nlGrid, pDet = pCurve, what = "Spherical spreading"),
  data.frame(nl = nlGrid, pDet = pPyram, what = "pyRAM, SKP summer, 25 Hz")
)

anchorP <- pDetGivenNL(90, detector, SL, TLpyram, truncationDistance = pyramTrunc)
refBoth <- data.frame(nl = nlGrid,
                      pDet = pmin(anchorP * 10^(-(nlGrid - 90) / 10), 1))

ggplot(bothDf, aes(x = nl, y = pDet, colour = what)) +
  geom_line(data = refBoth, aes(x = nl, y = pDet), inherit.aes = FALSE,
            linetype = "dashed", colour = "grey50", linewidth = 0.6) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = 84, colour = "firebrick", linetype = "dotted") +
  scale_y_log10() +
  xlab("Noise level (dB)") +
  ylab("Fraction of the study area you can hear into") +
  labs(colour = NULL,
       caption = "Dashed grey: one decade per 10 dB, anchored on the pyRAM curve") +
  theme(legend.position = "top")

```

`truncationDistance = pyramTrunc` changes what "the study area" on the y-axis
means. Without it, the denominator is the full 1000 km circle, including the
land eight of these bearings never reach. With it, the denominator is the actual
water area `cde()` would use, via the same `td^2`-per-sector logic `studyArea()`
applies. This curve is now the one that matches what `p_a` inside `cde()`
actually estimates for this site, not a hypothetical circle.

What to look for, and what it means.

**The slope should not be one decade per 10 dB.** At 25 Hz in 2000 m of water
the sound is in a waveguide, not a free field. Beyond a few water depths it
spreads more like a cylinder than a sphere, which makes TL climb more slowly
with range. Slower TL means a decibel of extra noise costs you a wider annulus,
so the curve should fall *faster* than the dashed line, not slower.

**The curve should be bumpy.** Convergence zones and bottom interaction add and
remove audible ocean at particular ranges. Wiggles are physics, not noise.

**The shoulder should be further left, or gone.** Flatter TL means longer
detection range, so it takes a quieter sea to run out of study area.

**One thing worth checking before trusting the output:** `pDetGivenNL`'s only
approximation is the TL bin width, and a bumpy curve is exactly the case where
too coarse a bin could blur real structure away rather than just smooth noise.

```{r pyram.binwidth, warning=FALSE}

binCheck <- data.frame(binWidth = c(1, 0.5, 0.25, 0.1))
binCheck$pDet84 <- vapply(binCheck$binWidth, function(b)
  pDetGivenNL(84, detector, SL, TLpyram, truncationDistance = pyramTrunc,
             binWidth = b), numeric(1))
binCheck$nBins <- vapply(binCheck$binWidth, function(b)
  length(seq(40, 140, by = b)), numeric(1))

binCheck

```

Trustworthy results from `pDetGivenNL` should be stable with with bin width.  
Nothing about using a real, bumpy curve instead of a synthetic one changes that.

### The bias using a more realistic TL

Using the same detection function `detFun(SNR)`, we apply the same mechanism.
The only difference is which `pDetGivenNL` curve does the filtering. We swap the
spherical TL for `TLpyram`, the more realistic transmission loss table computed
with pyRAM, while all other inputs stay the same.

```{r pyram.filter, warning=FALSE, fig.height=4}

finePyram <- seq(nlTrue - 5 * nlSd, nlTrue + 5 * nlSd, length.out = 400)
fNPyram   <- dnorm(finePyram, nlTrue, nlSd)
pdPyram   <- pDetGivenNL(finePyram, detector, SL, TLpyram,
                        truncationDistance = pyramTrunc)
wtPyram   <- fNPyram * pdPyram

filterPyramDf <- rbind(
  data.frame(nl = finePyram, y = fNPyram / max(fNPyram),
            what = "1. All calls (the truth)"),
  data.frame(nl = finePyram, y = pdPyram / max(pdPyram),
            what = "2. Chance of detection"),
  data.frame(nl = finePyram, y = wtPyram / max(wtPyram),
            what = "3. What you measure")
)

measuredMasked <- sum(finePyram * wtPyram) / sum(wtPyram)

ggplot(filterPyramDf, aes(x = nl, y = y, colour = what)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = c("1. All calls (the truth)"   = "firebrick",
                                 "2. Chance of detection"      = "seagreen",
                                 "3. What you measure"         = "steelblue")) +
  geom_vline(xintercept = nlTrue,       colour = "firebrick", linetype = "dotted",
            linewidth = 1.1) +
  geom_vline(xintercept = measuredMasked, colour = "steelblue", linetype = "dashed",
            linewidth = 1.1) +
  annotate("segment", x = measuredMasked, xend = nlTrue, y = 1.05, yend = 1.05,
           arrow = arrow(length = unit(2, "mm"), ends = "both")) +
  annotate("text", x = (measuredMasked + nlTrue) / 2, y = 1.12,
           label = sprintf("bias = %.2f dB", nlTrue - measuredMasked), size = 3) +
  xlab("Noise level (dB)") +
  ylab("Scaled to a common height") +
  labs(colour = NULL) +
  theme(legend.position = "top")

```

The bias using the more realistic TL can still be estimated the exact same way
and sits at `r round(nlTrue - measuredMasked, 2)` dB.

## Does land masking matter for this bias?

`pDetGivenNL` needs the same `truncationDistance` as whatever `pDetInArea` call
it is meant to support. Passing a different one, or none at all, does not
approximate the truth -- it silently produces a different estimate of NL, tied
to whichever land mask happened to be used, with no guarantee the two would
agree. Always pass the same truncation-distance vector used elsewhere in the
same analysis, so the land mask is treated identically by both functions rather
than becoming its own silent source of disagreement.

```{r pyram.seasons, eval=FALSE}

# WORTH DOING, NOT DONE. The bias is a property of the propagation, so it
# should vary by season as the sound speed profile changes, and by site. Loop
# the four seasons and the seven AWR sites and tabulate the bias. If it moves
# by more than a decibel between seasons, that is an argument for estimating NL
# seasonally, which is what Castro et al. (2024) recommend for f anyway.

```

# When does the correction still hold?

`nlFromDetections` substitutes the sample standard deviation of your detections
for the true standard deviation of the whole noise population, and that
substitution is where its residual bias comes from. The residual is small
whenever the local slope of `log(pDetGivenNL(nl))` stays close to constant
across the range the true noise distribution actually spans — not everywhere,
just there. A constant local slope means `log(pDetGivenNL)` is locally
exponential, and multiplying a Gaussian by an exponential shifts the mean
without touching the variance, which is exactly why the substitution costs
nothing in that regime.

Three things control how close to constant that local slope stays: how steep the
detector is, how far the propagation departs from a power law, and how wide the
true noise distribution is relative to both. None of the three determines it
alone — `pDetGivenNL` is a single curve built from detector and propagation
together, and the local slope's steadiness is a property of that combined curve,
not of either ingredient separately. The only way to find out which combinations
keep the local slope near-constant over the relevant range is to sweep all three
at once.

This residual behaves differently depending on where you are in parameter space.
What follows describes that behaviour directly, for the specific detectors,
propagation models, and NL widths swept here -- not necessarily as a general
property of nlFromDetections.

## A sweep, computed exactly

Implementation note: the residual can be computed without simulating a single
call. `meanObs` and its standard deviation are both weighted moments of the same
`fN * pDetGivenNL(nl)` product already used to build the three-curve plots, so
both come from the same deterministic integral, no repeated sampling and no
Monte Carlo noise to average out.

```{r residual.function}

#' Residual bias in nlFromDetections's recovered mean, computed exactly.
#'
#' Builds the same weighted mean and weighted variance the three-curve plots
#' use, then inverts predictSampledNL twice: once with the true sd (which
#' should recover nlTrue exactly, as a check), and once with the sample sd of
#' the detections (which is what nlFromDetections actually receives). The
#' difference between the second recovery and nlTrue is the residual.
residualNL <- function(detFun, TLtab, truncDist, sd, nlTrue = 84) {

  fine <- seq(nlTrue - 6 * sd, nlTrue + 6 * sd, length.out = 600)
  fN   <- dnorm(fine, nlTrue, sd)
  pd   <- pDetGivenNL(fine, detFun, SL, TLtab, truncationDistance = truncDist)
  wt   <- fN * pd

  if (sum(wt) <= 0 || !all(is.finite(wt))) return(NA_real_)

  meanObs  <- sum(fine * wt) / sum(wt)
  sigmaObs <- sqrt(sum((fine - meanObs)^2 * wt) / sum(wt))

  recoveredSample <- tryCatch(
    uniroot(function(mu) predictSampledNL(mu = mu, sigma = sigmaObs,
                                          detFun = detFun, SL = SL,
                                          TL = TLtab,
                                          truncationDistance = truncDist) - meanObs,
           interval = c(meanObs, meanObs + 40))$root,
    error = function(e) NA_real_)

  recoveredSample - nlTrue
}

```

```{r residual.sweep}

# Detector steepness: plogis's scale parameter, smaller is steeper.
detScaleGrid <- c(0.5, 1, 2, 4, 8)

# Width of the true NL distribution, dB.
sdGrid <- c(1, 2, 3, 4, 6, 8)

# Mean NL -- the same four landmark values already used to label detection
# radius in "Noise level versus audible area", so this reuses points already
# seen rather than introducing new ones. For spherical spreading, moving this
# only matters near the shoulder (a ceiling-at-1 effect, unrelated to
# propagation); for pyRAM, it also matters because it slides the Gaussian
# window along a curve with real wiggles in it, and whether a given nlSd
# catches a smooth stretch or a kink depends on where the window sits.
nlTrueGrid <- c(74, 84, 94, 104)

propModels <- list(
  spherical = list(TL = TL,      truncationDistance = max(TL[[1]])),
  pyRAM     = list(TL = TLpyram, truncationDistance = pyramTrunc)
)

sweepGrid <- expand.grid(detScale    = detScaleGrid,
                         nlSd        = sdGrid,
                         nlTrue      = nlTrueGrid,
                         propagation = names(propModels),
                         stringsAsFactors = FALSE)

# This runs residualNL() once per row. Every call re-bins the TL table (that
# is how pDetGivenNL works), and uniroot calls predictSampledNL, which calls
# pDetGivenNL again, several dozen times per row while it converges. Every row
# is independent of every other row, so this is embarrassingly parallel --
# same future.callr backend callDensity's own parallel functions use
# elsewhere. future.seed is deliberately omitted: residualNL never touches the
# RNG, so there is no seed for future to manage, and no need to force
# RNGkind() the way an actual simulation-based sweep would require.
future::plan(future.callr::callr, workers = future::availableCores(omit = 2))

sweepGrid$residual <- future_mapply(function(scale, sd, nlT, prop) {
  detFun <- function(snr) plogis(snr, location = 1, scale = scale)
  p <- propModels[[prop]]
  residualNL(detFun, p$TL, p$truncationDistance, sd, nlTrue = nlT)
}, sweepGrid$detScale, sweepGrid$nlSd, sweepGrid$nlTrue, sweepGrid$propagation,
future.packages = "callDensity")

future::plan(future::sequential)

```

```{r residual.plot, warning=FALSE, fig.height=5.5}

refPoint <- subset(sweepGrid, propagation == "spherical" & detScale == 2 &
                              nlSd == 4 & nlTrue == 84)

ggplot(sweepGrid, aes(x = nlSd, y = residual, colour = factor(detScale))) +
  geom_hline(yintercept = 0, colour = "grey60", linetype = "dashed") +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  geom_point(data = refPoint, aes(x = nlSd, y = residual), inherit.aes = FALSE,
            shape = 4, size = 4, stroke = 1.5, colour = "black") +
  facet_grid(propagation ~ nlTrue,
            labeller = labeller(nlTrue = as_labeller(function(x) paste0(x, " dB")))) +
  scale_colour_viridis_d(name = "Detector scale\n(smaller = steeper)") +
  xlab("True NL standard deviation (dB)") +
  ylab("Residual bias in nlFromDetections (dB)") +
  labs(caption = "Black X: the spherical, scale = 2, nlSd = 4, nlTrue = 84 case\nused throughout the rest of this document") +
  theme(legend.position = "right")

```

Across this grid, the residual grows as the detector gets steeper (smaller
`detScale`), and grows as `nlSd` gets wider, in both propagation models.

Two other patterns are visible here too. pyRAM is not uniformly more or less
biased than spherical: at 74 and 84 dB pyRAM is less biased, while at 94 dB and
especially 104 dB it is markedly more biased, including one combination --
`nlSd` 4-6 at `nlTrue` = 104 -- where `nlFromDetections` overshoots `nlTrue`
rather than undershooting it, the only sign flip anywhere in this grid. The
spherical row is also not flat across `nlTrue`: it peaks in magnitude around
84-94 dB and is smaller again at both 74 and 104.

These patterns hold for the detectors, propagation models, and NL widths tested
here. Nothing here establishes them as general, and no mechanism is derived for
the sign flip or for the non-monotonic pattern across `nlTrue`.



```{r residual.extremes}

sweepGrid[order(-abs(sweepGrid$residual)), ][1:5, ]
sweepGrid[order(abs(sweepGrid$residual)), ][1:5, ]

```

 The black X in the sweep plot above marks that case directly -- spherical,
scale = 2, nlSd = 4, nlTrue = 84.

## Looking at the curve directly

The aggregate residual says something odd happens for pyRAM around `nlTrue` =
104, `nlSd` 4-6. It cannot say what, because it is a single number per cell,
already collapsed from the curve that produced it. "The same curve, real
physics" earlier in this document already plots `pDetGivenNL` on a log axis over
the full 55-135 dB range, which implicitly shows the same `log(pDetGivenNL)`
this section's curvature argument is about -- slope on that plot is exactly the
quantity that matters. But reading curvature by eye off a full-range line plot
is exactly the wrong tool for finding one localised bend, so this zooms into the
specific window the anomaly lives in and plots the slope itself, rather than a
curve to eyeball for slope changes.

```{r pyram.slope.diagnostic, warning=FALSE, fig.height=4.5}

# Zoom to roughly the union of the nlSd = 4 and nlSd = 6 windows around
# nlTrue = 104, where the sign flip in the sweep shows up. Faceted across
# three of the detector scales that showed the sign flip in the extremes
# table (0.5, 1, 2), since the single fixed detector used everywhere else in
# this vignette does not establish whether this shape is specific to that
# one detector or holds more generally.
nlFine <- seq(65, 145, by = 0.5)
slopeDetScales <- c(0.5, 1, 2, 4, 8)

slopeDf <- do.call(rbind, lapply(slopeDetScales, function(sc) {
  detFunSc <- function(snr) plogis(snr, location = 1, scale = sc)

  pPyramFine <- pDetGivenNL(nlFine, detFunSc, SL, TLpyram,
                            truncationDistance = pyramTrunc)
  pSpherFine <- pDetGivenNL(nlFine, detFunSc, SL, TL)

  rbind(
    data.frame(nl    = nlFine[-1] - diff(nlFine) / 2,
              slope = diff(log(pPyramFine)) / diff(nlFine),
              what  = "pyRAM, SKP summer, 25 Hz",
              detScale = sc),
    data.frame(nl    = nlFine[-1] - diff(nlFine) / 2,
              slope = diff(log(pSpherFine)) / diff(nlFine),
              what  = "Spherical spreading",
              detScale = sc)
  )
}))

ggplot(slopeDf, aes(x = nl, y = slope, colour = detScale)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = 104, colour = "firebrick", linetype = "dotted") +
  annotate("rect", xmin = 104 - 4, xmax = 104 + 4, ymin = -Inf, ymax = Inf,
           fill = "grey50", alpha = 0.15) +
  annotate("rect", xmin = 104 - 6, xmax = 104 + 6, ymin = -Inf, ymax = Inf,
           fill = "grey50", alpha = 0.10) +
  facet_grid(what ~ detScale, 
            labeller = labeller(detScale = as_labeller(function(x) paste0("scale = ", x)))) +
  xlab("Noise level (dB)") +
  ylab("Local slope of log(pDetGivenNL) per dB") +
  scale_colour_viridis_c(name = "Detector scale\n(smaller = steeper)") +
  labs(colour = NULL,
       caption = "Shaded bands: the nlSd = 4 and nlSd = 6 integration windows around nlTrue = 104") +
  theme(legend.position = "top")

```

Across every scale tested, the same pattern holds in both propagation models: a
steeper detector (smaller `detScale`) shows more wiggle in the local-slope
curve, and a shallower one smooths it flat. At `scale = 4` and `8`, both curves
are close to featureless. At `0.5`, `1`, and `2`, both show a
dip-and-partial-recovery around 90-110 dB, and the pyRAM curve's recovery
clearly rises back above its own baseline slope into a genuine local increase --
the reversal behind the residual sweep's sign flip. This is consistent with a
specific mechanism. A steep detector responds sharply to a narrow slice of noise
level, so whatever structure the underlying curve has there -- pyRAM's real
bumps, or the truncation and source-level averaging built into the spherical
case -- shows up directly in `pDetGivenNL`'s local slope. A shallow detector
integrates its response over a much wider band of SNR, and that averaging blurs
the same structure away. This also supplies a mechanism, not just an observed
correlation, for the residual sweep's earlier finding that the residual grows as
the detector gets steeper.

One limit on how far this generalises: a perfect step-function detector, the
limit as `scale` approaches zero, was shown earlier in this vignette to give an
*exactly* log-linear curve under spherical spreading -- zero curvature, not
reduced curvature. Curvature cannot simply keep growing as `scale` keeps
shrinking forever, so the pattern seen here, across `0.5` to `8`, is real but is
not evidence it continues all the way to that limit.

The sharp features at the very right edge of each panel (~130-145 dB) have a
real physical explanation, now confirmed rather than assumed: the check below
shows the wiggle is identical across `binWidth` from 1 down to 0.05 dB
everywhere it appears, ruling out `pDetGivenNL`'s own numerical binning as the
cause. At these noise levels, the range at which SNR crosses the detector's
transition zone has shrunk to a tiny fraction of the study area -- around 1 km
at 130 dB under spherical spreading, about 0.175 km at 145 dB, several orders of
magnitude smaller than the 1000 km transects `pDetGivenNL` integrates over. A
detector's transition has some fixed width in SNR, and because TL compresses
range multiplicatively (equal steps in SNR correspond to shrinking steps in
range as range gets small), that fixed-width transition maps onto a badly
distorted range right where the area weighting is itself collapsing toward zero.
A perfect step-function detector has no transition width to distort this way,
and was shown earlier in this vignette to be exactly log-linear under spherical
spreading -- consistent with a real detector's finite smoothness being what
produces the wiggle here. Real detectors are smooth specifically because real
SNR measurements carry error, the same limitation already named in "What this
does not fix."

```{r pyram.slope.binwidth.check, warning=FALSE, fig.height=4}

# Does the wiggle at high NL survive finer TL binning, or is it partly (or
# wholly) a resolution artefact of pDetGivenNL's own binning? Same check as
# "Is the binning fine enough?" earlier in this vignette, run here instead,
# across the noise levels and detector scales where the wiggle shows up.
nlHighFine  <- seq(115, 145, by = 0.5)
binWidths   <- c(1, 0.5, 0.25, 0.1, 0.05)
checkScales <- c(0.5, 1, 2)

binCheckDf <- do.call(rbind, lapply(checkScales, function(sc) {
  detFunSc <- function(snr) plogis(snr, location = 1, scale = sc)
  do.call(rbind, lapply(binWidths, function(bw) {
    pPyramCheck <- pDetGivenNL(nlHighFine, detFunSc, SL, TLpyram,
                              truncationDistance = pyramTrunc, binWidth = bw)
    data.frame(nl       = nlHighFine[-1] - diff(nlHighFine) / 2,
              slope    = diff(log(pPyramCheck)) / diff(nlHighFine),
              binWidth = factor(bw), detScale = sc)
  }))
}))

ggplot(binCheckDf, aes(x = nl, y = slope, colour = binWidth)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ detScale,
            labeller = labeller(detScale = as_labeller(function(x) paste0("scale = ", x)))) +
  scale_colour_viridis_d(name = "binWidth (dB)") +
  xlab("Noise level (dB)") +
  ylab("Local slope of log(pDetGivenNL) per dB") +
  labs(caption = "If the wiggle is real physics, these lines should stack regardless of binWidth")

```


## Estimating NL's mean and standard deviation together, a 2D solver

`nlFromDetections` corrects the mean via a one-dimensional root find (`uniroot`
over `predictSampledNL`), and takes the standard deviation straight from the
detections without correcting it -- the substitution already named, at the start
of "When does the correction still hold?", as the source of its residual bias.
The natural next step is to solve for both the mean and the standard deviation
jointly instead of substituting one of them, which is what the two functions
below do.

Both directions of this inversion are built from the same weighted-moment
integral used throughout this vignette. The forward model,
`predictSampledMoments(mu, sigma, ...)`, evaluates a Gaussian with mean `mu` and
standard deviation `sigma` on a grid spanning `mu ± 6·sigma` (600 nodes),
weights it by `pDetGivenNL` at each point, and returns the resulting weighted
mean and weighted standard deviation -- the sample moments a detection-based
analysis would report, if the true population really had that `(mu, sigma)`.
This is the same forward calculation `predictSampledNL` already performs for the
mean alone, extended here to return both moments instead of one.

The inverse model, `recoverTrueMoments(obsMean, obsSd, ...)`, searches for the
`(mu, sigma)` pair whose forward-mapped moments best match a pair of observed
values, minimising the squared difference between predicted and observed mean
and sd. `sigma` is optimised on the log scale, so the search is unconstrained --
no boundary hack is needed to keep it positive -- using `optim`'s Nelder-Mead
method [@nelder1965simplex]: a derivative-free simplex search, evaluating the
objective only at the vertices of a moving simplex and reflecting, expanding, or
contracting it toward lower values, with no gradient or Hessian required. This
suits `predictSampledMoments`, which has no closed-form derivative available --
it is itself a numerical integral over `pDetGivenNL`, not a differentiable
formula -- and Nelder-Mead is `optim`'s own default method for exactly this kind
of low-dimensional, derivative-free search. The convenience has a real cost:
Nelder-Mead carries no guarantee of converging to even a local optimum, and is
known to stall at a non-stationary point on some smooth objective functions
[@mckinnon1998convergence]. It is used here for simplicity, not for any such
guarantee, started from `(obsMean + 2, log(obsSd))` -- a nudge in the direction
the bias is already known to run, not a principled prior.

Two diagnostics come back alongside the fitted values: `convergence`, which is
`0` only if `optim`'s own stopping rule fired, and the achieved objective
(reported elsewhere as `fitRMSE`), which is what actually says whether the fit
is good. Unlike `uniroot`, which guarantees a root given a valid bracket,
Nelder-Mead carries no such guarantee -- a real difference in robustness between
the one-dimensional correction already in the package and this two-dimensional
extension of it.

Before this solver ever sees real data, it is worth checking whether it can
recover a truth it is handed on a plate. The residual sweep above already has
ground truth built in -- every cell starts from a known `nlTrue` and `nlSd` --
so running the same two-parameter inversion on those same synthetic cells, and
checking whether it gets back to the values that generated them, is the natural
validation. It is also a fair test of the solver's own machinery on its own
terms, separate from any question about detector realism or real data.

```{r moment.functions}

#' Forward map: candidate true (mu, sigma) to the (mean, sd) you would
#' actually measure at detections. Same weighted-moment integral used
#' throughout this vignette, just returning both moments instead of one.
predictSampledMoments <- function(mu, sigma, detFun, SL, TL, truncDist,
                                  nNodes = 600) {
  fine <- seq(mu - 6 * sigma, mu + 6 * sigma, length.out = nNodes)
  fN   <- dnorm(fine, mu, sigma)
  pd   <- pDetGivenNL(fine, detFun, SL, TL, truncationDistance = truncDist)
  wt   <- fN * pd
  meanPred <- sum(fine * wt) / sum(wt)
  sdPred   <- sqrt(sum((fine - meanPred)^2 * wt) / sum(wt))
  c(mean = meanPred, sd = sdPred)
}

#' Inverse map: observed (mean, sd) to the candidate true (mu, sigma) that
#' would have produced them, via unconstrained least squares. sigma is
#' optimised on the log scale so no boundary hack is needed to keep it
#' positive. Returns the achieved objective alongside convergence, because
#' optim reporting convergence means its own stopping rule fired, not that
#' the fit is actually good -- a small residual is the thing that says that.
#' A proper package function would still want a dedicated 2D root-finder with
#' bracket guarantees, the way uniroot has for the 1D case; this does not.
recoverTrueMoments <- function(obsMean, obsSd, detFun, SL, TL, truncDist) {
  obsMean <- unname(obsMean)
  obsSd   <- unname(obsSd)
  obj <- function(par) {
    pred <- predictSampledMoments(par[1], exp(par[2]), detFun, SL, TL, truncDist)
    sum((pred - c(obsMean, obsSd))^2)
  }
  fit <- optim(c(obsMean + 2, log(obsSd)), obj, method = "Nelder-Mead",
              control = list(reltol = 1e-12, maxit = 2000))
  c(muTrue = fit$par[1], sigmaTrue = exp(fit$par[2]),
    convergence = fit$convergence, objective = fit$value)
}
```

```{r solver.validation}

# Deliberately a much smaller grid than the residual sweep -- each cell here
# costs an optim() run on top of what residualNL() already costs, so the full
# 5 x 6 x 4 x 2 grid would be prohibitively slow. Three detector scales
# (steep, middling, shallow), three true sd values, and both nlTrue = 84 and
# 104 (a "normal" cell and the one where the sign flip lives), across both
# propagation models.
recover2D <- function(detFun, TLtab, truncDist, sd, nlTrue = 84) {

  fine <- seq(nlTrue - 6 * sd, nlTrue + 6 * sd, length.out = 600)
  fN   <- dnorm(fine, nlTrue, sd)
  pd   <- pDetGivenNL(fine, detFun, SL, TLtab, truncationDistance = truncDist)
  wt   <- fN * pd

  if (sum(wt) <= 0 || !all(is.finite(wt))) {
    return(c(muRecovered = NA_real_, sigmaRecovered = NA_real_,
            muError = NA_real_, sigmaError = NA_real_,
            convergence = NA_integer_, fitRMSE = NA_real_))
  }

  meanObs  <- sum(fine * wt) / sum(wt)
  sigmaObs <- sqrt(sum((fine - meanObs)^2 * wt) / sum(wt))

  fit <- tryCatch(
    recoverTrueMoments(meanObs, sigmaObs, detFun, SL, TLtab, truncDist),
    error = function(e) c(muTrue = NA_real_, sigmaTrue = NA_real_,
                          convergence = NA_integer_, objective = NA_real_))

  c(muRecovered    = unname(fit["muTrue"]),
    sigmaRecovered = unname(fit["sigmaTrue"]),
    muError        = unname(fit["muTrue"]) - nlTrue,
    sigmaError     = unname(fit["sigmaTrue"]) - sd,
    convergence    = unname(fit["convergence"]),
    fitRMSE        = sqrt(unname(fit["objective"]) / 2))
}

detScaleGrid2D <- c(0.5, 2, 8)
sdGrid2D       <- c(2, 4, 8)
nlTrueGrid2D   <- c(84, 104)

validationGrid <- expand.grid(detScale    = detScaleGrid2D,
                              nlSd        = sdGrid2D,
                              nlTrue      = nlTrueGrid2D,
                              propagation = names(propModels),
                              stringsAsFactors = FALSE)

# One job per validation-grid row, plus the real "year" job from Beyond
# Counting Calls -- a genuinely different call shape (recoverTrueMoments
# directly on known real moments, not recover2D's forward-then-inverse
# round trip), but there is no reason to pay callr's worker-startup cost
# twice just because the two jobs look different. future_lapply over a list
# of closures dispatches both kinds through the same pool, concurrently.
validationJobs <- lapply(seq_len(nrow(validationGrid)), function(i) {
  row <- validationGrid[i, ]
  function() {
    detFun <- function(snr) plogis(snr, location = 1, scale = row$detScale)
    p <- propModels[[row$propagation]]
    recover2D(detFun, p$TL, p$truncationDistance, row$nlSd, nlTrue = row$nlTrue)
  }
})

yearJob <- function() recoverTrueMoments(88.6, 5.9, detector, SL, TLpyram, pyramTrunc)

allJobs <- c(validationJobs, list(year = yearJob))

future::plan(future.callr::callr, workers = future::availableCores(omit = 2))

allResults <- future_lapply(allJobs, function(f) f(), future.packages = "callDensity")

future::plan(future::sequential)

validationFits <- do.call(rbind, allResults[seq_len(nrow(validationGrid))])
validationGrid <- cbind(validationGrid, validationFits)

yearFit <- allResults[["year"]]

```

```{r solver.validation.plot, warning=FALSE, fig.height=4}

ggplot(validationGrid, aes(x = nlSd, y = muError, colour = factor(detScale))) +
  geom_hline(yintercept = 0, colour = "grey60", linetype = "dashed") +
  geom_point(size = 2) +
  geom_line(linewidth = 0.7) +
  facet_grid(propagation ~ nlTrue,
            labeller = labeller(nlTrue = as_labeller(function(x) paste0("nlTrue = ", x, " dB")))) +
  scale_colour_viridis_d(name = "Detector scale") +
  xlab("True NL standard deviation (dB)") +
  ylab("Recovered mu error (dB)") +
  labs(caption = "Should sit at zero everywhere, if the solver reliably recovers the truth")

```

```{r solver.validation.table}

validationGrid[order(-abs(validationGrid$muError)), 
              c("detScale", "nlSd", "nlTrue", "propagation",
                "muError", "sigmaError", "convergence", "fitRMSE")][1:10, ]

```

`muError` and `sigmaError` sit at roughly 1e-5 to 1e-6 dB across this whole
grid, with `convergence == 0` everywhere. That is a real validation of the
solver, not just a demonstration that it runs: whenever the truth actually is
Gaussian, this recovers it to a precision far tighter than anything else in this
vignette needs.

**On making this the default in `nlFromDetections`:** closer than it was, but
still not yet. This validation removes one of the concerns raised when this was
first built -- within the grid actually tested, the solver is trustworthy, not
just convergent. That grid is 36 cells, not the full residual sweep's 240, so it
has not been checked at every corner of detector, propagation, `nlTrue` and
`nlSd` this vignette has explored elsewhere; worth widening before leaning on
this claim at the true extremes of the parameter space. Two things still stand
between this and production, and neither is about the solver:

The Gaussian-mismatch risk demonstrated below is not something a better detector
or a better solver fixes. It is a property of whichever real time period the
estimate is built over, and needs its own check before any 2D-corrected estimate
is trusted, regardless of how good the solver itself is.

The performance cost is real and matters more in production than in a vignette:
every evaluation here costs several hundred `pDetGivenNL` calls worth of work,
repeated inside `optim`'s search, versus one bracketed `uniroot` search over a
single pre-built curve for the current 1D method. If `cde()` calls
`nlFromDetections` inside a bootstrap or Monte Carlo loop, that cost is paid on
every replicate. The residual sweep above exists precisely to show this cost is
only worth paying in specific corners of parameter space -- an opt-in path,
gated by that same sweep's criterion, protects the common case's speed while
still being available where the 1D residual is actually large.

# What happens when this meets real seasonal data

The validation above used synthetic cells with a known, exactly Gaussian
`nlTrue`/`nlSd` behind them. Real detections carry no such guarantee. The table
below is real per-season data from Kerguelen2015 (@castro2024, Beyond Counting
Calls); `NLmean`/`NLsd` are exactly the *observed*, biased moments
`nlFromDetections` receives as input in practice -- the same
`meanObs`/`sigmaObs` this vignette has been computing throughout, just from real
detections instead of a synthetic distribution.

```{r seasonal.table}

seasonalNL <- data.frame(
  season = c("summer", "autumn", "winter", "spring", "year"),
  nc     = c(13248, 19123, 32631, 12903, 77905),
  f      = c(0.464, 0.243, 0.265, 0.388, 0.314),
  T      = c(2075.4, 2199.7, 2197.3, 2167.8, 8640.2),
  pa     = c(0.0658, 0.0192, 0.0701, 0.1836, 0.0789),
  NLmean = c(92.4, 96.8, 87.5, 81.0, 88.6),
  NLsd   = c(2.6, 3.7, 3.5, 2.4, 5.9)
)

seasonalNL

```

**This still uses the vignette's placeholder detector, not the actual fitted
detection function for Kerguelen2015** -- illustrative of how the solver behaves
on real-shaped moments, not corrected values for the manuscript.

`sdGrid`'s 1 to 8 dB range already comfortably covers every `NLsd` value in this
table, and the validation above already showed the solver recovers the truth
reliably across that whole range whenever the truth is actually Gaussian, so
there is no separate sdGrid-coverage question left to answer here. What is worth
looking at is `year`, since it is the row most likely to expose what happens
when the Gaussian assumption itself is wrong; the four individual seasons would
mostly reconfirm what the synthetic sweep already established. `year`'s fit was
already computed alongside the validation grid above, in the same parallel
batch, rather than run again here.

The other two sweep axes have a rougher real-world check too. `nlTrueGrid` (74
to 104 dB) comfortably brackets Kerguelen2015's observed seasonal means (81.0 to
96.8 dB, from the table above), so the sweep's range of `nlTrue` was not an
arbitrary choice, even though it was chosen before seeing this data. 

`detScaleGrid` has no equivalent check available. Nothing in this vignette ties
`scale = 0.5` to `8` against the steepness of any detector actually fitted to
real Kerguelen2015 detections, so whether the tested range brackets a realistic
detector is unverified. **TODO:** both BCC and Common Ground fit
data-dependent detection functions (SCAM for BCC; SCAM and VGLM for Common
Ground) rather than a plain logistic, so there is no `scale` parameter sitting
in either model to compare against directly. A pseudo-scale is derivable: for a
standard logistic GLM, `plogis(SNR, location, scale)` matches `logit(p) = β0 +
β1·SNR` via `scale = 1/β1`, `location = -β0/β1`. Refitting a plain logistic on
top of the real data gives one number, but throws away whatever shape SCAM/VGLM
were fitted to capture in the first place -- a local version (finite-difference
slope of `qlogis(predict(fit, ...))` over whatever SNR range real detections
concentrate in, the same diagnostic already used in "Looking at the curve
directly") would be more faithful to the real detector's actual steepness where
it matters, at the cost of a curve instead of a single number. Not attempted
here.

The propagation side is anchored more directly: `TLpyram` is computed for this
exact site (Kerguelen2015, summer). It is reused unchanged for all four seasons
in the comparison below, even though the sound speed profile -- and therefore TL
itself -- genuinely varies by season, the same limitation already flagged as
"worth doing, not done" earlier in this vignette.

```{r seasonal.year}

data.frame(
  NLmean      = 88.6,
  muTrue      = yearFit["muTrue"],
  meanBias    = yearFit["muTrue"] - 88.6,
  NLsd        = 5.9,
  sigmaTrue   = yearFit["sigmaTrue"],
  sdBias      = yearFit["sigmaTrue"] - 5.9,
  convergence = yearFit["convergence"],
  fitRMSE     = sqrt(yearFit["objective"] / 2)
)

```

`year`'s recovered mean sits about 10.9 dB above the observed mean, and its
recovered sd (about 7.2 dB) is nearly double the largest individual season's sd
reported in the table above (3.7 dB, autumn). Both numbers came from a solver
just shown, on synthetic data, to recover the truth to within about 1e-5 dB.
That combination is the point: the solver is not the source of this discrepancy,
so whatever is left has to be a property of the real data it was asked to fit a
single Gaussian to.

`fitRMSE` cannot catch this, and not because it happens to be small here by
chance. `recoverTrueMoments` has exactly two free parameters matching exactly
two targets -- observed mean and observed sd -- so for essentially any finite
pair of target moments, some `(mu, sigma)` exists that reproduces them almost
exactly. `year`'s `fitRMSE` (about 1.1e-6) is barely larger than what a
well-behaved season's `fitRMSE` would look like, despite `year` being the case
there is every reason to distrust. `fitRMSE` checks whether the solver found *a*
fixed point, not whether fitting a single Gaussian was the right thing to do --
with only two moments feeding two parameters, it has no way to see that.

**Hypothesis, not an established mechanism:** `year` aggregates four seasons
whose own observed means span roughly 81 to 97 dB. A single Gaussian forced to
stand in for a mixture of differently-centred Gaussians would generally need to
inflate its own variance to absorb the *between-season* spread on top of each
season's *within-season* variance -- the same decomposition behind the law of
total variance. `year`'s recovered sd coming out well above every individual
season's is consistent with that story, and it would be a more direct signature
of this failure mode than anything `fitRMSE` can offer. But this vignette has
shown one data point, not a validated diagnostic: nothing here has checked that
this pattern reliably shows up for genuine non-Gaussian mixtures in general, or
that it stays quiet for nested windows that really are Gaussian throughout.
Treat it as a plausible explanation worth testing, not a mechanism this vignette
has established.

If the hypothesis holds up, it would point to a real constraint on `cde`'s `T`,
on top of the ones already known: choosing a time period over which to treat TL
and detection probability as constant would not be enough by itself, since the
noise level distribution within that same period would also need to be close
enough to Gaussian for `nlFromDetections`, 1D or 2D, to mean what it claims to
mean. `year` looks like a `T` chosen too coarse for that. Whether the individual
seasons are themselves coarse enough to show some of the same problem in
miniature is a separate open question this table cannot settle by itself --
their own biases (0.8 to 4.1 dB) are compatible with some residual
non-Gaussianity, not only with the small solver/detector residual the synthetic
validation already bounds. Checking whether recovered sd inflates when a
candidate `T` is built by aggregating finer sub-periods is a plausible
diagnostic for that question, contingent on the hypothesis above actually
validating.

Non-Gaussian and non-parametric noise level distributions are out of scope for
this vignette. Every function here -- `pDetGivenNL`, `predictSampledNL`,
`nlFromDetections`, and the exploratory 2D solver above -- assumes a Gaussian
`fN`. Real seasonal NL distributions from Kerguelen2015 are available and not
obviously Gaussian by eye (winter in particular looks bimodal, and `year` looks
like a clear mixture, consistent with the hypothesis above), which would make
them a natural next step if this ever gets pursued further, but fitting other
parametric families or a non-parametric density is a substantially harder
problem than anything solved here, and deliberately not attempted in this
vignette.


# What is still open

`pDetGivenNL`, `predictSampledNL` and `nlFromDetections` are all in the package
now, and `cde` calls `nlFromDetections` by default whenever `NL` is not
supplied, so `nlFromSnrInfo` is a deprecated wrapper rather than the active
path. Two things noted while building this are still genuinely open.

`pDetGivenNL` still duplicates work that `pDetInArea` already does. Both thin
the transmission loss table, broadcast the truncation distance across transects,
mask cells beyond truncation, and form area weights. That preparation should be
pulled out of `pDetInArea` into a shared helper and used by both. Doing it any
other way means the two drift apart.

The standard deviation is taken straight from the detections, and "When does the
correction still hold?" above is the sweep that decides whether that matters. A
first attempt at solving for the mean and standard deviation jointly now exists
("Does the 2D solver actually recover the truth?" and "What happens when this
meets real seasonal data?" above), built with `optim` rather than a proper
bracketed 2D root-finder. Validated against synthetic ground truth, it recovers
the truth to within about 1e-5 dB -- detector data was never the limiting
factor. Real seasonal data exposed a different, more fundamental problem
instead: the correction assumes the true NL distribution is Gaussian, and
whether that holds depends on how the time period `T` was chosen, not on the
solver or the detector. Promoting this into the package, and deciding whether it
becomes a default or an opt-in path, should wait on that question having its own
answer, not just on fixing the solver's remaining robustness gaps.

# What this does not fix

Four assumptions run through every plot above and none is touched by any of
this.

Calls are assumed spread evenly in two dimensions. The area weight in
`pDetGivenNL: r * (r <= td)` is that assumption written down. Clustered calling,
or calls concentrated at particular depths, changes it.

Noise is assumed independent of whether whales are calling. If loud hours are
also whale hours, the detection curve is not the only thing shaping curve 3, and
the correction is solving the wrong problem accurately.

Every detector here is assumed to be a function of SNR alone, `detFun(SNR)`, the
way `pDetGivenNL` is built. That is a real restriction, not just a
simplification for this vignette: it holds for detectors that threshold on
relative level, but plenty of real detectors don't work that way. Koogu's CNN,
for instance, scores a normalised spectrogram window on shape, not on relative
intensity, so its output is substantially decoupled from SNR. Nothing here
extends to a detector like that; doing so would mean characterising
`pDetGivenNL`-equivalent behaviour directly against RL and NL rather than
against their difference, which is a different function to build, not a
different argument to this one.

NL, RL and SNR are all assumed to be measured precisely and accurately. In
reality the best available is an estimate of each, and each will have an error
of its own. For example, the examples in the documentation of SNR estimation
software [`bsnr`](https://github.com/BrianSMiller/bsnr) already documents
several of the ways that SNR and NL estimation goes wrong (detection box size,
background/signal window choice, bandwidth trimming, etc). Whether that error is
small enough to ignore, or large enough to flow through detection function
fitting, the NL distribution estimated here, `p_a`, and ultimately bias in `Dc`,
is an open question this vignette has not attempted to address (but would be
worth adding to the list of related further work).

# References
