---
title: "callDensity: Estimating noise levels from detections"
output: 
  html_document: default
  rmarkdown::html_vignette:
    vignette: >
      %\VignetteEncoding{UTF-8}
      %\VignetteIndexEntry{callDensity: Estimating noise levels from detections}
      %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, fig.width = 6, fig.height = 3.5,
                      comment = "#>")
```

```{r setup}
library(callDensity)
library(ggplot2)
library(kableExtra)
```

# The problem in one paragraph

Call density needs a distribution of noise levels. The obvious place to get one
is the noise measured at each detection, since those measurements are already
in the capture history table. Those measurements are biased low. A quiet period
gives you a high SNR, a high SNR gives you a detection, so your detections
oversample the quiet periods. Measure noise only where you detected something
and you will conclude the ocean is quieter than it is.

This vignette shows how big that bias is, where it comes from, and how to undo
it.

# The one idea

`pDetInArea` already answers a question very close to the one we need. It asks:
given a distribution of noise levels, what fraction of the calls in the study
area get detected? That fraction is $\hat{p}_a$. It draws noise levels from a
normal distribution and averages over them.

Ask a slightly different question. **What would $\hat{p}_a$ be if the noise were
always exactly 84 dB?** Not drawn from a distribution. Fixed.

Same machinery, one line different. Call the answer `pDetGivenNL(84)`. Compute
it for 70 dB, 71 dB, 72 dB and so on, and you have a curve: probability of
detection against noise level.

That curve is the whole solution. It is not a new concept. It is $\hat{p}_a$
with the noise held still.

# Building the curve

`pDetGivenNL` is not in the package yet. Here it is as local code so that this
section runs today.

```{r pdetgivennl}

#' Probability of detecting a call somewhere in the study area, given a fixed
#' noise level.
#'
#' @param nl      Vector of noise levels (dB) to evaluate.
#' @param detFun  Either a detFun object, or a plain function of SNR returning
#'                probability.
#' @param TL      Transmission loss table. First column is range in m, the
#'                remaining columns are TL in dB for each radial transect.
#' @param SL      data.frame with mean and sd of the source level distribution.
#' @param truncationDistance Scalar, or one value per transect.
#' @param nSLnodes Number of quadrature nodes for the source level integral.
#' @return Numeric vector, same length as nl.
pDetGivenNL <- function(nl, detFun, TL, SL,
                        truncationDistance = max(TL[, 1]),
                        nSLnodes = 41) {

  range_m <- TL[, 1]
  tlMat   <- as.matrix(TL[, -1, drop = FALSE])

  # Area weights. Each range step stands for an annulus, and the area of an
  # annulus grows with r, so a step at 200 km carries ten times the weight of
  # a step at 20 km. Cells beyond the truncation distance carry no weight.
  truncationDistance <- rep_len(truncationDistance, ncol(tlMat))
  wMat <- outer(range_m, truncationDistance, function(r, td) r * (r <= td))
  totalWeight <- sum(wMat)

  # Source levels. Calls are not all equally loud, so average over the SL
  # distribution using a weighted grid.
  slNodes <- seq(SL$mean - 4 * SL$sd, SL$mean + 4 * SL$sd, length.out = nSLnodes)
  slWt    <- dnorm(slNodes, SL$mean, SL$sd)
  slWt    <- slWt / sum(slWt)

  # The detection function depends on one number, SNR. So evaluate it once on a
  # grid and interpolate, rather than calling predict() millions of times.
  pLookup <- if (is.function(detFun)) {
    detFun
  } else {
    snrGrid <- seq(-100, 100, by = 0.1)
    pGrid   <- predictDetFun(detFun, newdata = data.frame(SNR = snrGrid))$fit
    approxfun(snrGrid, pGrid, rule = 2)
  }

  vapply(nl, function(thisNL) {
    acc <- 0
    for (i in seq_along(slNodes)) {
      snr <- slNodes[i] - tlMat - thisNL      # sonar equation, every cell
      acc <- acc + slWt[i] * sum(pLookup(snr) * wMat)
    }
    acc / totalWeight
  }, numeric(1))
}

```

Now build a curve. The transmission loss table and source levels follow the
second vignette, and the detector is a plain logistic so this section does not
depend on anything else.

```{r build.curve}

R  <- 1e6
SL <- data.frame(mean = 190, sd = 4, sampleSize = 350)

TL <- simTLradials_20logR(maxRange = R, rangeStep = 1000, numTransects = 4)
TL <- TL[is.finite(rowSums(TL)) & TL[, 1] > 0, ]     # drop r = 0 if present

detector <- function(snr) plogis(snr, location = 1, scale = 2)

nlGrid <- seq(from = 55, to = 135, by = 1)
pCurve <- pDetGivenNL(nlGrid, detector, TL, SL)

curveDf <- data.frame(nl = nlGrid, pDet = pCurve)

```

## The graph that explains everything

```{r curve.plot, warning=FALSE, fig.height=4}

# A reference line falling exactly one decade per 10 dB, anchored on the curve
# at 90 dB.
anchor <- curveDf$pDet[curveDf$nl == 90]
refDf  <- data.frame(nl = nlGrid, pDet = anchor * 10^(-(nlGrid - 90) / 10))

ggplot(curveDf, aes(x = nl, y = pDet)) +
  geom_line(data = refDf, linetype = "dashed", colour = "grey50",
            linewidth = 0.6) +
  geom_line(linewidth = 0.9, colour = "steelblue") +
  geom_vline(xintercept = 84, colour = "firebrick", linetype = "dotted") +
  annotate("text", x = 84.5, y = 1e-4, label = "true mean NL",
           colour = "firebrick", hjust = 0, size = 3) +
  scale_y_log10() +
  xlab("Noise level (dB)") +
  ylab("Probability of detection in area") +
  labs(caption = "Dashed grey: one decade per 10 dB")

```

Read this plot slowly, because the rest of the vignette is a consequence of it.

**The straight portion.** Over most of its range the curve falls by exactly one
decade for every 10 dB of noise, matching the dashed line. That is geometry and
nothing else. Raise the noise 10 dB and every call needs 10 dB more received
level to be heard. Spherical spreading means 10 dB of received level is a
factor of 3.16 in range, which is a factor of 10 in area. Ten times less ocean,
ten times fewer detections. It does not matter what the detector is: a greedy
detector and a fussy detector both lose 90% of their calls, because what
changed is how much ocean is left, not how good the detector is.

This is the picture behind the algebra. On a log axis, "detections fall
tenfold per 10 dB" is a straight line of slope $-1$ per decade. Nothing more.

**The shoulder on the left.** Below about 65 dB the curve flattens and stops
falling. There is no more ocean to gain. Every call inside the study area is
already being detected, so making the sea quieter buys you nothing.

That shoulder is important. It is exactly where the tidy algebra stops working,
and it is why this vignette computes a curve instead of fitting a slope.

**Where your data sits.** The red line marks the true mean noise level. It sits
on the straight portion, which is the interesting case: noise variation there
translates directly into detection variation.

# Why your measurements are biased

Now use the curve. Nature draws a noise level for each call from a normal
distribution. The detection process then throws most of them away, and it
throws away loud ones far more eagerly than quiet ones. The curve says exactly
how much more eagerly.

```{r filter.plot, warning=FALSE, fig.height=4}

nlTrue  <- 84
nlSd    <- 4

fine  <- seq(nlTrue - 5 * nlSd, nlTrue + 5 * nlSd, length.out = 400)
fN    <- dnorm(fine, nlTrue, nlSd)
pd    <- pDetGivenNL(fine, detector, TL, SL)
wt    <- fN * pd

# Scale each curve to a common height so the shift is visible
filterDf <- rbind(
  data.frame(nl = fine, y = fN / max(fN),  what = "1. All calls (the truth)"),
  data.frame(nl = fine, y = pd / max(pd),  what = "2. Chance of detection"),
  data.frame(nl = fine, y = wt / max(wt),  what = "3. What you measure")
)

meanObs <- sum(fine * wt) / sum(wt)

ggplot(filterDf, aes(x = nl, y = y, colour = what)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = nlTrue,  colour = "grey30", linetype = "dotted") +
  geom_vline(xintercept = meanObs, colour = "grey30", linetype = "dashed") +
  annotate("segment", x = meanObs, xend = nlTrue, y = 1.05, yend = 1.05,
           arrow = arrow(length = unit(2, "mm"), ends = "both")) +
  annotate("text", x = (meanObs + nlTrue) / 2, y = 1.12,
           label = sprintf("bias = %.2f dB", nlTrue - meanObs), size = 3) +
  xlab("Noise level (dB)") +
  ylab("Scaled to a common height") +
  labs(colour = NULL) +
  theme(legend.position = "top")

```

Curve 1 is the truth. Curve 2 is the detection curve from the previous plot,
now drawn on a linear axis over the range the noise actually occupies. Curve 3
is curve 1 multiplied by curve 2, and it is what your detections show you.

Multiplying by a curve that leans left drags the peak left. That is the entire
bias. Nothing subtle is happening.

Two things worth noticing, because both get used later.

Curve 3 is narrower than curve 1 by very little. The width survives the
filtering almost intact, which means the standard deviation of your measured
noise is close to the truth even though the mean is not. That leaves only one
unknown to solve for.

The size of the shift depends on how steeply curve 2 leans, and on how wide
curve 1 is. It does not depend on where the detector sits. Move the detector
threshold and curve 2 changes height, not slope.

# Undoing it

The previous plot is a prediction machine. Give it a candidate true mean, and
it tells you what mean you should have measured.

You know what you measured. So run it backwards.

```{r predict.function}

#' Predict the mean noise level you would measure at detections, given a
#' candidate true noise distribution.
predictSampledNL <- function(mu, sigma, detFun, TL, SL,
                             truncationDistance = max(TL[, 1])) {
  nl <- seq(mu - 6 * sigma, mu + 6 * sigma, length.out = 121)
  wt <- dnorm(nl, mu, sigma) *
        pDetGivenNL(nl, detFun, TL, SL, truncationDistance)
  sum(nl * wt) / sum(wt)
}

```

```{r rootfind.plot, warning=FALSE, fig.height=3.5}

muTry <- seq(80, 92, by = 1)
muObs <- vapply(muTry, function(m)
  predictSampledNL(m, nlSd, detector, TL, SL), numeric(1))

rootDf <- data.frame(muTrue = muTry, muMeasured = muObs)

ggplot(rootDf, aes(x = muTrue, y = muMeasured)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey70", linetype = "dashed") +
  geom_line(linewidth = 0.9, colour = "steelblue") +
  geom_hline(yintercept = meanObs, colour = "firebrick", linetype = "dotted") +
  geom_vline(xintercept = nlTrue,  colour = "firebrick", linetype = "dotted") +
  annotate("text", x = 80.2, y = meanObs + 0.3, hjust = 0, size = 3,
           colour = "firebrick", label = "what you measured") +
  xlab("Candidate true mean NL (dB)") +
  ylab("Mean NL you would measure at detections") +
  labs(caption = "Dashed grey: the 1:1 line you would get with no bias")

```

The blue line sits below the 1:1 line by the size of the bias. It rises
steadily, so there is exactly one candidate that reproduces your measurement.
Find it by reading across from the red horizontal line to the blue curve, then
down. That is what `uniroot` does.

```{r nlfromdetections}

#' Estimate the true noise level distribution from noise measured at detections.
#'
#' PLACEHOLDER. This will become nlFromDetections() in the package, replacing
#' nlFromSnrInfo(). See "What is left to build" below.
nlFromDetections <- function(snrInfo, detFun, TL, SL,
                             truncationDistance = max(TL[, 1]),
                             searchWidth = 25) {

  obsMean <- mean(snrInfo$NoiseRL, na.rm = TRUE)
  sigma   <- sd(snrInfo$NoiseRL,   na.rm = TRUE)

  # The bias is always positive, so the true mean is above the measured one.
  gap <- function(mu) {
    predictSampledNL(mu, sigma, detFun, TL, SL, truncationDistance) - obsMean
  }
  mu <- uniroot(gap, interval = c(obsMean, obsMean + searchWidth))$root

  data.frame(mean = mu, sd = sigma,
             sampleSize = sum(!is.na(snrInfo$NoiseRL)))
}

```

# Where this leaves the old function

`nlFromSnrInfo` corrects the same bias by adding the SNR at which the detection
function reaches 0.5. That quantity is a property of the detector. The plots
above say the bias is a property of the geometry and the noise variance. The
two happen to be close for the parameters of the second vignette and they part
company elsewhere.

```{r sweep.placeholder, eval=FALSE}

# PLACEHOLDER - see callDensity3 scratchpad, to be folded in here.
#
# Two sweeps, both at Casey2019 scale (214 h subsample, every 8th detection
# adjudicated):
#
#   A. Move the detector threshold from -5 to +10 dB SNR. The bias needed
#      stays near 3 dB throughout while the old correction tracks the
#      threshold exactly.
#
#   B. Vary sd(NL) from 1 to 10 dB. The bias grows with the square of sd
#      while the old correction does not move.
#
# Add nlFromDetections as a third column to both. Expect it to sit on the
# bias-needed line in both sweeps.

```

# Does it depend on the transmission loss model?

No, and this is the point of computing a curve rather than fitting a slope.
`pDetGivenNL` reads transmission loss out of a lookup table. It never writes
down $20\log_{10}r$. The dashed reference line on the first plot is a
diagnostic, not an assumption.

```{r tl.placeholder, eval=FALSE}

# PLACEHOLDER.
#
# Rebuild the first plot with a transmission loss table that is not spherical
# spreading. Candidates, in increasing order of realism:
#
#   1. Mixed spherical to cylindrical: 20*log10(rt) + 10*log10(r/rt)
#   2. A pyRAM table from 2026-CopernicusPyRAM, Kerguelen2015 at 26 Hz
#
# Expect the straight portion to change slope for (1), and to become bumpy for
# (2) where convergence zones add and remove area. Expect nlFromDetections to
# recover the true mean in both cases, because it uses the curve rather than
# its slope.

```

# What is left to build

Everything above runs, but three of the pieces live in this file rather than in
the package.

`pDetGivenNL` duplicates work that `pDetInArea` already does. Both thin the
transmission loss table, broadcast the truncation distance across transects,
mask cells beyond truncation, and form area weights. That preparation should be
pulled out of `pDetInArea` into a shared helper and used by both. Doing it any
other way means the two drift apart.

`nlFromDetections` needs to replace the `is.null(NL)` branch inside `cde`, which
currently calls `nlFromSnrInfo`. Everything it needs is already an argument to
`cde`. `nlFromSnrInfo` then becomes a deprecated wrapper, following the same
pattern as `fitSNRdetectionFunc`.

The standard deviation is taken straight from the detections. The second plot
shows why that is nearly right and not exactly right: filtering narrows curve 3
a little. Solving for the mean and the standard deviation together would remove
that, at the cost of turning a one dimensional root find into a two dimensional
optimisation. Worth measuring the residual first and only fixing it if it
matters.

# What this does not fix

Two assumptions run through every plot above and neither is touched by any of
this.

Calls are assumed spread evenly in two dimensions. The area weight
 in `pDetGivenNL` is that assumption written down. Clustered
calling, or calls concentrated at particular depths, changes it.

Noise is assumed independent of whether whales are calling. If loud hours are
also whale hours, the detection curve is not the only thing shaping curve 3,
and the correction is solving the wrong problem accurately.

# References
