---
title: "callDensity_snrThreshold"
output: 
  html_document: default
  pdf_document: default
  rmarkdown::html_vignette:
    vignette: >
      %\VignetteEncoding{UTF-8}
      %\VignetteIndexEntry{callDensity_snrThreshold}
      %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.height = 3.75, 
  comment = "#>"
)
```

```{r setup}
library(callDensity)
library(ggplot2)
library(scales)
library(kableExtra)
```

# Further exploration and topics in callDensity package (TODO)

1)  Spatial distribution of calls
    A)  Effect of land, ice, and other zero-probability areas
    B)  Effect of non-uniform distribution of calls in area
2)  **Truncation in SNR**
3)  **Detectors with different slopes**
4)  Minimum number of recaptures/effort required to get a reliable estimate
5)  More realistic Transmission Loss models and effects of mismatch

As usual, we start with the density equation:

\begin{equation}
D_c = \frac{N_c(1-c)}{kTp_aA}
\end{equation}

where: $D_c$ is call density, $N_c$ is number of calls, $c$ is false discovery rate, $k$ is number of sensors (here always 1), $T$ is the duration of data analysed (in hours), $p_a$ is the probability of detection in the study area, and $A$ is the study area (in km\^2).

## Effects of truncation in SNR

Scenario: Antarctic blue whales in a quiet ocean with spherical spreading of sounds. 

In this example, we investigate the effects of excluding detections with an SNR that are below a threshold value.  

```{r call.spatial.distribution, warning=FALSE, message=FALSE}

n = 1e6; # number of simulated calls (regardless of whether detected or not); 
R = 1e6; # radius of 1000 km (but in m)
k <- 1   # number of sensors
minDate <- as.POSIXct("2025-01-01")
maxDate <- as.POSIXct("2026-01-01")
Time <- as.numeric(difftime(maxDate,minDate,unit="days"))/365 # in years

# Call density, D_c is n/A 
A = studyArea(R/1e3) # Circular study area in km^2
TrueCallDensity = n/(A*Time) # calls/km^2/time

# 1) Generate n uniformly distributed calls within radius R and time period Time
sim <- simCallLocation(n=n, R=R, minDate=minDate, maxDate=maxDate)

```
Sonar equation parameters for this scenario:

\begin{equation}
TL = 20  log_{10}  d
\end{equation}
\begin{equation}
SNR = SL-TL-NL
\end{equation}


```{r sonar.eq}

SL <- data.frame(mean=190, sd=4, sampleSize=350) # True SL distribution for sim
NL = data.frame(mean=84, sd = 4, sampleSize = n)  # True NL distribution for sim
tlFunc <- function(r)20*log10(r)

# Simulate the true acoustic properties of each call and false positive 
sim <- simCallAcoustics(sim, SL, NL, TL = tlFunc)

# Per-transect estimate of TL (used by callDensity package to estimate pDet)
TL <- simTLradials_20logR(maxRange=R, rangeStep=100, numTransects=8)

```

### Simulate detectors:

```{r parameterise.detectors, warning=FALSE, message=FALSE}

#Specify parameters for detector 1
# Good detector with low threshold (location) and low false positive rate.
det1params = data.frame( 
  location=1,    # AKA intercept?
  scale=1,       # AKA slope?
  func='plogis', # logistic function
  c=0.1,         # False discovery rate
  fpMean=0,      # Mean of distribution of false positives (in dB SNR)
  fpSD = 2      # Standard deviation false positive distribution (in dB SNR)
)

# Detector 2 has a higher threshold (location), scale, and false positive rate
# than detector 1.
det2params = data.frame( 
  location=3,    # AKA intercept?
  scale=0.5,       # AKA slope?
  func='plogis', # logistic function
  c=0.3,         # False discovery rate
  fpMean=0,      # Mean of distribution of false positives (in dB SNR)
  fpSD = 2      # Standard deviation false positive distribution (in dB SNR)
)
```

Now we create a function that will take the detector parameters above and simulate the detection process. We then call this function with each set of parameters to create a simulation for each detector.

```{r simulate.detectors, warning=FALSE, message=FALSE}

# Simulate two detectors each with the different parameters
simDet1<- simulateDetector(det1params,sim)
simDet2<- simulateDetector(det2params,sim)

```

## SNR Truncation

Apply SNR truncation to the detectors by removing all detections that were less than the threshold.

```{r}
snrTruncationThreshold <- 0; # dB

# True number of calls above threshold
nAboveThreshold <- sum(sim$snr>snrTruncationThreshold) 

# Variables from sims that were truncated in SNR get a 't' as a suffix
simDet1t <- subset(simDet1, simDet1$snr >= snrTruncationThreshold)
simDet2t <- subset(simDet2, simDet2$snr >= snrTruncationThreshold)

# Number of detections above threshold for each detector
Nc1 <- sum(simDet1t$detect_table)
Nc2 <- sum(simDet2t$detect_table)

```

View the spatial detection density (which is not the same as spatial call density since it  accounts for neither detection probability nor time).

```{r detector.plots, warning=FALSE, message=FALSE}

p1  <- plotSpatialDetections(simDet1)
p2  <- plotSpatialDetections(simDet2)
p1t <- plotSpatialDetections(simDet1t)+ylab("Truncated\nY (kM)")
p2t <- plotSpatialDetections(simDet2t)+ylab("Truncated\nY (kM)")

gridExtra::grid.arrange(p1,p2,p1t,p2t,nrow=2)

dist1 <- plotDetectionDistribution(simDet1)
dist2 <- plotDetectionDistribution(simDet2)
dist1t<- plotDetectionDistribution(simDet1t)+ylab("Truncated count")
dist2t<- plotDetectionDistribution(simDet2t)+ylab("Truncated count")
gridExtra::grid.arrange(dist1,dist2,dist1t,dist2t,nrow=1)

```

## Subsample data for detector characterisation

@castro2024 subsample approximately 200 hours evenly spaced throughout the year to create detector characterisation curves (detection vs SNR). Here we subsample 1 hour in every 41 hours yielding 214 subsamples that are 1 hour in duration.

```{r subsample.for.detector.characterisation}

subsampleDet1 <- subsampleSimInTime(simDet1,interval = "23 hour")
subsampleDet2 <- subsampleSimInTime(simDet2,interval = "23 hour")

subsampleDet1t <- subsampleSimInTime(simDet1t,interval = "23 hour")
subsampleDet2t <- subsampleSimInTime(simDet2t,interval = "23 hour")

print(paste("True positives in original  subsample of detector 1: ", 
            with(subsampleDet1,sum(groundTruth & detect_table)) ) )
print(paste("True positives in truncated subsample of detector 1: ", 
            with(subsampleDet1t,sum(groundTruth & detect_table)) ) )

print(paste("True positives in original  subsample of detector 2: ", 
            with(subsampleDet2,sum(groundTruth & detect_table)) ) )
print(paste("True positives in truncated subsample of detector 2: ", 
            with(subsampleDet2t,sum(groundTruth & detect_table)) ) )

```

## Detection matching between detector 1 and 2

Both detectors (simulated subsets) operate on the same set of calls (i.e. they are different subsets of the same underlying simulation). So we can find detections from one detector that match those of the other detector by merging the detections into a capture history table.

Presently, false positives have randomly generated date-times, so have a negligible chance of matching between the two detectors.

*TODO: Create false positives in a way that better approximates their real-world occurrence (i.e. in a way that allows them to be matched between detectors).*

```{r match detections from detector1 and detector2}

# Create a capture history table from the two tables of detections (All
# simulated calls in the subset whether detected or not)
capHistTab<- simsTocaptureHistoryTable(subsampleDet1,subsampleDet2)
capHistTabt<- simsTocaptureHistoryTable(subsampleDet1t,subsampleDet2t)

# Total number of true positive detections for detector1 OR detector2
nTrueDetectedSubset <- with(capHistTab,sum(detect_table1 & groundTruth1 |   
                                        detect_table2 & groundTruth2) )
nTrueDetectedSubsett <- with(capHistTabt,sum(detect_table1 & groundTruth1 |   
                                        detect_table2 & groundTruth2) )

# Total number of positive detections for detector1 OR detector2
nPositiveSubset <- with(capHistTab,sum(detect_table1 | detect_table2))
nPositiveSubsett <- with(capHistTabt,sum(detect_table1 | detect_table2))

```

Convert this capture history table into an SNRinfo table, and treat detector1 as ground truth.

```{r format.output.for.callDensity.package}

# Convert capture history into SNRinfo format (removing detections )
SNRinfo <- capHistTosnrInfo(capHistTab)
SNRinfot<- capHistTosnrInfo(capHistTabt)

# Detector 2 true & false positives
n_p_subsampleDet2 <- sum(capHistTab$detect_table2)
n_p_subsampleDet2t <- sum(capHistTabt$detect_table2) 

# Detectir 2 false positives (Positive on detector 2, but not on 1)
n_fp_subsample <- sum(capHistTab$detect_table2 & !capHistTab$detect_table1)
n_fp_subsamplet <- sum(capHistTabt$detect_table2 & !capHistTabt$detect_table1)

# False discovery rate is number.false.positives/number.predicted.positive
c_subsample <- n_fp_subsample/n_p_subsampleDet2
c_subsamplet<- n_fp_subsamplet/n_p_subsampleDet2t

n_subsample <- dim(SNRinfo)[1]  # Number positive detections in SUBset
n_subsamplet <- dim(SNRinfot)[1]  # Number positive detections in SUBset

truncationDistances <- R 

# Number positive detections in FULL set
Nc <- sum(simDet2$detect_table)
Nct <- sum(simDet2t$detect_table)       

```

## Calculate call densities
### SNR Truncation ONLY for SNR Detection Function

First we calculate call density using the original subset (without SNR truncation). These results are given the suffix .glm. 

We compare the call density and probability of detection in the area of original subset to those derived using SNR truncatiot but only truncating in SNR to derive the SNR-detection function, and NOT applying SNR truncation to the Monte-Carlo model nor to derivation of false positive rate, c.  

```{r GLM p.det with independently estimated NL, warning=FALSE, message=FALSE}

snrDetFun.glm <- callDensity::fitSNRdetectionFunc(SNRinfo,  modelType = "glm")
snrDetFun.glmt<- callDensity::fitSNRdetectionFunc(SNRinfot, modelType = "glm")

snrDetFun.glm$r.sq <- with(summary(snrDetFun.glm),  1 - deviance/null.deviance)
snrDetFun.glmt$r.sq<- with(summary(snrDetFun.glmt), 1 - deviance/null.deviance)

# Estimate of NL distribution. NOTE: Noise levels estimated from mean NL of
# non-truncated detections plus intercept of SNR detection function (truncated
# or not).
NLind <- nlFromSnrInfo(SNRinfo,  snrDetFun.glm)
NLindt<- nlFromSnrInfo(SNRinfo, snrDetFun.glmt)

Pa.glm <- callDensity::pDetInArea(snrDetFun.glm, SL=SL, TLlookup = TL, NL=NLind,
           output.resolution.m=100, outerloop = 100, truncationDistance=R, # m
           snrTruncationThreshold =-Inf)
pa.glm <- Pa.glm$overall

Pa.glmt <- callDensity::pDetInArea(snrDetFun.glmt, SL=SL, TLlookup = TL, NL=NLindt,
           output.resolution.m=100, outerloop = 100, truncationDistance=R, # m
           snrTruncationThreshold =-Inf)
pa.glmt <- Pa.glmt$overall


# Density estimation equation
Dc.glm <- (Nc * (1 - c_subsample) )/( k * A * pa.glm * Time )
Dc.glmt <- (Nc * (1 - c_subsample) )/( k * A * pa.glmt * Time )


results.glm <- data.frame(Dc=Dc.glm, SampleSize=n_subsample, A = A,
                      NLmean = NLind$mean, NLsd = NLind$sd, 
                      model='glm', c=c_subsample, Pa=pa.glm,  
                      row.names = 'All SNR')
results.glmt <- data.frame(Dc=Dc.glmt, SampleSize=n_subsamplet, A = A,# Environment
                      NLmean = NLind$mean, NLsd = NLind$sd,      # Noise
                      model='glm', c=c_subsample, Pa=pa.glmt,  
                      row.names = paste('Truncated',snrTruncationThreshold,'dB'))

```


Compare with the true results

```{r Call.Density.Results, echo=TRUE, warning=FALSE}

resultsTrue = data.frame(Dc = TrueCallDensity, SampleSize = n, A = A,# Environment 
                      NLmean = NL$mean, NLsd = NL$sd,            # Noise
                        # Detector params
                        model=det2params$func, c = det2params$c, 
                        Pa = mean(simDet2$p_det,na.rm=TRUE), 
                      row.names='Truth')

# True density and PA of calls above threshold
Dct <- nAboveThreshold/(A*Time) 
Pat <- mean(simDet2$p_det[simDet2$snr > snrTruncationThreshold ],na.rm=T)

results<-rbind( resultsTrue, results.glm, results.glmt)

kableExtra::kbl(results, digits = c(4,3,0,1, 2, NA, 3, 4), 
                col.names= c('Dc','(sub)Sample Size', 'A','NL_mean','NL_sd',
                             'Model','c','P_a')) %>% 
  kableExtra::kable_classic(full_width=FALSE)


```

In this example, the estimate of call density with SNR truncation is further from the true value than the estimate without. 

## SNR Truncation everywhere (snrDetFun, pDetInArea, c, Nc)

Alternatively, we can apply SNR truncation for every parameter that goes into the call density equation. 

This means applying SNR truncation when tabulating the number of calls, Nc; when estimating the density of calls within the area, pa; as well as applying SNR truncation when estimating the false discovery rate, c. 

Applying SNR truncation in this way will create a result that is not comparable to the above results. This is because we are now estimating the density only of calls above our SNR threshold, whereas previously we were estimating the density of all calls.  

```{r}

# True value of false discovery rate for truncated sim
c_truncated_actual = sum(simDet2t$groundTruth==FALSE)/sum(simDet2t$detect_table)

# True density of calls that are above threshold
resultsTruncated = data.frame(Dc = Dct, SampleSize=nAboveThreshold,
                      A = A, NLmean = NL$mean, NLsd = NL$sd,            # Noise
                      model=det2params$func, 
                      c = c_truncated_actual, Pa = mean(simDet2t$p_det,na.rm=TRUE), 
                      row.names=paste('Truth (>',snrTruncationThreshold,'dB)') ) 

# Probability of detection in teh area of calls that are above threshold (NOT all calls)
Pa.allt <- callDensity::pDetInArea(snrDetFun.glmt, SL=SL, TLlookup = TL, NL=NLindt,
           output.resolution.m=100, outerloop = 100, truncationDistance=R, # m
           snrTruncationThreshold =snrTruncationThreshold)
pa.allt <- Pa.allt$overall
Dc.allt <- (Nct * (1 - c_subsamplet) )/( k * A * pa.allt * Time )

results.allt <- data.frame(Dc=Dc.allt, SampleSize=n_subsamplet, A = A,# Environment
                      NLmean = NLindt$mean, NLsd = NLindt$sd,      # Noise
                      model='glm', c=c_subsamplet, Pa=pa.allt,  
                      row.names = paste('Truncated',snrTruncationThreshold,'dB'))

resultst<-rbind( resultsTruncated, results.allt)

kableExtra::kbl(resultst, digits = c(4,3,0,1, 2, NA,  3, 4), 
                col.names= c('Dc','(sub)Sample Size', 'A','NL_mean','NL_sd',
                             'Model','c','P_a')) %>% 
  kableExtra::kable_classic(full_width=FALSE)

```

Applying truncation throughout all the terms of the call density equation seems to produce a correct estimate of call density. 

However, interpretation of these densities is less satisfying because they are potentially only meaningfully comparable to densities of calls measured using the same thresholds. 

It is not obvious whether these densities of calls above an SNR threshold are still comparable across sites that have different SL, NL, or TL profiles. While I suspect they are comparable, but  will depend on the accuracy of per-site estimates of SL, NL, and TL. Non-truncated call densities also depend on the accuracy of acoustic parameters, but estimates that utilize SNR truncation may be more sensitive to mismatch between parameter and reality due to smaller sample sizes. The exploration of this topic is beyond the scope of this vignette (left as an exercise for an interested party or as an item to add to the TODO list). 

Additionally, it might be possible to alter the Monte-Carlo simulation to keep track of the proportion of calls that were excluded due to SNR truncation. Then, the density of above-threshold calls could be scaled by this proportion to convert the answer back into a non-truncated call density. Implementation of this idea is again beyond the scope of this vignette. 

## VGLM and truncation

Placeholder to demonstrate that adjudicated capture recapture produces same results when truncating in SNR. 

```{r VGLM p.det, warning=FALSE, message=FALSE}
library(VGAM)
### Adjudicated capture recapture (capture history table with false positives
### removed by judge)
adjudicated <- subset(capHistTab,
                (capHistTab$groundTruth1 | capHistTab$groundTruth2) &
                (capHistTab$detect_table1 | capHistTab$detect_table2))
adjudicatedt <- subset(capHistTabt,
                (capHistTabt$groundTruth1 | capHistTabt$groundTruth2) &
                (capHistTabt$detect_table1 | capHistTabt$detect_table2))

# Number of adjudicated positive detections
n_adj <- dim(adjudicatedt)[1]

adjudicatedt$SNR <-   rowMeans(
  subset(adjudicatedt,select=c('snr1','snr2')) ,na.rm=T) #dB

#False discovery rate of truncated
tp <- sum(with(capHistTabt,(detect_table2 & groundTruth2) ))
fp <- sum(with(capHistTabt,(detect_table2 & !groundTruth2) ))
# precision <- sum(tp)/length(tp)
adj_c <- fp/(fp+tp)

observerNames = c("detect_table1", "detect_table2")
snrDetFun.vglm <- fitSNRvglm(adjudicatedt, observerNames,
                             whichObserver = "detect_table2")

# Estimate of NL distribution
noiseTab <- capHistTab[capHistTab$detect_table2 | capHistTab$detect_table1,]
noiseTab$NoiseRL <- rowMeans(
  subset(noiseTab,select=c('noiseRMSdB1','noiseRMSdB2')) ,na.rm=T) #dB
NLadj<- nlFromSnrInfo(noiseTab, snrDetFun.vglm)

Pa.vglm <- callDensity::pDetInArea(snrDetFun.vglm, SL=SL, TLlookup = TL, NL=NLadj,
           output.resolution.m=100,
           outerloop = 10,
           truncationDistance=R, # in m
           snrTruncationThreshold = snrTruncationThreshold)
pa.vglm <- Pa.vglm$overall


# Density estimation equation
Dc.vglm <- (Nct * (1 - adj_c) )/( k * A * pa.vglm * Time )

results.vglm <- data.frame(Dc=Dc.vglm, SampleSize=n_adj, A = A,# Environment
                      NLmean = NLadj$mean, NLsd = NLadj$sd,# Noise
                      model='vglm',  c=adj_c, Pa=pa.vglm,  
                      row.names = 'ACR')

```


```{r}

resultst<-rbind( resultsTruncated, results.allt, results.vglm)

kableExtra::kbl(resultst, digits = c(4,3,0,1, 2, NA,  3, 4), 
                col.names= c('Dc','(sub)Sample Size', 'A','NL_mean','NL_sd',
                             'Model','c','P_a')) %>% 
  kableExtra::kable_classic(full_width=FALSE)


```
The adjudicated capture-recapture model seems to produce a correct estimate of density of above-threshold calls. But something might not be quite right since the values for c seem a bit off. Further investigation of the code is warranted. 


## Inspect detection functions for insights

```{r compare snr detection functions, warning=FALSE}
SNR <- seq(-20,20,0.5);
snrs <- data.frame(SNR)

cols = colorRampPalette(c('#EDB120','#D95319','#0072BD','#7E2F8E','#77AC30',
                        '#4DBEEE','#A2142F'))
preds.glm= predict(snrDetFun.glm, type='response',newdata=snrs)
preds.glmt= predict(snrDetFun.glmt, type='response',newdata=snrs)
preds.vglm=predict(snrDetFun.vglm,type='response',newdata=snrs)
preds.vglm0=predict(snrDetFun.vglm,type='response',newdata=snrs,
                    type.fitted='onempall0')# One minus probability of all zeros
pr =apply(cbind(preds.vglm[,2],preds.vglm0),1,prod)


par(cex=0.8,mar=c(3.1,3.1,0.25,0),mgp=c(2.1,1,0))
plot(snrs$SNR, xlim= c(-5,10),  ylim= c(0,1),
     xlab="SNR", ylab = "P(Detection)", type='n')
grid(nx = NULL, ny = NULL, lty = 2, col = "gray", lwd = 2)
lines(snrs$SNR,preds.glm,col=cols(7)[4],lwd=2)
lines(snrs$SNR,preds.glmt,col=cols(7)[5],lwd=2)
lines(snrs$SNR,pr,col=cols(7)[1],lwd=2)    # vglm 1-p(all zero)
points(simDet2t$snr,simDet2t$p_det,pch='.')     # Actual probability of detection
legend( "bottomright", 
        legend =c("TRUE",
                  "GLM-No SNR Truncation",
                  "GLM-SNR Truncated",
                  "VGLM-SNR Truncated"
                  # "Perception bias: detector1 (VGLM)",
                  # "Perception bias: detector2 (VGLM)",
                  # "1-p(all zeros) (VGLM)"
                  ),
        col=c('black', cols(7)[4], cols(7)[5], 
              cols(7)[1] #, cols(7)[6],  cols(7)[2]), 
        ),
        lty=rep('solid',6),
        lwd=rep(2,6)
)
```


# References
