---
title: "callDensity: Two detector simulation"
output: 
  html_document: default
  pdf_document: default
  rmarkdown::html_vignette:
    vignette: >
      %\VignetteEncoding{UTF-8}
      %\VignetteIndexEntry{callDensity: Two detector simulation}
      %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.height = 3.75, 
  comment = "#>"
)
```

```{r setup}
library(callDensity)
library(ggplot2)
library(scales)
library(kableExtra)
```

# Further exploration and topics in callDensity package (TODO)

1)  Spatial distribution of calls
    A)  Effect of land, ice, and other zero-probability areas
    B)  Effect of non-uniform distribution of calls in area
2)  **Truncation in SNR**
3)  **Detectors with different slopes**
4)  Minimum number of recaptures/effort required to get a reliable estimate
5)  More realistic Transmission Loss models and effects of mismatch

As usual, we start with the density equation:

\begin{equation}
D_c = \frac{N_c(1-c)}{kTp_aA}
\end{equation}

where: $D_c$ is call density, $N_c$ is number of calls, $c$ is false discovery rate, $k$ is number of sensors (here always 1), $T$ is the duration of data analysed (in hours), $p_a$ is the probability of detection in the study area, and $A$ is the study area (in km\^2).

## Effects of truncation in SNR

Scenario: Antarctic blue whales in a quiet ocean with spherical spreading of sounds. 

In this example, we investigate the effects of excluding detections with an SNR that are below a threshold value.  

```{r call.spatial.distribution, warning=FALSE, message=FALSE}

# 1) Generate a distribution of n calls within radius R of the hydrophone with a
# known call density Uniform random in circle

# For ABW ABZ calls, 1 whale call every 10 s would be a maximum of 3.1536M calls
# In a continuous year of recording:  365*24*60*60/10 = 3.1536e6 

n = 1e6; # number of simulated calls (regardless of whether detected or not); 
R = 1e6; # radius of 1000 km 


# Call density, D_c is n/A 
A = pi*(R/1e3)^2 # Circular study area in km^2
TrueCallDensity = n/A # calls/km^2

k <- 1      # number of sensors
Time <- 1   # years

r = R*sqrt(runif(n));
t = 2*pi*runif(n);
sim <- data.frame(x = r*cos(t), y = r*sin(t), groundTruth=1);

# Date and time of detections span 1 full year (calendar year of)

minDate <- as.POSIXct("2025-01-01",tz='UTC')
maxDate <- as.POSIXct("2025-12-31",tz='UTC')

duration_s <- as.numeric(difftime(maxDate,minDate,unit="sec"))
sim$datetime <- minDate + sort(runif(n, 0, duration_s))

sim$d <- sqrt(sim$x^2+sim$y^2)

print(paste(TrueCallDensity,"calls per km^2 per unit time"))

```
Sonar equation parameters for this scenario:

\begin{equation}
TL = 20  log_{10}  d
\end{equation}
\begin{equation}
SNR = SL-TL-NL
\end{equation}


```{r sonar.eq}

SL <- data.frame(mean=190, sd=4, sampleSize=350) # True SL distribution for sim
sim$SL <- rnorm(n, mean=SL$mean, sd=SL$sd)       # Realised SL for each call

NL = data.frame(mean=84, sd = 4, sampleSize = n)  # True NL distribution for sim
sim$noiseRMSdB <- rnorm(n, mean=NL$mean, sd=NL$sd)# Noise for each call

range_m = seq(from=5, to=1e6, by=100)

# Spherical spreading for a single transect
tlTransectSpherical <- 20*log10(range_m)

tlTransect <- tlTransectSpherical
sim$TL <- 20*log10(sim$d)

# Create additional transects here to better match with our real-world approach
TL <- data.frame(
  range_m,
  tl000 = tlTransect,
  tl090 = tlTransect,
  tl180 = tlTransect,
  tl270 = tlTransect
)
sim$signalRMSdB <- with(sim,SL-TL)
sim$snr <- with(sim,SL-TL-noiseRMSdB)

```

### Simulate detectors:

| Detector | Description | Location/SNR 'threshold' | Scale/slope | False positive rate |
|---------------|---------------|---------------|---------------|---------------|
| Detector 1 | Reliable and good detector, meant to emulate a human analyst. | Lower | Same | Lower |
| Detector 2 | Automated detector, meant to emulate a signal processing algorithm. | Higher | Same | Higher |


```{r parameterise.detectors, warning=FALSE, message=FALSE}

#Specify parameters for detector 1
# Good detector with low threshold (location) and low false positive rate.
det1params = data.frame( 
  location=1,    # AKA intercept?
  scale=1,       # AKA slope?
  func='plogis', # logistic function
  c=0.1,         # False discovery rate
  fpMean=0,      # Mean of distribution of false positives (in dB SNR)
  fpSD = 1      # Standard deviation false positive distribution (in dB SNR)
)

# Detector 2 has a higher threshold (location), scale, and false positive rate
# than detector 1.
det2params = data.frame( 
  location=3,    # AKA intercept?
  scale=2,       # AKA slope?
  func='plogis', # logistic function
  c=0.3,         # False discovery rate
  fpMean=0,      # Mean of distribution of false positives (in dB SNR)
  fpSD = 1      # Standard deviation false positive distribution (in dB SNR)
)
```

Now we create a function that will take the detector parameters above and simulate the detection process. We then call this function with each set of parameters to create a simulation for each detector.

```{r simulate.detectors, warning=FALSE, message=FALSE}
simulateDetector <- function(detParams,sim){
  
  # Probability of detection for each simulated call
  sim$p_det = plogis(sim$snr,
                     location=detParams$location,
                     scale=detParams$scale)
  
  # Apply Bernoulli trial to SNRs to see which were detected
  sim$detect_table <- as.logical(rbinom(dim(sim)[1],size=1,prob=sim$p_det))
  
  # Calculate number of false positives given the number of true positive
  # detections for the detector and the specified false discovery rate, c, for
  # that detector.
  n_tp_det1 <- sum(sim$detect_table) 
  n_fp_det1 <- n_tp_det1/(1-detParams$c)-n_tp_det1
  
  # Store false positives in data.frame with same format as sim
  fp <- data.frame(matrix(ncol=length(sim),nrow=n_fp_det1))
  colnames(fp) <- colnames(sim)
  
  # groundTruth=0 means that these are false positives (handy to keep in mind for
  # when we include these in sim)
  fp$groundTruth <- FALSE
  fp$detect_table <- TRUE
  
  # Generate the right number of false positives uniformly over same time period
  # as true positives.
  fp$datetime <- minDate + sort(runif(n_fp_det1, 0, duration_s))  
  
  # False positives don't have a location, SL, or TL, but do have NL and SNR
  # Use same distribution of NL as true positives
  fp$noiseRMSdB <- rnorm(n_fp_det1, mean=NL$mean, sd=NL$sd)
  
  # SNR distribution of false positives broadly matches that of Casey2019 human
  # analysts
  fp$snr <- rnorm(n_fp_det1, mean=detParams$fpMean, sd=detParams$fpSD) 
  
  fp$signalRMSdB <- fp$noiseRMSdB + fp$snr
  
  # Combine false positives into the simulation
  sim <- rbind(sim,fp) 
  
  # We've generated the right number of false positive detections now and merged
  # these into our simulation. 
  
  # The false positives are missing locations, distances, SL, and p_dets. But
  # not sure that this actually matters in any meaningful way.

  # Add factor column to track true positive, false positivs & missed detections
  sim$group <- factor(ifelse(sim$groundTruth,
                             ifelse(sim$detect_table, 
                                    "TruePositive","FalseNegative"),
                             "FalsePositive"),
                      levels=c("FalsePositive","TruePositive","FalseNegative")
  )
  
  return(sim)
  
}

# Simulate two detectors each with the different parameters
simDet1<- simulateDetector(det1params,sim)
simDet2<- simulateDetector(det2params,sim)

```

## SNR Truncation

Apply SNR truncation to the detectors by removing all detections that were less than the threshold.

```{r}
snrTruncationThreshold <- 0; # dB
nAboveThreshold <- sum(sim$snr>snrTruncationThreshold)
simDet1 <- subset(simDet1, simDet1$snr >= snrTruncationThreshold)
simDet2 <- subset(simDet2, simDet2$snr >= snrTruncationThreshold)

```

View the spatial detection density (which is not the same as spatial call density since it  accounts for neither detection probability nor time).

```{r detector.plots, warning=FALSE, message=FALSE}

plotSpatialDetections <- function(sim){
  
  # Spatial distribution (excludes false positives)
  ggplot(data=sim, aes(x=x/1e3, y=y/1e3, weight=detect_table) )+
    geom_bin_2d(alpha=1,binwidth=c(10,10))+
    coord_equal()+
    xlab("X location (km)")+
    ylab("Y location (km)")
}

plotSpatialDetections(simDet1)
plotSpatialDetections(simDet2)


plotDetectionDistribution <- function(sim){
  # SNR Distribution (includes false positives)
  ggplot(data=sim, aes(x=snr, group=detect_table, fill=detect_table) )+
    geom_histogram()
  
  ggplot(data=sim, aes(x=d/1e3, y=snr, colour=detect_table))+
    geom_point(alpha=0.1, size=1)+
    xlab("Distance (km)")+
    ylab("SNR (dB)")+
    scale_x_log10(limits=c(0.1,1000),
                  labels=label_log(),
                  breaks=10^(-1:7),
                  minor_breaks = rep(1:9,6)*(10^rep(-2:7,each=9)))+
    annotation_logticks(sides='') +
    theme_bw()
  
  
  # SNR Distribution
  # ggplot(data=sim, aes(x=snr, group=group) )+
  #   geom_histogram(position = position_stack())
  
  ggplot(data=sim, aes(x=snr, group=group) )+
    geom_histogram(position = position_identity())+
    facet_wrap(group~., ncol=1,scales="free_y")

}

dist1 <- plotDetectionDistribution(simDet1)
dist2 <- plotDetectionDistribution(simDet2)
gridExtra::grid.arrange(dist1,dist2,nrow=1)

```

## Subsample data for detector characterisation

@castro2024 subsample approximately 200 hours evenly spaced throughout the year to create detector characterisation curves (detection vs SNR). Here we subsample 1 hour in every 41 hours yielding 214 subsamples that are 1 hour in duration.

```{r subsample.for.detector.characterisation}

subStart <- seq(from=minDate,to=maxDate,by='41 hour')
subEnd <- subStart+3600

# There's probably a more elegant way to do this, but loop over each subsample
# period and find all calls in that subset.
simDet1$subset <- 0
simDet2$subset <- 0
for (i in 1:length(subStart)){
  simDet1$subset <- simDet1$subset | 
    (simDet1$datetime >= subStart[i] & simDet1$datetime <= subEnd[i])
  simDet2$subset <- simDet2$subset | 
    (simDet2$datetime >= subStart[i] & simDet2$datetime <= subEnd[i])
}


subsampleDet1 <- subset(simDet1, simDet1$subset)
# Total positives, true-positives, and false positives in detector2 subsample
n_p_subsampleDet1 <- sum(subsampleDet1$detect_table) 
n_tp_subsampleDet1 <- with(subsampleDet1,sum(groundTruth & detect_table))
n_fp_subsampleDet1 <- sum(subsampleDet1$groundTruth==FALSE)

subsampleDet2 <- subset(simDet2, simDet2$subset)
# Total positives, true-positives, and false positives in detector2 subsample
n_p_subsampleDet2 <- sum(subsampleDet2$detect_table) 
n_tp_subsampleDet2 <- with(subsampleDet2,sum(groundTruth & detect_table))
n_fp_subsampleDet2 <- sum(subsampleDet2$groundTruth==FALSE) 


```

## Detection matching between detector 1 and 2

Both detectors (simulated subsets) operate on the same set of calls (i.e. they are different subsets of the same underlying simulation). So we can find detections from one detector that match those of the other detector by merging the detections into a capture history table.

Presently, false positives have randomly generated date-times, so have a negligible chance of matching between the two detectors.

*TODO: Create false positives in a way that better approximates their real-world occurrence (i.e. in a way that allows them to be matched between detectors).*

```{r match detections from detector1 and detector2}

# Create a capture history table from the two tables of detections
captureHistoryTable <- function(subsampleDet1, subsampleDet2){
  capHistTab <- merge(x=subsampleDet1, y=subsampleDet2, all=TRUE,
                      by = c('datetime'), suffixes = c('1','2'))
  
  # Missed detections will be NA by default. Make them 0 instead.
  capHistTab$detect_table1[is.na(capHistTab$detect_table1)]<-0
  capHistTab$detect_table2[is.na(capHistTab$detect_table2)]<-0
  
  # groundtruth and group columns will have NA by default, so need to copy them
  # from other detector. TODO:Look into powerjoin, dplyr::coalesce and
  # coalesce_join (https://alistaire.rbind.io/blog/coalescing-joins/) for more
  # elegant solutions
  capHistTab$groundTruth1[is.na(capHistTab$groundTruth1)]<-
    capHistTab$groundTruth2[is.na(capHistTab$groundTruth1)]
  capHistTab$groundTruth2[is.na(capHistTab$groundTruth2)]<-
    capHistTab$groundTruth1[is.na(capHistTab$groundTruth2)]
  
  capHistTab$group1[is.na(capHistTab$group1)]<-
    capHistTab$group2[is.na(capHistTab$group1)]
  capHistTab$group2[is.na(capHistTab$group2)]<-
    capHistTab$group1[is.na(capHistTab$group2)]
  
  return (capHistTab)
}

capHistTab<- captureHistoryTable(subsampleDet1,subsampleDet2)

# All simulated calls in the subset (detected or not)
dim(capHistTab)

# Total number of true positive detections for detector1 OR detector2
nDetectedSubset <- with(capHistTab,sum(detect_table1 & groundTruth1 |   
                                        detect_table2 & groundTruth2) )
nDetectedSubset

# Total number of positive detections for detector1 OR detector2
nPositiveSubset <- with(capHistTab,sum(detect_table1 | detect_table2))
nPositiveSubset

```

Convert this capture history table into an SNRinfo table, and treat detector1 as ground truth.

```{r format.output.for.callDensity.package}

# Function to convert the capture history table into the 'SNRinfo' format used
# by the callDensity package
capHistTosnrInfo <- function(capHistTab){

  # Assume detect_table1 is ground truth, so only keep rows where
  # detect_table1==1 (includes both true and false positives though)
  capHistTab <- capHistTab[capHistTab$detect_table1==1, ]

  # Detected refers to the automated detector, here detect_table2
  Detected<-capHistTab$detect_table2
    
  # Add time rows for compatibility with callDensity format
  t <- capHistTab$datetime
  season <- time2season(t)
  month <- time2monthCode(t)
  
  # CallRL and NoiseRL here in dB
  CallRL<-as.numeric(rowMeans(
    subset(capHistTab,select=c('signalRMSdB1','signalRMSdB2')) ,na.rm=T) )#dB
  NoiseRL<-as.numeric(rowMeans(
    subset(capHistTab,select=c('noiseRMSdB1','noiseRMSdB2')) ,na.rm=T) )#dB

  # SNR<-as.numeric(capHistTab$signalRMSdB-capHistTab$noiseRMSdB) # SNR measured as in DH
  SNR<-CallRL-NoiseRL # to check
  SNRinfo<-data.frame(Detected,CallRL,NoiseRL,SNR,t,month,season)

  # Check for SNR=Inf: put in a warning to user here, since infinite SNR is
  # likely indicative of a problem with the inputs
  lines<-c(which(SNRinfo$SNR=="Inf", arr.ind = TRUE))
  if ( length(lines) >0) { # Run this line if inf occurs
    warning( sprintf('%g of the SNR were infinite:\n',length(lines)) )
    SNRinfo <- SNRinfo[-lines,]
  }

  return(SNRinfo)
}

# SNRinfo is used to calculate snrDetFun for detector2 assuming detector1 is the
# ground truth. This is an observer ground (OG) so includes false positives from
# detector1, but does not include detections from detector2 not detected on
# detector1
SNRinfo <- capHistTosnrInfo(capHistTab)

# Positive on detector 2, but not on 1
n_fp_subsample <- sum(capHistTab$detect_table2 & !capHistTab$detect_table1)

# False discovery rate is number.false.positives/number.predicted.positive
c_subsample <- n_fp_subsample/n_p_subsampleDet2
n_subsample <- dim(SNRinfo)[1]  # Number positive detections in SUBset

truncationDistances <- R 

# Number positive detections in FULL set
Nc <- sum(simDet2$detect_table)       

# Noise level from subsampled Observer ground-truth
NLsamp <-SNRinfo %>% dplyr::summarise(mean=mean(NoiseRL,na.rm = TRUE),
                             sd=sd(NoiseRL,na.rm = TRUE),
                             sampleSize=dplyr::n()-sum(is.na(NoiseRL)))

NLtable <- rbind(NL,NLsamp)
rownames(NLtable)<- c('Actual','Estimated from subsample')
kable(NLtable,caption = 'Noise level distributions')%>% 
  kableExtra::kable_classic(full_width=FALSE)

```

## Calculate call densities using callDensity package

The detector was a logistic function, so fitting a GLM should provide good results.

We also fit a capture-recapture model to these adjudicated data using the VGAM package with the function 'vglm' [@yee2015].

A GAM or SCAM should also be able to provide a good fit to a logistic function, but to keep this concise we leave fitting those as an exercise for the reader (see callDensity vignette).

```{r GLM p.det with independently estimated NL, warning=FALSE, message=FALSE}

snrDetFun.glm <- callDensity::fitSNRdetectionFunc(SNRinfo,
                                              modelType = "glm")
snrDetFun.glm$r.sq <- with(summary(snrDetFun.glm), 1 - deviance/null.deviance)

# Estimate of NL distribution
NLind<- nlFromSnrInfo(SNRinfo, snrDetFun.glm)


Pa.NLi <- callDensity::pDetInArea(snrDetFun.glm, SL=SL, TLlookup = TL, NL=NLind,
           transectFile = 'pDet_test_NLind_year_transects.csv',
           simResultsFile = 'simResult_test_NLind_year.txt',
           paFile= 'Pa_test_NLind_year.txt',
           output.resolution.m=100,
           outerloop = 100,
           truncationDistance=R, # in m
           snrTruncationThreshold =snrTruncationThreshold)
pa.NLi <- Pa.NLi$overall


# Density estimation equation
Dc.NLi <- (Nc * (1 - c_subsample) )/( k * A * pa.NLi * Time )

results.NLind <- data.frame(Dc=Dc.NLi, SampleSize=n_subsample, A = A,# Environment
                      NLmean = NLind$mean, NLsd = NLind$sd,      # Noise
                        # Detector params
                        model='glm', R2=snrDetFun.glm$r.sq,
                        AIC =AIC(snrDetFun.glm), c=c_subsample, Pa=pa.NLi,  
                      row.names = 'NL independent')


ggplot(data=Pa.NLi$meanOfAllTransects, aes(x=range_m/1e3,y=pDet))+
  geom_point(alpha=0.1, size=0.1)+
  xlab("Range (km)")+
  ylab("Probability of detection\n(Mean across all transects)")+
  scale_x_log10(limits=c(0.1,1000),
                labels=label_log(),
                breaks=10^(-1:7),
                minor_breaks = rep(1:9,6)*(10^rep(-2:7,each=9)))+
  annotation_logticks(sides='') +
  theme_bw()


```

```{r Call.Density.Results, echo=TRUE, warning=FALSE}

results = data.frame(Dc = TrueCallDensity, SampleSize = n, A = A,# Environment 
                      NLmean = NL$mean, NLsd = NL$sd,            # Noise
                        # Detector params
                        model=det2params$func, R2=NA,  
                        AIC =NA, c = det2params$c, 
                        Pa = mean(simDet2$p_det,na.rm=TRUE), 
                      row.names='Truth')

results<-rbind( results, results.NLind)

kableExtra::kbl(results, digits = c(4,3,0,1, 2, NA, 3, 3, 3, 4), 
                col.names= c('Dc','(sub)Sample Size', 'A','NL_mean','NL_sd',
                             'Model','R^2','AIC','c','P_a')) %>% 
  kableExtra::kable_classic(full_width=FALSE)


```

We fit a capture-recapture model to these adjudicated data using the VGAM package with the function 'vglm' [@yee2015].

```{r VGLM p.det, warning=FALSE, message=FALSE}
library(VGAM)
### Adjudicated capture recapture model
adjudicated <- subset(capHistTab,capHistTab$groundTruth1 &
                (capHistTab$detect_table1 | capHistTab$detect_table2))

summary(adjudicated$group1)
summary(adjudicated$group2)

# Number of adjudicated positive detections
n_adj <- dim(adjudicated)[1]
n_adj

# Combine SNR of both detectors (shouldn't be necessary unless we've somehow
# made them different)
adjudicated$SNR <-   rowMeans(
  subset(adjudicated,select=c('snr1','snr2')) ,na.rm=T) #dB

# Subset used for calculating false discovery rate of detector 2. Actually, we
# use the fact that the false discovery rate is equal to (1-precision). 
tp <- subset(capHistTab, as.logical(capHistTab$detect_table2), 
              select = c('groundTruth2') )
precision <- sum(tp)/dim(tp)[1]
adj_c <- 1-precision

observerNames = c("detect_table1", "detect_table2")
snrDetFun.vglm <- fitSNRvglm(adjudicated, observerNames, whichObserver = "detect_table2")

# Estimate of NL distribution
NLadj<- nlFromSnrInfo(SNRinfo, snrDetFun.glm)

Pa.vglm <- callDensity::pDetInArea(snrDetFun.vglm, SL=SL, TLlookup = TL, NL=NLadj,
           transectFile = 'pDet_test_vglm_year_transects.csv',
           simResultsFile = 'simResult_test_vglm_year.txt',
           paFile= 'Pa_test_vglm_year.txt',
           output.resolution.m=100,
           outerloop = 10,
           truncationDistance=R, # in m
           snrTruncationThreshold = snrTruncationThreshold)
pa.vglm <- Pa.vglm$overall


# Density estimation equation
Dc.vglm <- (Nc * (1 - adj_c) )/( k * A * pa.vglm * Time )

results.vglm <- data.frame(Dc=Dc.vglm, SampleSize=n_adj, A = A,# Environment
                      NLmean = NLadj$mean, NLsd = NLadj$sd,# Noise
                      # Detector params
                        model='vglm', R2=NA,
                        AIC =AIC(snrDetFun.vglm), c=adj_c, Pa=pa.vglm,  
                      row.names = 'Adjudicated Capture-recapture\nNL independent')

ggplot(data=Pa.vglm$meanOfAllTransects, aes(x=range_m/1e3,y=pDet))+
  geom_point(alpha=0.1, size=0.1)+
  xlab("Range (km)")+
  ylab("Probability of detection\n(Mean across all transects)")+
  scale_x_log10(limits=c(0.1,1000),
                labels=label_log(),
                breaks=10^(-1:7),
                minor_breaks = rep(1:9,6)*(10^rep(-2:7,each=9)))+
  annotation_logticks(sides='') +
  theme_bw()

```

```{r}

results<-rbind( results, results.vglm)

kableExtra::kbl(results, digits = c(4,3,0,1, 2, NA, 3, 3, 3, 4), 
                col.names= c('Dc','(sub)Sample Size', 'A','NL_mean','NL_sd',
                             'Model','R^2','AIC','c','P_a')) %>% 
  kableExtra::kable_classic(full_width=FALSE)


```

The estimate of Dc derived from the vglm model is very close to the true value for this simulation.

Adjudicated capture-recapture along with independently estimated NL yields more accurate estimates of c and P_a.

## Inspect detection functions for insights

```{r compare snr detection functions, warning=FALSE}
SNR <- seq(-20,20,0.5);
snrs <- data.frame(SNR)

cols = colorRampPalette(c('#EDB120','#D95319','#0072BD','#7E2F8E','#77AC30',
                        '#4DBEEE','#A2142F'))
preds.glm= predict(snrDetFun.glm, type='response',newdata=snrs)
preds.vglm=predict(snrDetFun.vglm,type='response',newdata=snrs)
preds.vglm0=predict(snrDetFun.vglm,type='response',newdata=snrs,
                    type.fitted='onempall0')# One minus probability of all zeros
pr =apply(cbind(preds.vglm[,2],preds.vglm0),1,prod)


par(cex=0.8,mar=c(3.1,3.1,0.25,0),mgp=c(2.1,1,0))
plot(snrs$SNR, xlim= c(-5,10),  ylim= c(0,1),
     xlab="SNR", ylab = "P(Detection)", type='n')
grid(nx = NULL, ny = NULL, lty = 2, col = "gray", lwd = 2)
lines(snrs$SNR,preds.glm,col=cols(7)[4],lwd=2)
lines(snrs$SNR,preds.vglm[,1],col=cols(7)[5],lwd=2) # Percep bias detector1
lines(snrs$SNR,preds.vglm[,2],col=cols(7)[6],lwd=2) # Percep bias detector2
lines(snrs$SNR,preds.vglm0,col=cols(7)[2],lwd=2) # 1-
lines(snrs$SNR,pr,col=cols(7)[1],lwd=2)    # vglm 1-p(all zero)
points(simDet2$snr,simDet2$p_det,pch='.')     # Actual probability of detection
legend( "bottomright", 
        legend =c("Detection function TRUE",
                  "Detection function GLM",
                  "Detection function VGLM",
                  "Perception bias: detector1 (VGLM)",
                  "Perception bias: detector2 (VGLM)",
                  "1-p(all zeros) (VGLM)"
                  ),
        col=c('black', cols(7)[4], cols(7)[1], cols(7)[5], cols(7)[6], cols(7)[2]), 
        lty=rep('solid',6),
        lwd=rep(2,6)
)
```


# References
