---
title: "Parallel performance benchmarking"
output: 
  html_document: default
  rmarkdown::html_vignette:
    vignette: >
      %\VignetteIndexEntry{Parallel performance benchmarking}
      %\VignetteEncoding{UTF-8}
      %\VignetteEngine{knitr::rmarkdown}
  pdf_document: default
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.height = 3.75, 
  comment = "#>"
)
```

```{r setup}
library(callDensity)
library(ggplot2)
library(VGAM)
```

# Parallel performance benchmarking

The `pDetInArea` function performs a Monte Carlo simulation that can be
parallelized via the `parallel` argument in `cde()`. This vignette helps you
find the optimal number of parallel workers for your hardware by benchmarking
`cde()` with both lightweight (scam) and heavyweight (vglm) detection
functions across a range of worker counts.

**Important:** The optimal worker count depends on your CPU, memory, and
operating system. Results from one machine won't transfer to another. Run
this vignette on the machine you plan to use for production analyses.

## Key findings from development

During development on a 64-core Windows machine, we found:

- **scam/gam/glm models:** Per-iteration work is too lightweight for
  parallel overhead. `parallel=TRUE` is *slower* than serial. Use the
  default `parallel=FALSE`.
- **vglm models:** `VGAM::predict` is expensive enough to benefit from
  parallelism. A 9--12x speedup was achieved with 16--30 workers using
  `future.callr` as the backend.
- **Too many workers hurts:** Beyond ~50% of available cores, resource
  exhaustion (memory, file handles) can cause crashes on Windows. More
  workers also means more serialization overhead per chunk.
- **`future.callr`** is the recommended backend on Windows. The default
  `future::multisession` can serialize execution under RStudio's knit
  button.

## Generate simulated test data

We use the package's simulation functions to create a self-contained test
dataset with known ground truth.

```{r simulate.calls}
set.seed(42)

n <- 1e6
R <- 1e6          # radius in m
k <- 1
minDate <- as.POSIXct("2025-01-01")
maxDate <- as.POSIXct("2026-01-01")
Time <- as.numeric(difftime(maxDate, minDate, unit = "hours"))

A <- studyArea(R / 1e3)  # km^2

# Source level and noise level distributions
SL <- data.frame(mean = 189, sd = 8, sampleSize = 350)
NL <- data.frame(mean = 90, sd = 5, sampleSize = 1000)

# Simulate call locations
sim <- simCallLocation(n = n, R = R, minDate = minDate, maxDate = maxDate)

# Simulate acoustic properties (SL, NL, TL, SNR)
sim <- simCallAcoustics(sim, SL = SL, NL = NL)

cat("Simulated", n, "calls within", R/1e3, "km radius\n")
cat("Study area:", round(A), "km^2\n")
```

## Simulate detectors and create capture histories

We simulate two detectors: one with good sensitivity (detector 1) and one
slightly worse (detector 2). This mirrors the real-world scenario where we
have a human analyst and an automated detector.

```{r simulate.detectors}
# Detector 1: good sensitivity (like a human analyst)
det1params <- data.frame(location = 2, scale = 2, c = 0.05, func = 'logistic',
                         fpMean = -5, fpSD = 3)
simDet1 <- simulateDetector(det1params, sim)

# Detector 2: slightly worse (like an automated detector)
det2params <- data.frame(location = 4, scale = 2.5, c = 0.10, func = 'logistic',
                         fpMean = -3, fpSD = 4)
simDet2 <- simulateDetector(det2params, sim)

# Subsample to create annotated subset (as in real analyses)
subsampleDet1 <- subsampleSimInTime(simDet1, minDate, maxDate, 
                                     interval = '41 hour', duration = 3600)
subsampleDet2 <- subsampleSimInTime(simDet2, minDate, maxDate, 
                                     interval = '41 hour', duration = 3600)

# Create capture history table
capHistTab <- simsTocaptureHistoryTable(subsampleDet1, subsampleDet2)

cat("Capture history table:", nrow(capHistTab), "rows\n")
```

## Simulate transmission loss radials

```{r simulate.TL}
TL <- simTLradials_20logR(maxRange = R, rangeStep = 20, numTransects = 8)
truncDist <- matrix(rep(R, 8), nrow = 1)
season <- factor("year")

cat("TL table:", nrow(TL), "range steps x", ncol(TL)-1, "transects\n")
```

## Fit detection functions

We fit both a scam (lightweight) and a vglm (heavyweight) detection
function so we can benchmark both code paths.

```{r fit.scam, warning=FALSE, message=FALSE}
# OG-style capture history for scam
ch_scam <- capHistTab
ch_scam$detect_table1 <- as.logical(ch_scam$groundTruth1)
ch_scam$detect_table2 <- as.logical(ch_scam$detect_table2)
ch_scam$snr_table2 <- ch_scam$snr2
ch_scam$signalRMSdB_table2 <- ch_scam$signalRMSdB2
ch_scam$noiseRMSdB_table2 <- ch_scam$noiseRMSdB2

# Fit scam detection function
SNRinfo_scam <- capHist2snrInfo(ch_scam)
detFun_scam <- fitSNRdetectionFunc(SNRinfo_scam, modelType = 'scam',
                                    numKnots = 5)
cat("scam model fitted:", class(detFun_scam)[1], "\n")
```

```{r fit.vglm, warning=FALSE, message=FALSE}
# Adjudicated data for VGLM (true positives only from both detectors)
adjudicated <- subset(capHistTab, capHistTab$groundTruth1 &
                        (capHistTab$detect_table1 | capHistTab$detect_table2))

observerNames <- c("detect_table1", "detect_table2")
detFun_vglm <- fitSNRvglm(adjudicated, observerNames,
                            whichObserver = "detect_table2")

# CR-style capture history for cde
ch_vglm <- capHistTab
ch_vglm$detect_table1 <- as.logical(
  ch_vglm$detect_table1 | ch_vglm$detect_table2)
ch_vglm$detect_table2 <- as.logical(ch_vglm$detect_table2)
ch_vglm$snr_table2 <- ch_vglm$snr2
ch_vglm$signalRMSdB_table2 <- ch_vglm$signalRMSdB2
ch_vglm$noiseRMSdB_table2 <- ch_vglm$noiseRMSdB2

Nc <- sum(simDet2$detect_table)

cat("vglm model fitted:", paste(class(detFun_vglm), collapse=", "), "\n")
cat("Total detections (Nc):", Nc, "\n")
```

## Configure benchmark parameters

Adjust these to suit your machine. Start conservative and increase
`max_workers` if your system has headroom.

```{r config}
# Number of Monte Carlo iterations for benchmarking.
# Use 100 for a quick survey (~15 min), 1000 for production timings.
outerloop <- 256

# Worker counts to test. Adjust based on your core count.
n_cores <- parallelly::availableCores()
cat("Available cores:", n_cores, "\n")

# Test a range from serial through ~75% of cores.
# Going beyond ~50% of cores may cause resource exhaustion on Windows.
worker_counts <- unique(sort(c(
  4,
  min(8, n_cores - 1),
  min(16, n_cores - 1),
  min(round(n_cores / 2), n_cores - 1),
  min(round(n_cores * 0.75), n_cores - 1)
)))
worker_counts <- worker_counts[worker_counts >= 4]

cat("Worker counts to test:", worker_counts, "\n")
```

## Run benchmarks

```{r benchmark, message=FALSE, warning=FALSE}
# Helper: run a timed cde call
run_timed_cde <- function(ch, detFun, model_type, outerloop, parallel) {
  elapsed <- system.time(
    cde(Nc = Nc, capHistTab = ch, snrDetFun = detFun,
        SL = SL, TL = TL, T = Time, A = A,
        modelType = model_type, season = season,
        outerloop = outerloop, truncationDistance = truncDist,
        output.resolution.m = 100, siteCode = "bench",
        parallel = parallel)
  )["elapsed"]
  as.numeric(elapsed)
}

results <- data.frame()

# --- Serial baselines --------------------------------------------------------
cat("\n=== Serial baselines ===\n")

cat("  scam serial... ")
t_scam_serial <- run_timed_cde(ch_scam, detFun_scam, "scam", outerloop, FALSE)
cat(round(t_scam_serial, 1), "s\n")

cat("  vglm serial... ")
t_vglm_serial <- run_timed_cde(ch_vglm, detFun_vglm, "vglm", outerloop, FALSE)
cat(round(t_vglm_serial, 1), "s\n")

results <- rbind(results, data.frame(
  model = c("scam", "vglm"), workers = 0, outerloop = outerloop,
  elapsed_s = c(t_scam_serial, t_vglm_serial),
  stringsAsFactors = FALSE
))

# --- Parallel benchmarks -----------------------------------------------------
for (nw in worker_counts) {
  cat(sprintf("\n=== %d workers ===\n", nw))

  # Set plan and warm up
  future::plan(future.callr::callr, workers = nw)

  cat("  Warming up... ")
  warmup_t <- system.time(
    cde(Nc = Nc, capHistTab = ch_vglm, snrDetFun = detFun_vglm,
        SL = SL, TL = TL, T = Time, A = A,
        modelType = "vglm", season = season, outerloop = 5,
        truncationDistance = truncDist, output.resolution.m = 100,
        siteCode = "warmup", parallel = TRUE)
  )["elapsed"]
  cat(round(as.numeric(warmup_t), 1), "s\n")

  # scam parallel
  cat("  scam parallel... ")
  t_scam <- run_timed_cde(ch_scam, detFun_scam, "scam", outerloop, TRUE)
  cat(round(t_scam, 1), "s\n")

  # vglm parallel
  cat("  vglm parallel... ")
  t_vglm <- run_timed_cde(ch_vglm, detFun_vglm, "vglm", outerloop, TRUE)
  cat(round(t_vglm, 1), "s\n")

  results <- rbind(results, data.frame(
    model = c("scam", "vglm"), workers = nw, outerloop = outerloop,
    elapsed_s = c(t_scam, t_vglm),
    stringsAsFactors = FALSE
  ))
}

# Clean up
future::plan('sequential')
```

## Results

```{r results.table}
# Add speedup column
serial_times <- results[results$workers == 0, c("model", "elapsed_s")]
names(serial_times)[2] <- "serial_s"
results <- merge(results, serial_times, by = "model")
results$speedup <- round(results$serial_s / results$elapsed_s, 1)

# Sort for display
results <- results[order(results$model, results$workers), ]
rownames(results) <- NULL

knitr::kable(results[, c("model", "workers", "elapsed_s", "speedup")],
             digits = c(NA, 0, 1, 1),
             col.names = c("Model", "Workers", "Elapsed (s)", "Speedup"),
             caption = paste("Benchmark results: outerloop =", outerloop))
```

```{r results.plot, fig.width=6, fig.height=4}
ggplot(results, aes(x = workers, y = speedup, colour = model, group = model)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  annotate("text", x = max(results$workers) * 0.8, y = 0.8,
           label = "serial baseline", colour = "grey50", size = 3) +
  scale_x_continuous(breaks = c(0, worker_counts)) +
  labs(
    title = "Parallel speedup vs number of workers",
    subtitle = paste("outerloop =", outerloop, " |  cores =", n_cores),
    x = "Number of workers",
    y = "Speedup (serial time / parallel time)",
    colour = "Model type"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Interpretation

Look at the plot above to find your optimal worker count:

- **scam/gam/glm models:** If the scam line stays near or below 1.0x (the
  dashed baseline), parallel execution adds overhead without benefit. Use
  `parallel=FALSE` for these models.

- **vglm models:** The vglm line should rise with more workers. The optimal
  point is where the curve starts to flatten --- adding more workers beyond
  that gives diminishing returns and increases memory pressure.

- **If any test crashed** with a `FutureLaunchError`, that worker count
  exceeds your system's resources. Use fewer workers.

## Recommended configuration

Based on your results, set the plan in your analysis script:

```r
# Replace N with your optimal worker count from the plot above
future::plan(future.callr::callr, workers = N)

# Then call cde with parallel=TRUE for vglm models only
result <- cde(..., modelType = "vglm", parallel = TRUE)
```

For scam/gam/glm models, omit `parallel` or set it to `FALSE`:

```r
result <- cde(..., modelType = "scam")  # serial by default
```

## Save results

```{r save.results}
outfile <- sprintf("parallel_benchmarks_%s.csv",
                   format(Sys.time(), "%Y%m%d_%H%M"))
write.csv(results, file = outfile, row.names = FALSE)
cat("Results saved to", outfile, "\n")
```
