---
title: "Where capture-recapture estimates of p_a break down (serial)"
author: "Brian Miller"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5)
library(callDensity)
library(ggplot2)
```

# The question

Call density depends on $p_a$: the mean probability of hearing a call anywhere
in the study area. Nearly every call is far away and faint, so $p_a$ is decided
almost entirely by how well the detector hears faint calls.

But a detector can only be measured on calls somebody heard, and those are the
loud ones. Calls that every observer missed leave no trace. So the detection
function is fitted where the loud calls are, and then used to answer a question
about faint ones. The region that decides the answer is extrapolated.

These notes ask how far that extrapolation can be pushed before $p_a$ stops
being usable, and what property of the partner observer decides it.

Two things are compared.

**Capture-recapture (CR).** Both observers' detections are modelled jointly
with a positive-Bernoulli model (VGAM `posbernoulli.t`), conditioning on a call
having been detected by at least one observer. The sample is calls somebody
heard.

**Oracle.** Detector 2's own detections over *every call that was made*,
including the ones nobody heard. Not a method: this sample cannot exist. It is
the control. It differs from CR in exactly one respect, SNR coverage, so any
difference between them is attributable to coverage alone.

The hypothesis is that a **shallow** partner keeps contributing detections
further down into the faint region, so it constrains the curve where $p_a$
lives. A steep partner is a step function: above its threshold it hears
everything, below it nothing, so the faint region has no recaptures and the
curve is free to go where it likes.

Note this is about the partner's **slope**, not its recall. Those come apart.

# Setup

```{r simulate-fn}
#' Simulate two detectors listening to the same calls.
#'
#' The random number generator kind is forced, not just seeded. Otherwise a
#' parallel backend that switches to L'Ecuyer-CMRG would silently produce a
#' different realisation from the same seed, and cde's internal draws would
#' differ too. With the kind forced, every cell is fully deterministic and the
#' serial and parallel versions of this document must agree exactly.
#'
#' @param det1scale,det2scale Logistic scale. Small means a steep, nearly
#'   step-like detector. Large means a shallow one that degrades gradually.
#' @param det1location,det2location SNR at which the detector hears half of
#'   calls. Held at 0 here: earlier sweeps showed location has little effect.
#' @param fdr False discovery rate of both detectors. Held at 0: CR recovers
#'   the false discovery rate exactly, so it is not the subject here.
#' @param n Number of calls.
#' @param seed Random seed. A grid dimension, not a constant: one realisation
#'   is one ticket in a lottery.
simulatePair <- function(det1scale = 2, det2scale = 4,
                         det1location = 0, det2location = 0,
                         fdr = 0, n = 5e4, seed = 1) {

  RNGkind("Mersenne-Twister")
  set.seed(seed)

  R       <- 1e6                                  # study radius, m
  minDate <- as.POSIXct("2025-01-01")
  maxDate <- as.POSIXct("2026-01-01")
  Time    <- as.numeric(difftime(maxDate, minDate, unit = "days")) / 365
  A       <- studyArea(R / 1e3)

  SL <- data.frame(mean = 190, sd = 4, sampleSize = 350)
  NL <- data.frame(mean = 84,  sd = 4, sampleSize = n)

  det1 <- data.frame(location = det1location, scale = det1scale,
                     func = "plogis", c = fdr, fpMean = 0, fpSD = 2)
  det2 <- data.frame(location = det2location, scale = det2scale,
                     func = "plogis", c = fdr, fpMean = 0, fpSD = 2)

  sim     <- simCallLocation(n = n, R = R, minDate = minDate, maxDate = maxDate)
  sim     <- simCallAcoustics(sim, SL, NL, TL = function(r) 20 * log10(r))
  simDet1 <- simulateDetector(det1, sim)
  simDet2 <- simulateDetector(det2, sim)

  capHistTab <- simsTocaptureHistoryTable(simDet1, simDet2)

  # The adjudicated verdict is truth. Restricted to calls at least one detector
  # flagged, which is the sample a positive-Bernoulli model is defined on.
  chCR <- capHistTab[capHistTab$detect_table1 == 1 |
                     capHistTab$detect_table2 == 1, ]
  chCR$detect_observer1 <- as.integer(chCR$detect_table1)
  chCR$detect_observer2 <- as.integer(chCR$detect_table2)
  chCR$detect_table1    <- as.integer(chCR$groundTruth1)

  real <- simDet2$groundTruth == 1

  list(
    chCR   = chCR,
    SL     = SL,
    NL     = NL,
    TL     = simTLradials_20logR(maxRange = R, rangeStep = 1000,
                                 numTransects = 4),
    A      = A,
    Time   = Time,
    Nc     = sum(simDet2$detect_table),

    # The oracle sample. Every call that was made, whether anyone heard it or
    # not. Same quantity as the CR sample, but with full SNR coverage.
    oracleSnr = data.frame(Detected = simDet2$detect_table[real],
                           SNR      = simDet2$snr[real]),

    # The truth.
    trueDc       = n / (A * Time),
    truePa       = mean(simDet2$p_det, na.rm = TRUE),
    trueC        = fdr,
    det2location = det2location,
    det2scale    = det2scale
  )
}

#' The CR sample: adjudicated real calls, both observers' verdicts retained.
#' capHist2snrInfo drops the per-observer columns a VGLM needs, so build it here.
snrInfoCR <- function(s) {
  g   <- s$chCR[s$chCR$detect_table1 == 1, ]
  out <- data.frame(
    detect_observer1 = g$detect_observer1,
    detect_observer2 = g$detect_observer2,
    Detected         = g$detect_observer2,
    CallRL           = as.numeric(g$signalRMSdB),
    NoiseRL          = as.numeric(g$noiseRMSdB),
    SNR              = as.numeric(g$signalRMSdB - g$noiseRMSdB),
    t                = g$t,
    month            = g$month,
    season           = g$season
  )
  out[is.finite(out$SNR), ]
}

fitCR <- function(snrCR) {
  fitDetFun(snrCR, modelType = "vglm",
            yColNames     = c("detect_observer1", "detect_observer2"),
            whichObserver = "detect_observer2")
}

#' Probability at least one observer detects, as a plain function of SNR. This
#' is the filter that selected the CR sample, so it is what the noise
#' correction has to undo. VGAM computes it as type.fitted "onempall0".
pUnionFromVglm <- function(model, snrGrid = seq(-100, 100, by = 0.1)) {
  pu <- VGAM::predictvglm(model, newdata = data.frame(SNR = snrGrid),
                          type = "response", type.fitted = "onempall0")
  stats::approxfun(snrGrid, as.vector(pu), rule = 2)
}

runOne <- function(label, s, detFun, nl, modelType) {
  r <- cde(Nc = s$Nc, capHistTab = s$chCR, snrDetFun = detFun,
           SL = s$SL, TL = s$TL, A = s$A, T = s$Time, NL = nl,
           modelType = modelType)
  data.frame(row        = label,
             truePa     = s$truePa, pa = r$pa,
             paFraction = r$pa / s$truePa,
             Dc         = r$Dc,
             DcFraction = r$Dc / s$trueDc,
             CV.pa      = r$CV.pa,
             stringsAsFactors = FALSE)
}
```

# The sweep

`det1scale` runs from 0.5 (a hard threshold) to 8 (a detector that degrades
very gradually). `det2scale` takes two values so the target detector is once
steep and once shallow. Ten seeds per cell, because a single realisation says
nothing about the spread.

```{r sweep-fn}
sweepCell <- function(det1scale, det2scale, seed) {

  s     <- simulatePair(det1scale = det1scale, det2scale = det2scale,
                        seed = seed)
  snrCR <- snrInfoCR(s)

  funCR     <- fitCR(snrCR)
  funOracle <- fitDetFun(s$oracleSnr, modelType = "glm")

  nlCR <- nlFromDetections(snrCR, pUnionFromVglm(funCR), s$SL, s$TL)

  out <- rbind(
    runOne("CR",     s, funCR,     nlCR,  "vglm"),
    runOne("oracle", s, funOracle, s$NL,  "glm")
  )

  cbind(data.frame(row = out$row, det1scale = det1scale,
                   det2scale = det2scale, seed = seed,
                   nCR = nrow(snrCR),
                   snrP05 = as.numeric(stats::quantile(snrCR$SNR, 0.05)),
                   stringsAsFactors = FALSE),
        out[, setdiff(names(out), "row")])
}

detGrid <- expand.grid(det1scale = c(0.5, 1, 2, 4, 8),
                       det2scale = c(1, 4),
                       seed      = 1:10,
                       KEEP.OUT.ATTRS = FALSE)
```

```{r sweep-run}
sweep <- do.call(rbind, lapply(seq_len(nrow(detGrid)), function(i) {
  g <- detGrid[i, ]
  tryCatch(sweepCell(g$det1scale, g$det2scale, g$seed),
           error = function(e) {
             message(sprintf("cell %d failed: %s", i, conditionMessage(e)))
             NULL
           })
}))
```

# Does the partner's slope decide it?

`paFraction` is the estimated probability of detection divided by the true one.
A usable method sits on the dashed line in every cell and every realisation.
Each point is one seed.

```{r plot-pafraction}
sweep$method <- factor(sweep$row, levels = c("oracle", "CR"))

ggplot(sweep, aes(x = factor(det1scale), y = paFraction, colour = method)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15,
                                             dodge.width = 0.75),
             alpha = 0.7, size = 1.5) +
  facet_wrap(~ det2scale, labeller = label_both) +
  scale_colour_manual(values = c(oracle = "steelblue", CR = "firebrick")) +
  labs(x = "detector 1 scale (small = steep threshold, large = gradual)",
       y = expression(hat(p)[a] / p[a]),
       title = "Estimated over true probability of detection",
       subtitle = "10 realisations per cell. The oracle sees every call; CR sees only calls somebody heard.") +
  theme_bw()
```

The oracle is the control. It uses the same `cde`, the same calls and the same
detector 2. It differs from CR only in that its sample covers the whole SNR
range. If the oracle sits on the line while CR scatters, the scatter is
coverage.

```{r plot-dcfraction}
ggplot(sweep, aes(x = factor(det1scale), y = DcFraction, colour = method)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15,
                                             dodge.width = 0.75),
             alpha = 0.7, size = 1.5) +
  facet_wrap(~ det2scale, labeller = label_both) +
  scale_colour_manual(values = c(oracle = "steelblue", CR = "firebrick")) +
  labs(x = "detector 1 scale (small = steep threshold, large = gradual)",
       y = expression(hat(D)[c] / D[c]),
       title = "Estimated over true call density") +
  theme_bw()
```

## Does the model know when it is wrong?

`CV.pa` is the coefficient of variation `cde` reports for $p_a$. It measures how
well the curve is pinned down *where the data is*. It cannot know about the
extrapolation. If the red points sit low while `paFraction` scatters widely,
the confidence intervals are confidently wrong.

```{r plot-cv}
sweep$absErr <- abs(sweep$paFraction - 1)

ggplot(sweep, aes(x = CV.pa, y = absErr, colour = method)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.7) +
  scale_colour_manual(values = c(oracle = "steelblue", CR = "firebrick")) +
  labs(x = "CV of p_a reported by cde",
       y = "actual |relative error| in p_a",
       title = "Reported uncertainty against actual error",
       subtitle = "Points above the dashed line are errors the model did not see coming") +
  theme_bw()
```

# The mechanism, drawn

The curves themselves, for one realisation. Black is detector 2's true
detection function, which we know exactly because we simulated it. The vertical
line marks the 5th percentile of the CR sample: almost no data exists to the
left of it.

```{r curves-fn}
curvesFor <- function(det1scale, det2scale, seed = 1) {
  s       <- simulatePair(det1scale = det1scale, det2scale = det2scale,
                          seed = seed)
  snrCR   <- snrInfoCR(s)
  snrGrid <- seq(-25, 30, by = 0.25)

  funCR     <- fitCR(snrCR)
  funOracle <- fitDetFun(s$oracleSnr, modelType = "glm")

  rbind(
    data.frame(SNR = snrGrid, curve = "true",
               p = stats::plogis(snrGrid, s$det2location, s$det2scale)),
    data.frame(SNR = snrGrid, curve = "CR",
               p = predictDetFun(funCR,
                                 newdata = data.frame(SNR = snrGrid))$fit),
    data.frame(SNR = snrGrid, curve = "oracle",
               p = predictDetFun(funOracle,
                                 newdata = data.frame(SNR = snrGrid))$fit)
  ) |>
    transform(det1scale = det1scale, det2scale = det2scale,
              snrP05 = as.numeric(stats::quantile(snrCR$SNR, 0.05)))
}

curves <- do.call(rbind, lapply(c(0.5, 1, 2, 4, 8), curvesFor,
                                det2scale = 4, seed = 1))
```

```{r plot-curves, fig.height = 4}
ggplot(curves, aes(x = SNR, y = p, colour = curve)) +
  geom_vline(aes(xintercept = snrP05), linetype = "dotted", colour = "grey40") +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ det1scale, nrow = 1, labeller = label_both) +
  scale_colour_manual(values = c(true = "black", CR = "firebrick",
                                 oracle = "steelblue")) +
  labs(y = "P(detection by detector 2)",
       title = "Detector 2's detection function, det2scale = 4",
       subtitle = "Dotted line: 5th percentile of the CR sample. Left of it, CR is extrapolating.") +
  theme_bw()
```

The curves can look close and still give a badly wrong $p_a$, because the plot
gives every dB equal weight and $p_a$ does not: nearly all calls sit in the
faint tail.

# Summary

```{r summary-table}
summ <- do.call(rbind, lapply(
  split(sweep, list(sweep$row, sweep$det1scale, sweep$det2scale)),
  function(x) data.frame(
    method    = x$row[1],
    det1scale = x$det1scale[1],
    det2scale = x$det2scale[1],
    nCR       = round(mean(x$nCR)),
    snrP05    = round(mean(x$snrP05), 1),
    paFrac_mean = mean(x$paFraction),
    paFrac_min  = min(x$paFraction),
    paFrac_max  = max(x$paFraction),
    CV.pa       = mean(x$CV.pa)
  )))
summ <- summ[order(summ$det2scale, summ$det1scale, summ$method), ]

knitr::kable(summ, digits = 3, row.names = FALSE)
```

`nCR` is the size of the CR sample and `snrP05` is its 5th percentile: the point
below which the curve is extrapolating. If the hypothesis holds, `snrP05`
should fall as `det1scale` rises, and `paFraction` should tighten as it does.
