---
title: "Observer ground truth and capture-recapture compared on simulated data"
author: "Brian Miller"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(callDensity)
```

# The question

Call density estimation needs three things the data do not give you directly:
the probability of detecting a call somewhere in the study area, $p_a$, the
false discovery rate, $c$, and the distribution of noise levels. The first two
are estimated by comparing two observers.

There are two ways to do that comparison.

**Observer ground truth (OG).** One observer is assumed to be correct. The
second observer's detection function is fitted to the first observer's
detections with an ordinary binomial model. Anything the second observer finds
that the first did not is counted as a false positive.

**Capture-recapture (CR).** Neither observer is assumed correct. Both
observers' detections are modelled jointly with a positive-Bernoulli model
(VGAM `posbernoulli.t`), which conditions on a call having been detected by at
least one observer. Calls are adjudicated against common ground truth, so a
false positive is a false positive.

These notes ask whether CR recovers the true call density where OG does not,
using simulated data where the truth is known exactly.

`cde` always treats `detect_table1` as truth. The methods differ in what goes
there, and in which model is fitted. Nothing else changes.

- OG puts detector 1's raw detections in `detect_table1`.
- CR puts the adjudicated verdict there.

An **oracle** is included throughout. It is not a method. It fits the detection
function to detector 2's own detections over *every call that was made*, not
only the ones an observer happened to notice, and it is given the true noise.
That is a sample you need and can never have. It is here for one purpose: to
separate errors in the estimator from errors in the fit. If the oracle returns
the right density, the machinery works and any error is in the detection
function.

# Setup

```{r simulate-fn}
#' Simulate two detectors listening to the same calls.
#'
#' @param det1location,det1scale,det2location,det2scale Logistic detector
#'   parameters. Location is the SNR at which the detector hears half of calls;
#'   scale controls how sharply it falls off. Small scale means a steep,
#'   nearly step-like detector.
#' @param fdr False discovery rate of both detectors. The proportion of each
#'   detector's detections that are not calls at all.
#' @param n Number of calls.
#' @param seed Random seed. Fixed across the sweep, so every cell sees the same
#'   calls and differences are attributable to the detectors.
#' @return A list holding everything the methods need, plus the truth.
simulatePair <- function(det1location = 1, det1scale = 2,
                         det2location = 2, det2scale = 4,
                         fdr = 0, n = 5e4, seed = 42) {

  set.seed(seed)

  R       <- 1e6                                  # study radius, m
  minDate <- as.POSIXct("2025-01-01")
  maxDate <- as.POSIXct("2026-01-01")
  Time    <- as.numeric(difftime(maxDate, minDate, unit = "days")) / 365
  A       <- studyArea(R / 1e3)

  SL <- data.frame(mean = 190, sd = 4, sampleSize = 350)
  NL <- data.frame(mean = 84,  sd = 4, sampleSize = n)

  det1 <- data.frame(location = det1location, scale = det1scale,
                     func = "plogis", c = fdr, fpMean = 0, fpSD = 2)
  det2 <- data.frame(location = det2location, scale = det2scale,
                     func = "plogis", c = fdr, fpMean = 0, fpSD = 2)

  sim     <- simCallLocation(n = n, R = R, minDate = minDate, maxDate = maxDate)
  sim     <- simCallAcoustics(sim, SL, NL, TL = function(r) 20 * log10(r))
  simDet1 <- simulateDetector(det1, sim)
  simDet2 <- simulateDetector(det2, sim)

  capHistTab <- simsTocaptureHistoryTable(simDet1, simDet2)

  # OG: detector 1's raw detections are asserted to be truth.
  chOG <- capHistTab

  # CR: the adjudicated verdict is truth. Here the verdict is groundTruth,
  # which is what adjudicating against common ground truth recovers. Restricted
  # to calls at least one detector flagged, which is the sample a
  # positive-Bernoulli model is defined on.
  chCR <- capHistTab[capHistTab$detect_table1 == 1 |
                     capHistTab$detect_table2 == 1, ]
  chCR$detect_observer1 <- as.integer(chCR$detect_table1)
  chCR$detect_observer2 <- as.integer(chCR$detect_table2)
  chCR$detect_table1    <- as.integer(chCR$groundTruth1)

  real <- simDet2$groundTruth == 1

  list(
    chOG   = chOG,
    chCR   = chCR,
    SL     = SL,
    NL     = NL,
    TL     = simTLradials_20logR(maxRange = R, rangeStep = 1000,
                                 numTransects = 4),
    A      = A,
    Time   = Time,
    Nc     = sum(simDet2$detect_table),

    # The oracle sample. Detector 2's own detections over every call that was
    # made, not only the ones an observer noticed. Not achievable in practice.
    oracleSnr = data.frame(Detected = simDet2$detect_table[real],
                           SNR      = simDet2$snr[real]),

    # The truth. truePa is a property of detector 2 alone: the mean probability
    # that detector 2 hears a call, over the calls that were made. It cannot
    # depend on detector 1.
    trueDc = n / (A * Time),
    truePa = mean(simDet2$p_det, na.rm = TRUE),
    trueC  = fdr
  )
}
```

The detection functions are fitted to three different samples.

```{r samples-fn}
#' Detector 1's detections, scored by detector 2. The OG sample.
snrInfoOG <- function(s) capHist2snrInfo(s$chOG, "year")

#' Adjudicated real calls, with both detectors' verdicts retained. The CR
#' sample. `capHist2snrInfo` drops the per-observer columns a VGLM needs, so
#' this builds the sample directly.
snrInfoCR <- function(s) {
  g   <- s$chCR[s$chCR$detect_table1 == 1, ]
  out <- data.frame(
    detect_observer1 = g$detect_observer1,
    detect_observer2 = g$detect_observer2,
    Detected         = g$detect_observer2,
    CallRL           = as.numeric(g$signalRMSdB),
    NoiseRL          = as.numeric(g$noiseRMSdB),
    SNR              = as.numeric(g$signalRMSdB - g$noiseRMSdB),
    t                = g$t,
    month            = g$month,
    season           = g$season
  )
  out[is.finite(out$SNR), ]
}
```

Noise measured at detections is biased low: a quiet period gives a high SNR,
and a high SNR gives a detection. Undoing that bias needs the detection
function of the process that *selected the sample*. Under CR that is the
probability at least one detector heard the call, which VGAM computes already
as `type.fitted = "onempall0"`.

```{r punion-fn}
#' Probability at least one observer detects, as a plain function of SNR.
pUnionFromVglm <- function(model, snrGrid = seq(-100, 100, by = 0.1)) {
  pu <- VGAM::predictvglm(model,
                          newdata     = data.frame(SNR = snrGrid),
                          type        = "response",
                          type.fitted = "onempall0")
  stats::approxfun(snrGrid, as.vector(pu), rule = 2)
}
```

```{r runone-fn}
runOne <- function(label, s, ch, detFun, nl, modelType) {
  r <- cde(Nc = s$Nc, capHistTab = ch, snrDetFun = detFun,
           SL = s$SL, TL = s$TL, A = s$A, T = s$Time, NL = nl,
           modelType = modelType)
  data.frame(row        = label,
             trueC      = s$trueC,     c      = r$c,
             trueNlMean = s$NL$mean,   nlMean = nl$mean,
             truePa     = s$truePa,    pa     = r$pa,
             Dc         = r$Dc,
             DcFraction = r$Dc / s$trueDc,
             stringsAsFactors = FALSE)
}
```

# Test 1: no false positives

Two imperfect detectors listen to the same calls. Neither ever produces a false
positive, so the true false discovery rate is exactly zero. If a method reports
a non-zero $c$, that number is an artefact of the method.

The table changes one thing per row, so each row attributes one error.

Row three is **not a method**. It is an ablation: OG's ordinary binomial model,
handed the correct verdict. It answers a question that would otherwise go
unanswered — is CR better simply because it has the right verdict, or does the
VGLM do real work?

```{r ladder-fn}
ladder <- function(fdr) {

  s <- simulatePair(fdr = fdr)

  snrOG <- snrInfoOG(s)
  snrCR <- snrInfoCR(s)

  funOG       <- fitDetFun(snrOG, modelType = "glm")
  funAblation <- fitDetFun(snrCR, modelType = "glm")      # ablation only
  funOracle   <- fitDetFun(s$oracleSnr, modelType = "glm")  # oracle only
  funCR       <- fitDetFun(snrCR, modelType = "vglm",
                           yColNames     = c("detect_observer1",
                                             "detect_observer2"),
                           whichObserver = "detect_observer2")

  nlCR <- nlFromDetections(snrCR, pUnionFromVglm(funCR), s$SL, s$TL)

  rbind(
    data.frame(row = "truth", trueC = s$trueC, c = s$trueC,
               trueNlMean = s$NL$mean, nlMean = s$NL$mean,
               truePa = s$truePa, pa = s$truePa,
               Dc = s$trueDc, DcFraction = 1),
    runOne("OG",                     s, s$chOG, funOG,       s$NL, "glm"),
    runOne("ablation: verdict, glm", s, s$chCR, funAblation, s$NL, "glm"),
    runOne("CR",                     s, s$chCR, funCR,       s$NL, "vglm"),
    runOne("CR, noise estimated",    s, s$chCR, funCR,       nlCR, "vglm"),
    runOne("oracle: true curve",     s, s$chCR, funOracle,   s$NL, "glm")
  )
}
```

`trueC` and `c` are the true and estimated false discovery rate. `trueNlMean`
and `nlMean` are the true and assumed mean noise level in dB. `truePa` and `pa`
are the true and estimated probability of detecting a call somewhere in the
study area. `DcFraction` is estimated density as a fraction of the truth.

```{r test1}
knitr::kable(ladder(fdr = 0), digits = c(0, 3, 4, 2, 2, 4, 4, 5, 3),
             row.names = FALSE)
```

**OG fails on $c$, not on the detection function.** Detector 1 is assumed
correct and is not, so every real call detector 2 finds and detector 1 misses
is recorded as a false positive. $p_a$ is close to right, because given SNR the
two detectors are independent and fitting detector 2's curve on detector 1's
detections is unbiased. The whole error is the spurious $c$, which scales
density by $(1-c)$.

**Fixing the verdict alone makes things worse.** The ablation's sample holds
only calls someone heard, so at low SNR it contains detector 2's successes and
none of the calls both detectors missed. An ordinary binomial model fitted to
it sees roughly $p_2/(p_1+p_2)$ rather than $p_2$, and concludes detector 2
hears almost everything.

**The oracle shows the estimator is sound.** Given the right curve and the
right noise, `cde` returns the right density.

# Test 2: false discovery rate of 0.1

Reporting $c = 0$ when the truth is zero is necessary but not sufficient. This
test gives both detectors a real false discovery rate of 0.1, and asks whether
each method *recovers* it.

```{r test2}
knitr::kable(ladder(fdr = 0.1), digits = c(0, 3, 4, 2, 2, 4, 4, 5, 3),
             row.names = FALSE)
```

# Does it hold generally?

The tests above use one detector pair. This sweep varies the location and scale
of both detectors independently, at both false discovery rates. All cells share
one call realisation, so differences are attributable to the detectors rather
than to the calls.

Each cell runs the two methods as you would actually use them, plus the oracle:

- **OG** with `nlFromSnrInfo`, which is current practice.
- **CR** with `nlFromDetections`, using the union as the selecting curve.
- **oracle** with the true curve and the true noise.

Neither method is given the true noise. The oracle is given everything.

The key comparison is `truePa` against `pa`. **`truePa` is a property of
detector 2 alone.** Within a fixed detector 2 it does not change, so any
movement in `pa` across detector 1 configurations is estimation error.

```{r sweep-fn}
sweepCell <- function(det1location, det1scale, det2location, det2scale, fdr) {

  s <- simulatePair(det1location = det1location, det1scale = det1scale,
                    det2location = det2location, det2scale = det2scale,
                    fdr = fdr)

  snrOG <- snrInfoOG(s)
  snrCR <- snrInfoCR(s)

  funOG     <- fitDetFun(snrOG, modelType = "glm")
  funOracle <- fitDetFun(s$oracleSnr, modelType = "glm")
  funCR     <- fitDetFun(snrCR, modelType = "vglm",
                         yColNames     = c("detect_observer1",
                                           "detect_observer2"),
                         whichObserver = "detect_observer2")

  nlOG <- nlFromSnrInfo(snrOG, funOG)
  nlCR <- nlFromDetections(snrCR, pUnionFromVglm(funCR), s$SL, s$TL)

  out <- rbind(
    runOne("OG",     s, s$chOG, funOG,     nlOG, "glm"),
    runOne("CR",     s, s$chCR, funCR,     nlCR, "vglm"),
    runOne("oracle", s, s$chCR, funOracle, s$NL, "glm")
  )

  cbind(
    data.frame(row = out$row, fdr = fdr,
               det1location = det1location, det1scale = det1scale,
               det2location = det2location, det2scale = det2scale,
               stringsAsFactors = FALSE),
    out[, setdiff(names(out), "row")]
  )
}

grid <- expand.grid(
  det1location = c(0, 4),
  det1scale    = c(1, 4),
  det2location = c(0, 4),
  det2scale    = c(1, 4),
  fdr          = c(0, 0.1),
  KEEP.OUT.ATTRS = FALSE
)
```

```{r sweep-run}
# sweep <- do.call(rbind, lapply(seq_len(nrow(grid)), function(i) {
#   g <- grid[i, ]
#   tryCatch(
#     sweepCell(g$det1location, g$det1scale,
#               g$det2location, g$det2scale, g$fdr),
#     error = function(e) {
#       message(sprintf("cell %d failed: %s", i, conditionMessage(e)))
#       NULL
#     }
#   )
# }))
# 
# sweep$paFraction <- sweep$pa / sweep$truePa
# sweep$nlResidual <- sweep$nlMean - sweep$trueNlMean
library(future.apply)
future::plan(future::multisession, workers = 16)

sweep <- do.call(rbind, future_lapply(seq_len(nrow(grid)), function(i) {
  g <- grid[i, ]
  tryCatch(
    sweepCell(g$det1location, g$det1scale,
              g$det2location, g$det2scale, g$fdr),
    error = function(e) {
      message(sprintf("cell %d failed: %s", i, conditionMessage(e)))
      NULL
    }
  )
}, future.seed = TRUE, future.packages = "callDensity"))

future::plan(future::sequential)

sweep$paFraction <- sweep$pa / sweep$truePa
sweep$nlResidual <- sweep$nlMean - sweep$trueNlMean
```

```{r summarise-fn}
#' Mean, min and max of one column, by method and false discovery rate.
summariseBy <- function(d, what) {
  s <- do.call(rbind, lapply(split(d, list(d$row, d$fdr)), function(x) {
    data.frame(row  = x$row[1], fdr = x$fdr[1],
               mean = mean(x[[what]]),
               min  = min(x[[what]]),
               max  = max(x[[what]]))
  }))
  s[order(s$fdr, s$row), ]
}

#' Mean, min and max of one column for one method, split by detector 1's scale.
summariseByDet1 <- function(d, what, method = "CR") {
  x <- d[d$row == method, ]
  s <- do.call(rbind, lapply(split(x, list(x$det1scale, x$fdr)), function(z) {
    data.frame(det1scale = z$det1scale[1], fdr = z$fdr[1],
               mean = mean(z[[what]]), min = min(z[[what]]),
               max = max(z[[what]]))
  }))
  s[order(s$fdr, s$det1scale), ]
}

```

## Density

`DcFraction` is call density as a fraction of its true value. A correct method
sits at 1 in every cell, whatever the detectors are. The spread matters more
than the mean: a method that is right on average and 50% out in individual
cells is not usable.

```{r sweep-dc}
knitr::kable(summariseBy(sweep, "DcFraction"), digits = 3, row.names = FALSE)
```

## False discovery rate

`c` should equal `trueC` in every cell.

```{r sweep-c}
knitr::kable(summariseBy(sweep, "c"), digits = 4, row.names = FALSE)
```

## Probability of detection

`pa` should equal `truePa` in every cell, so `paFraction` should be 1.

```{r sweep-pa}
knitr::kable(summariseBy(sweep, "paFraction"), digits = 3, row.names = FALSE)
```

## Noise level

`nlResidual` is the assumed mean noise level minus the true one, in dB. It
should be 0. The oracle is given the true noise, so its residual is 0 by
construction and is shown only as a check.

```{r sweep-nl}
knitr::kable(summariseBy(sweep, "nlResidual"), digits = 2, row.names = FALSE)
```

```{r sweep-summarise-by-det1-scale-location}

knitr::kable(summariseByDet1(sweep, "DcFraction"), digits = 3, row.names = FALSE)
knitr::kable(summariseByDet1(sweep, "paFraction"), digits = 3, row.names = FALSE)

```


## Every cell

`trueDc` is the same in every cell, at `r round(simulatePair()$trueDc, 5)`
calls per km^2^ per year, so it is omitted; `DcFraction` carries it.

```{r sweep-all}
knitr::kable(sweep[order(sweep$fdr, sweep$row,
                         sweep$det2location, sweep$det2scale,
                         sweep$det1scale, sweep$det1location), ],
             digits = 3, row.names = FALSE)
```

# Open questions

**The noise measured at detections is doing two jobs.** It is a covariate in
fitting $p(\mathrm{SNR})$, and it parameterises the noise distribution that
`pDetInArea` draws from. Those are different quantities with different biases,
and tying them together is the only reason `nlFromDetections` needs the
selecting curve at all. `noiseLevelDistribution` reads an independent noise
table, measured on a fixed schedule regardless of whether anything was calling.
Where such a table exists, nothing needs correcting. Whether an independently
derived noise distribution can be made commensurate with the SNRs the detection
function was fitted on is a separate question, and probably the more important
one: SNR = RL − NL only holds if NL is measured the same way RL was, in the
same band, over the same averaging duration.
